---
title: "2024 Mifish Elas02 CO1 plots"
output: html_notebook
---


Notebook for making plots from 2024 eDNA data (trawl and hand samples only- not expedition)

Import data tables generated with [/MiFish-Elas-CO1-handsamples-2020-2024.nb.html](MiFish-Elas-CO1-handsamples-2020-2024 notebook)

# Load packages

```{r}
# rm(list = ls())
library(tidyverse)
library(phyloseq)
library(vegan)
library(RColorBrewer)
library(microbiome)
library(metagMisc)
library(fantaxtic)
library(ggrepel)
library(paletteer)
library(pals)
library(compositions)
library(hms)
library(ggvegan)
```

#  Import cleand up data tables
## Mifish-
```{r}
asv_count_mifish <- read.csv("filtered_data/mifish_2024_glommed_asv_table.csv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count_mifish) <- asv_count_mifish$X
asv_count_mifish <- asv_count_mifish %>% select(-X)
asv_count_mifish_mat <- as.matrix(asv_count_mifish)

asv_taxonomy_mifish <- read.csv("filtered_data/mifish_2024_glommed_tax_table.csv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_taxonomy_mifish) <- asv_taxonomy_mifish$X1stRepresentative_ASV
asv_taxonomy_mifish <- asv_taxonomy_mifish %>% select(-X1stRepresentative_ASV)
asv_taxonomy_mifish_mat <- as.matrix(asv_taxonomy_mifish)

sample_metadata_mifish <- read.csv("filtered_data/mifish_2024_sample_data.csv", header=TRUE, stringsAsFactors=TRUE)
row.names(sample_metadata_mifish) <- sample_metadata_mifish$X
sample_metadata_mifish <- sample_metadata_mifish %>% select(-X)


ASV_2024_mifish = otu_table(asv_count_mifish_mat, taxa_are_rows = TRUE)
TAX_2024_mifish = tax_table(asv_taxonomy_mifish_mat)
samples_2024_mifish = sample_data(sample_metadata_mifish)

phylo_2024_mifish <- phyloseq(ASV_2024_mifish, TAX_2024_mifish, samples_2024_mifish)

```


## Elas02
```{r}
asv_count_elas <- read.csv("filtered_data/elas_2024_glommed_asv_table.csv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count_elas) <- asv_count_elas$X
asv_count_elas <- asv_count_elas %>% select(-X)
asv_count_elas_mat <- as.matrix(asv_count_elas)

asv_taxonomy_elas <- read.csv("filtered_data/elas_2024_glommed_tax_table.csv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_taxonomy_elas) <- asv_taxonomy_elas$X1stRepresentative_ASV
asv_taxonomy_elas <- asv_taxonomy_elas %>% select(-X1stRepresentative_ASV)
asv_taxonomy_elas_mat <- as.matrix(asv_taxonomy_elas)

sample_metadata_elas <- read.csv("filtered_data/elas_2024_sample_data.csv", header=TRUE, stringsAsFactors=TRUE)
row.names(sample_metadata_elas) <- sample_metadata_elas$X
sample_metadata_elas <- sample_metadata_elas %>% select(-X)


ASV_2024_elas = otu_table(asv_count_elas_mat, taxa_are_rows = TRUE)
TAX_2024_elas = tax_table(asv_taxonomy_elas_mat)
samples_2024_elas = sample_data(sample_metadata_elas)

phylo_2024_elas <- phyloseq(ASV_2024_elas, TAX_2024_elas, samples_2024_elas)

```

## Common names
Import common names database and set color scheme
```{r}
commonnames <- read_csv(file = "../eDNA-databases/commonnames.csv")

myColors_sciname <- setNames(commonnames$Color, commonnames$Species)
myColors_comname <- setNames(commonnames$Color, commonnames$CommonName)
myColors_Sci_comname <- setNames(commonnames$Color, commonnames$Species.CommonName)


head(myColors_sciname)
head(myColors_comname)
head(myColors_Sci_comname)
```


# Plot positive controls
Note- there are no Elas-2 positive controls, just MiFish

Filter them out into new ps object:
```{r}
# Make new df with positive controls and remove ASVs that are not represented
ps2024_mifish_posctl <- subset_samples(phylo_2024_mifish, controls=="positive")

# Use metagMisc to remove ASV rows with 0 abundance for all positive samples
ps2024_mifish_posctl <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps2024_mifish_posctl,
  minabund = 0,
  relabund = FALSE)

# Make a rel abun ps object also
ps2024_mifish_posctl_ra <- microbiome::transform(ps2024_mifish_posctl, transform = "compositional")
```

Plot rel abun in pos ctls
```{r}
mifish_posctl_barplot <- plot_bar(ps2024_mifish_posctl, fill = "CommonName", x = "replicates", title = "2024 Hand Samples: Positive Controls") + 
  scale_y_continuous(name="Read Abundance")+
  theme_minimal() +
  theme(axis.title.x=element_blank(), legend.title = element_blank())+ 
  scale_fill_manual(values = myColors_comname) 
mifish_posctl_barplot

mifish_posctl_barplot_ra <- plot_bar(ps2024_mifish_posctl_ra, fill = "CommonName", x = "replicates", title = "2024 Hand Samples: Positive Controls") + 
  scale_y_continuous(name="Read Abundance")+
  theme_minimal() +
  theme(axis.title.x=element_blank(), legend.title = element_blank()) + 
  scale_fill_manual(values = myColors_comname) 
mifish_posctl_barplot_ra

ggsave(plot = mifish_posctl_barplot_ra, filename = "figures-2024eDNA/mifish_posctl_barplot_ra.jpg", width = 7, height = 4, units = "in")

```


## Bubble Plots
### Prepare data
Remove pos controls, glom by Species.CommonName, and convert to dataframe
#### Mifish
```{r}
phylo_2024_mifish_glommed <- tax_glom(phylo_2024_mifish, "Species.CommonName")
mifish_df <- subset_samples(phylo_2024_mifish_glommed, !controls %in% c("positive"))
mifish_df <- psmelt(mifish_df)

#relative abund ps object
phylo_2024_mifish_glommed_ra <- tax_glom(phylo_2024_mifish, "Species.CommonName")
mifish_df_ra <- subset_samples(phylo_2024_mifish_glommed_ra, !controls %in% c("positive")) 
mifish_df_ra <- microbiome::transform(mifish_df_ra, transform = "compositional") 
mifish_df_ra <- psmelt(mifish_df_ra)

# replace zeroes in the table with NA
mifish_df[mifish_df == 0] <- NA
mifish_df_ra[mifish_df_ra == 0] <- NA

# and remove rows with NAs (so they don't appear as small dots in plot)
mifish_df <-  filter(mifish_df, !is.na(Abundance))
mifish_df_ra <-  filter(mifish_df_ra, !is.na(Abundance))


# and view
mifish_df
mifish_df_ra
```


#### Elas
```{r}
phylo_2024_elas_glommed <- tax_glom(phylo_2024_elas, "Species.CommonName")
elas_df <- subset_samples(phylo_2024_elas_glommed, !controls %in% c("positive"))
elas_df <- psmelt(elas_df)

#relative abund ps object
phylo_2024_elas_glommed_ra <- tax_glom(phylo_2024_elas, "Species.CommonName")
elas_df_ra <- subset_samples(phylo_2024_elas_glommed_ra, !controls %in% c("positive")) 
elas_df_ra <- microbiome::transform(elas_df_ra, transform = "compositional") 
elas_df_ra <- psmelt(elas_df_ra)

# replace zeroes in the table with NA
elas_df[elas_df == 0] <- NA
elas_df_ra[elas_df_ra == 0] <- NA

# and remove rows with NAs (so they don't appear as small dots in plot)
elas_df <-  filter(elas_df, !is.na(Abundance))
elas_df_ra <-  filter(elas_df_ra, !is.na(Abundance))


# and view
elas_df
elas_df_ra
```

### Prepare for plot

#### Remove controls from dataframes
```{r}
mifish_df <- mifish_df %>% filter(is.na(controls))
mifish_df_ra <- mifish_df_ra %>% filter(is.na(controls))

elas_df <- elas_df %>% filter(is.na(controls))
elas_df_ra <- elas_df_ra %>% filter(is.na(controls))
```


#### Sort levels

```{r}
# sort station numbers
sitelevels <- as.character(sort(as.numeric(unique(c(mifish_df$sites)))))
sitelevels

# put east/west in geographic/sensible order (not alphabetic)
mifish_df$Bayside_f= factor(mifish_df$Bayside, levels=c('WEST','EAST'))
mifish_df_ra$Bayside_f= factor(mifish_df_ra$Bayside, levels=c('WEST','EAST'))
elas_df$Bayside_f= factor(elas_df$Bayside, levels=c('WEST','EAST'))
elas_df_ra$Bayside_f= factor(elas_df_ra$Bayside, levels=c('WEST','EAST'))

# put months in time/sensible order (not alphabetic)
mifish_df$Month_f= factor(mifish_df$Month, levels=sort(unique(mifish_df$Month)))
mifish_df_ra$Month_f= factor(mifish_df_ra$Month, levels=sort(unique(mifish_df$Month)))
elas_df$Month_f= factor(elas_df$Month, levels=sort(unique(mifish_df$Month)))
elas_df_ra$Month_f= factor(elas_df_ra$Month, levels=sort(unique(mifish_df$Month)))

```

### Mifish Bubble Plots
#### By bayside
```{r}
# First average replicates at same sites
mifish_df_bayside <- mifish_df %>%
    group_by(sites, Bayside_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
mifish_df_bayside


bubbleplot_mifish_Bayside <- ggplot(mifish_df_bayside, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 9), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Bayside_f, scales = "free", space = "free", drop= TRUE)
bubbleplot_mifish_Bayside
  
# save
ggsave(plot = bubbleplot_mifish_Bayside, filename = "figures-2024eDNA/bubbleplot_mifish_Bayside.jpg", width = 10, height = 7, units = "in")
```

#### By month
```{r}
# First average replicates from same month
mifish_df_month <- mifish_df %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
mifish_df_month


bubbleplot_mifish_Month <- ggplot(mifish_df_month, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_wrap(~Month_f, drop= TRUE, nrow = 1)  
bubbleplot_mifish_Month
  
# save
ggsave(plot = bubbleplot_mifish_Month, filename = "figures-2024eDNA/bubbleplot_mifish_Month.jpg", width = 22, height = 8, units = "in")
```

#### By month, trawl sites only

```{r}
# First average replicates from same month
mifish_df_month_trawl <- mifish_df %>%
  filter(sites %in% c(1,2,3,4,5,6,7,8,9,10,11,12,13)) %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
mifish_df_month_trawl


bubbleplot_mifish_Month_trawlonly <- ggplot(mifish_df_month_trawl, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_wrap(~Month_f, drop= TRUE, nrow = 1)  
bubbleplot_mifish_Month_trawlonly
  
# save
ggsave(plot = bubbleplot_mifish_Month_trawlonly, filename = "figures-2024eDNA/bubbleplot_mifish_Month_trawlonly.jpg", width = 12, height = 8, units = "in")
```



#### By month, other sites

First checked labels and "SEA GRASS" AND "EEL GRASS" should be the same. Fix
```{r}
#check before
unique(mifish_df$Habitat)

mifish_df <- mifish_df %>%
  mutate(Habitat = case_when(
    Habitat == "SEAGRASS" ~ "EELGRASS",
    TRUE ~ Habitat  # Keep all other values unchanged
  ))

#and after
unique(mifish_df$Habitat)
```

Plot by othet habitat types:
Eelgrass
```{r}
# First average replicates from same month
mifish_df_month_eelgrass <- mifish_df %>%
  filter(Habitat == "EELGRASS") %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
mifish_df_month_eelgrass


bubbleplot_mifish_Month_eelgrass <- ggplot(mifish_df_month_eelgrass, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Month_f, drop = TRUE, scales = "free")
bubbleplot_mifish_Month_eelgrass

# save
ggsave(plot = bubbleplot_mifish_Month_eelgrass, filename = "figures-2024eDNA/bubbleplot_mifish_Month_eelgrass.jpg", width = 9, height = 5, units = "in")
```


Clam Sanctuary
```{r}
# First average replicates from same month
mifish_df_month_clamsanct <- mifish_df %>%
  filter(Habitat == "CLAM SANCTUARY") %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
mifish_df_month_clamsanct


bubbleplot_mifish_Month_clamsanct <- ggplot(mifish_df_month_clamsanct, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Month_f, drop = TRUE, scales = "free")
bubbleplot_mifish_Month_clamsanct

# save
ggsave(plot = bubbleplot_mifish_Month_clamsanct, filename = "figures-2024eDNA/bubbleplot_mifish_Month_clamsanct.jpg", width = 9, height = 3.5, units = "in")
```


Oyster reef
```{r}
# First average replicates from same month
mifish_df_month_oysterreef <- mifish_df %>%
  filter(Habitat == "OYSTER REEF") %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
mifish_df_month_oysterreef


bubbleplot_mifish_Month_oysterreef <- ggplot(mifish_df_month_oysterreef, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Month_f, drop = TRUE, scales = "free")
bubbleplot_mifish_Month_oysterreef

# save
ggsave(plot = bubbleplot_mifish_Month_oysterreef, filename = "figures-2024eDNA/bubbleplot_mifish_Month_oysterreef.jpg", width = 9, height = 4.5, units = "in")
```






### Elas02 Bubble Plots
#### By bayside
```{r}
# First average replicates at same sites
elas_df_bayside <- elas_df %>%
    group_by(sites, Bayside_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
elas_df_bayside


bubbleplot_elas_Bayside <- ggplot(elas_df_bayside, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 9), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Bayside_f, scales = "free", space = "free", drop= TRUE)
bubbleplot_elas_Bayside
  
# save
ggsave(plot = bubbleplot_elas_Bayside, filename = "figures-2024eDNA/bubbleplot_elas_Bayside.jpg", width = 10, height = 3, units = "in")
```


#### By month
```{r}
# First average replicates from same month
elas_df_month <- elas_df %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
elas_df_month


bubbleplot_elas_Month <- ggplot(elas_df_month, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_wrap(~Month_f, drop= TRUE, nrow = 1)  
bubbleplot_elas_Month
  
# save
ggsave(plot = bubbleplot_elas_Month, filename = "figures-2024eDNA/bubbleplot_elas_Month.jpg", width = 22, height = 3, units = "in")
```

#### By month, trawl sites only

```{r}
# First average replicates from same month
elas_df_month_trawl <- elas_df %>%
  filter(sites %in% c(1,2,3,4,5,6,7,8,9,10,11,12,13)) %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
elas_df_month_trawl


bubbleplot_elas_Month_trawlonly <- ggplot(elas_df_month_trawl, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_wrap(~Month_f, drop= TRUE, nrow = 1)  
bubbleplot_elas_Month_trawlonly
  
# save
ggsave(plot = bubbleplot_elas_Month_trawlonly, filename = "figures-2024eDNA/bubbleplot_elas_Month_trawlonly.jpg", width = 11, height = 2.5, units = "in")
```



#### By month, other sites

First checked labels and "SEA GRASS" AND "EEL GRASS" should be the same. Fix
```{r}
#check before
unique(elas_df$Habitat)

elas_df <- elas_df %>%
  mutate(Habitat = case_when(
    Habitat == "SEAGRASS" ~ "EELGRASS",
    TRUE ~ Habitat  # Keep all other values unchanged
  ))

#and after
unique(elas_df$Habitat)
```

Plot by othet habitat types:
Eelgrass
```{r}
# First average replicates from same month
elas_df_month_eelgrass <- elas_df %>%
  filter(Habitat == "EELGRASS") %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
elas_df_month_eelgrass


bubbleplot_elas_Month_eelgrass <- ggplot(elas_df_month_eelgrass, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Month_f, drop = TRUE, scales = "free")
bubbleplot_elas_Month_eelgrass

# save
ggsave(plot = bubbleplot_elas_Month_eelgrass, filename = "figures-2024eDNA/bubbleplot_elas_Month_eelgrass.jpg", width = 9, height = 2.5, units = "in")
```


Clam Sanctuary
```{r}
# First average replicates from same month
elas_df_month_clamsanct <- elas_df %>%
  filter(Habitat == "CLAM SANCTUARY") %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
elas_df_month_clamsanct


bubbleplot_elas_Month_clamsanct <- ggplot(elas_df_month_clamsanct, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Month_f, drop = TRUE, scales = "free")
bubbleplot_elas_Month_clamsanct

# save
ggsave(plot = bubbleplot_elas_Month_clamsanct, filename = "figures-2024eDNA/bubbleplot_elas_Month_clamsanct.jpg", width = 7, height = 2.5, units = "in")
```


Oyster reef
```{r}
# First average replicates from same month
elas_df_month_oysterreef <- elas_df %>%
  filter(Habitat == "OYSTER REEF") %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
elas_df_month_oysterreef


bubbleplot_elas_Month_oysterreef <- ggplot(elas_df_month_oysterreef, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Month_f, drop = TRUE, scales = "free")
bubbleplot_elas_Month_oysterreef

# save
ggsave(plot = bubbleplot_elas_Month_oysterreef, filename = "figures-2024eDNA/bubbleplot_elas_Month_oysterreef.jpg", width = 9, height = 3, units = "in")
```


