---
title: "ShiRP eDNA Time Series- Mifish Elas02 CO1 plots"
output: html_notebook
---


Notebook for making plots from eDNA data (trawl and manual samples only- not expedition)

Import data tables generated with [/MiFish-Elas-CO1-handsamples-2020-2025.nb.html](MiFish-Elas-CO1-handsamples-2020-2025 notebook)

# Load packages

```{r}
# rm(list = ls())
library(tidyverse)
library(phyloseq)
library(vegan)
library(RColorBrewer)
library(microbiome)
library(metagMisc)
library(fantaxtic)
library(ggrepel)
library(paletteer)
library(pals)
library(compositions)
library(hms)
library(vegan)
library(ggvegan)
library(sf)
library(patchwork)
```

# Global vars
update years and output directory for figures, as needed
```{r}
mifish_years <- 2020:2025
elas_years <- c(2021, 2023, 2024, 2025)
co1_years <- c(2024, 2025)

data_dir <-  "filtered_data"
out_dir <- "figures-timeseries"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
```


#  Import cleand up data tables
## Mifish
Function that imports cleaned up data tables and turns into phyloseq object for all mifish data, any year:
```{r}
read_matrix_first_col_as_rownames <- function(path) {
  df <- read.csv(path, header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)

  # Use first column as rownames (whatever its name is)
  rn <- df[[1]]
  df <- df[, -1, drop = FALSE]
  rownames(df) <- rn

  as.matrix(df)
}

read_sampledata_first_col_as_rownames <- function(path) {
  df <- read.csv(path, header = TRUE, stringsAsFactors = TRUE, check.names = FALSE)

  rn <- df[[1]]
  df <- df[, -1, drop = FALSE]
  rownames(df) <- rn

  df
}

make_phyloseq_mifish <- function(year, data_dir = "filtered_data") {

  asv_path  <- file.path(data_dir, sprintf("mifish_%d_glommed_asv_table.csv", year))
  tax_path  <- file.path(data_dir, sprintf("mifish_%d_glommed_tax_table.csv", year))
  samp_path <- file.path(data_dir, sprintf("mifish_%d_sample_data.csv", year))

  asv_mat <- read_matrix_first_col_as_rownames(asv_path)

  # Tax table is different: rownames column might not be first; handle by name if present
  tax_df <- read.csv(tax_path, header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
  if ("X1stRepresentative_ASV" %in% names(tax_df)) {
    rownames(tax_df) <- tax_df$X1stRepresentative_ASV
    tax_df <- tax_df[, setdiff(names(tax_df), "X1stRepresentative_ASV"), drop = FALSE]
  } else {
    # fallback: assume first column contains rownames
    rn <- tax_df[[1]]
    tax_df <- tax_df[, -1, drop = FALSE]
    rownames(tax_df) <- rn
  }
  tax_mat <- as.matrix(tax_df)

  samp_df <- read_sampledata_first_col_as_rownames(samp_path)

  phyloseq(
    otu_table(asv_mat, taxa_are_rows = TRUE),
    tax_table(tax_mat),
    sample_data(samp_df)
  )
}

```

Run
```{r}
phylo_mifish <- setNames(lapply(mifish_years, make_phyloseq_mifish), as.character(mifish_years))
```



## Elas02
Function
```{r}
make_phyloseq_elas <- function(year, data_dir = "filtered_data") {

  asv_path  <- file.path(data_dir, sprintf("elas_%d_glommed_asv_table.csv", year))
  tax_path  <- file.path(data_dir, sprintf("elas_%d_glommed_tax_table.csv", year))
  samp_path <- file.path(data_dir, sprintf("elas_%d_sample_data.csv", year))

  # ASV table: first column = rownames (robust to X / ...1 / blank header)
  asv_mat <- read_matrix_first_col_as_rownames(asv_path)

  # Tax table: prefer REVAMP-style column name, otherwise fall back to first column
  tax_df <- read.csv(tax_path, header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
  if ("X1stRepresentative_ASV" %in% names(tax_df)) {
    rownames(tax_df) <- tax_df$X1stRepresentative_ASV
    tax_df <- tax_df[, setdiff(names(tax_df), "X1stRepresentative_ASV"), drop = FALSE]
  } else {
    rn <- tax_df[[1]]
    tax_df <- tax_df[, -1, drop = FALSE]
    rownames(tax_df) <- rn
  }
  tax_mat <- as.matrix(tax_df)

  # Sample metadata: first column = rownames (robust)
  samp_df <- read_sampledata_first_col_as_rownames(samp_path)

  phyloseq(
    otu_table(asv_mat, taxa_are_rows = TRUE),
    tax_table(tax_mat),
    sample_data(samp_df)
  )
}

```

Run
```{r}

phylo_elas <- setNames(
  lapply(elas_years, make_phyloseq_elas),
  as.character(elas_years)
)
```

## CO1 - trawl only
Keep in mind this is filtered to only those species expected to be caught by trawl (eg. pelagic/ swimming species plus all fish. removed many planktonic and benthic invertebrates)
```{r}
make_phyloseq_co1 <- function(year, data_dir = "filtered_data") {

  asv_path  <- file.path(data_dir, sprintf("co1_%d_glommed_asv_table_fortrawlcomparisononly.csv", year))
  tax_path  <- file.path(data_dir, sprintf("co1_%d_glommed_tax_table_fortrawlcomparisononly.csv", year))
  samp_path <- file.path(data_dir, sprintf("co1_%d_sample_data_fortrawlcomparisononly.csv", year))

  # ASV table
  asv_mat <- read_matrix_first_col_as_rownames(asv_path)

  # Tax table (prefer REVAMP column name; otherwise fall back to first col)
  tax_df <- read.csv(tax_path, header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
  if ("X1stRepresentative_ASV" %in% names(tax_df)) {
    rownames(tax_df) <- tax_df$X1stRepresentative_ASV
    tax_df <- tax_df[, setdiff(names(tax_df), "X1stRepresentative_ASV"), drop = FALSE]
  } else {
    rn <- tax_df[[1]]
    tax_df <- tax_df[, -1, drop = FALSE]
    rownames(tax_df) <- rn
  }
  tax_mat <- as.matrix(tax_df)

  # Sample metadata
  samp_df <- read_sampledata_first_col_as_rownames(samp_path)

  phyloseq(
    otu_table(asv_mat, taxa_are_rows = TRUE),
    tax_table(tax_mat),
    sample_data(samp_df)
  )
}

```

Add Taxon.CommonName rank if missing (for CO1_TOTAL, use Species as display label)
```{r}
add_taxon_commonname_from_species <- function(ps) {
  if (is.null(ps) || ntaxa(ps) == 0) return(ps)

  tt <- as.data.frame(tax_table(ps), stringsAsFactors = FALSE)

  # If Taxon.CommonName already exists, do nothing
  if ("Taxon.CommonName" %in% colnames(tt)) return(ps)

  # If Species exists, copy it over
  if ("Species" %in% colnames(tt)) {
    tt$`Taxon.CommonName` <- as.character(tt$Species)
    tax_table(ps) <- tax_table(as.matrix(tt))
    return(ps)
  }

  # Make something usable so plotting doesn't break
  tt$`Taxon.CommonName` <- rownames(tt)
  tax_table(ps) <- tax_table(as.matrix(tt))
  ps
}

```


Run
```{r}
phylo_co1 <- setNames(
  lapply(co1_years, make_phyloseq_co1),
  as.character(co1_years)
)
```



## CO1 - everything else

```{r}
make_phyloseq_co1_total <- function(year, data_dir = "filtered_data") {

  asv_path  <- file.path(data_dir, sprintf("co1_%d_glommed_asv_table_total.csv", year))
  tax_path  <- file.path(data_dir, sprintf("co1_%d_glommed_tax_table_total.csv", year))
  samp_path <- file.path(data_dir, sprintf("co1_%d_sample_data_total.csv", year))

  # ASV table
  asv_mat <- read_matrix_first_col_as_rownames(asv_path)

  # Tax table (prefer REVAMP column name; otherwise fall back to first col)
  tax_df <- read.csv(tax_path, header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)

  if ("X1stRepresentative_ASV" %in% names(tax_df)) {
    rownames(tax_df) <- tax_df$X1stRepresentative_ASV
    tax_df <- tax_df[, setdiff(names(tax_df), "X1stRepresentative_ASV"), drop = FALSE]
  } else {
    rn <- tax_df[[1]]
    tax_df <- tax_df[, -1, drop = FALSE]
    rownames(tax_df) <- rn
  }

    # ---- FORCE a usable label column for plotting/pruning ----
  # For TOTAL COI, many taxa lack CommonName; use Species as fallback.
  # Create/overwrite Taxon.CommonName so the rest of your script can stay unchanged.
  if (!"Taxon.CommonName" %in% names(tax_df)) {
    tax_df$`Taxon.CommonName` <- NA_character_
  }

  # Start with whatever is already there
  lab <- as.character(tax_df$`Taxon.CommonName`)
  lab[lab == ""] <- NA_character_

  # Fill from CommonName if it exists
  if ("CommonName" %in% names(tax_df)) {
    cn <- as.character(tax_df$CommonName)
    cn[cn == ""] <- NA_character_
    lab[is.na(lab) & !is.na(cn)] <- cn[is.na(lab) & !is.na(cn)]
  }

  # Fill from Species if it exists
  if ("Species" %in% names(tax_df)) {
    sp <- as.character(tax_df$Species)
    sp[sp == ""] <- NA_character_
    lab[is.na(lab) & !is.na(sp)] <- sp[is.na(lab) & !is.na(sp)]
  }

  tax_df$`Taxon.CommonName` <- lab


  tax_mat <- as.matrix(tax_df)

  # Sample metadata
  samp_df <- read_sampledata_first_col_as_rownames(samp_path)

  phyloseq(
    otu_table(asv_mat, taxa_are_rows = TRUE),
    tax_table(tax_mat),
    sample_data(samp_df)
  )
}


# CO1_TOTAL phyloseq list (same years vector unless you change it)
phylo_co1_total <- setNames(
  lapply(co1_years, make_phyloseq_co1_total),
  as.character(co1_years)
)

# helper: prune phyloseq to a set of Taxon.CommonName values (robust even if some missing)
# IMPORTANT: returns NULL if the prune would leave 0 taxa (phyloseq can't hold 0-dim otu_table)
prune_ps_to_taxon_commonname <- function(ps, taxa_keep, tax_rank = "Taxon.CommonName") {
  if (is.null(ps)) return(NULL)
  if (ntaxa(ps) == 0 || nsamples(ps) == 0) return(ps)

  tt <- as.data.frame(tax_table(ps))
  if (!tax_rank %in% names(tt)) return(ps)

  taxa_keep <- unique(as.character(taxa_keep))
  taxa_keep <- taxa_keep[!is.na(taxa_keep) & taxa_keep != ""]

  keep_taxa_ids <- rownames(tt)[as.character(tt[[tax_rank]]) %in% taxa_keep]
  keep_taxa_ids <- keep_taxa_ids[!is.na(keep_taxa_ids)]

  if (length(keep_taxa_ids) == 0) {
    return(NULL)  # <- avoids invalid 0-dimension otu_table
  }

  prune_taxa(keep_taxa_ids, ps)
}

prune_ps_to_species <- function(ps, species_keep, tax_rank = "Species") {
  if (is.null(ps)) return(NULL)
  if (ntaxa(ps) == 0 || nsamples(ps) == 0) return(ps)

  tt <- as.data.frame(tax_table(ps))
  if (!tax_rank %in% names(tt)) return(ps)

  species_keep <- unique(as.character(species_keep))
  species_keep <- species_keep[!is.na(species_keep) & species_keep != ""]

  keep_taxa_ids <- rownames(tt)[as.character(tt[[tax_rank]]) %in% species_keep]
  keep_taxa_ids <- keep_taxa_ids[!is.na(keep_taxa_ids)]

  if (length(keep_taxa_ids) == 0) return(NULL)

  prune_taxa(keep_taxa_ids, ps)
}




```



## Common names
Import common names database to set consistent color scheme
```{r}
commonnames <- read_csv(file = "../eDNA-databases/commonnames.csv")

# Species -> "Taxon.CommonName" display label (e.g. "Species (Common)")
display_lookup <- setNames(commonnames$Taxon.CommonName, commonnames$Taxon)

make_taxon_display <- function(taxon_commonname, species, commonname) {
  tc <- as.character(taxon_commonname); tc[tc == ""] <- NA_character_
  sp <- as.character(species);          sp[sp == ""] <- NA_character_
  cn <- as.character(commonname);       cn[cn == ""] <- NA_character_

  # Prefer the Taxon.CommonName label when it exists,
  # else fall back to CommonName, else Species
  dplyr::coalesce(tc, cn, sp)
}


myColors_sciname <- setNames(commonnames$Color, commonnames$Taxon)
myColors_comname <- setNames(commonnames$Color, commonnames$CommonName)
myColors_Sci_comname <- setNames(commonnames$Color, commonnames$Taxon.CommonName)

# helper
get_db_color <- function(x, default = "grey60") {
  if (length(x) == 0) return(default)

  x <- as.character(x)
  x[x == ""] <- NA_character_

  out <- rep(default, length(x))

  # prefer Taxon.CommonName mapping (already includes "Sci (Common)" style 
  hit1 <- myColors_Sci_comname[x]
  ok1  <- !is.na(hit1)
  out[ok1] <- unname(hit1[ok1])

  # then scientific name
  need <- !ok1 & !is.na(x)
  if (any(need)) {
    hit2 <- myColors_sciname[x[need]]
    ok2  <- !is.na(hit2)
    out[need][ok2] <- unname(hit2[ok2])
  }

  # then common name
  need2 <- (out == default) & !is.na(x)
  if (any(need2)) {
    hit3 <- myColors_comname[x[need2]]
    ok3  <- !is.na(hit3)
    out[need2][ok3] <- unname(hit3[ok3])
  }

  out
}

make_fill_values <- function(labels, default = "grey60") {
  labels <- unique(as.character(labels))
  labels <- labels[!is.na(labels) & labels != ""]
  setNames(get_db_color(labels, default = default), labels)
}

make_taxon_display <- function(tc = NULL, sp = NULL, cn = NULL) {
  # tc = Taxon.CommonName (may already be "Sci (Common)")
  # sp = Species (scientific)
  # cn = CommonName (common)
  tc <- as.character(tc); sp <- as.character(sp); cn <- as.character(cn)

  tc[tc == ""] <- NA_character_
  sp[sp == ""] <- NA_character_
  cn[cn == ""] <- NA_character_

  out <- tc

  # If tc already looks like "Sci (Common)", keep it
  has_combo <- !is.na(out) & grepl("\\(.+\\)", out)

  # Otherwise, if we have Species + CommonName, build "Species (CommonName)"
  need_combo <- !has_combo & !is.na(sp) & !is.na(cn)
  out[need_combo] <- paste0(sp[need_combo], " (", cn[need_combo], ")")

  # Fallbacks
  out[is.na(out) | out == ""] <- dplyr::coalesce(sp, cn, out)

  out
}


head(myColors_sciname)
head(myColors_comname)
head(myColors_Sci_comname)
```

# Plot positive controls

Function to pull out pos control samples
```{r}
make_posctl_objects <- function(ps, control_col = "controls", control_value = "positive") {

  out <- tryCatch({

    md <- as.data.frame(sample_data(ps))
    if (!control_col %in% colnames(md)) return(NULL)

    ctrl_vec <- trimws(tolower(as.character(md[[control_col]])))
    keep_ids <- rownames(md)[ctrl_vec == tolower(control_value)]
    if (length(keep_ids) == 0) return(NULL)

    ps_pos <- prune_samples(keep_ids, ps)
    if (nsamples(ps_pos) == 0) return(NULL)

    # trim taxa absent from all positive controls
    ps_pos <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
      ps_pos, minabund = 0, relabund = FALSE
    )

    if (nsamples(ps_pos) == 0 || ntaxa(ps_pos) == 0) return(NULL)

    ps_pos_ra <- microbiome::transform(ps_pos, transform = "compositional")

    list(ps_pos = ps_pos, ps_pos_ra = ps_pos_ra)

  }, error = function(e) {
    # if phyloseq/metagMisc throws errors, skip
    message("Skipping year due to error: ", conditionMessage(e))
    return(NULL)
  })

  out
}


```


Function to plot pos control plots
```{r}
make_posctl_barplot <- function(ps_obj, ylab_txt) {

  d <- psmelt(ps_obj)
  fill_vals <- make_fill_values(d$Taxon.CommonName)

  plot_bar(ps_obj, fill = "Taxon.CommonName", x = "replicates") +
    scale_y_continuous(name = ylab_txt) +
    theme_minimal() +
    scale_fill_manual(values = fill_vals) +
    guides(fill = guide_legend(nrow = 7, byrow = TRUE)) +
    theme(
      legend.position   = "bottom",
      legend.title      = element_blank(),
      legend.text       = element_text(size = 6, lineheight = 0.7),
      legend.key.size   = unit(0.35, "cm"),
      legend.spacing.x  = unit(0.15, "cm"),
      legend.spacing.y  = unit(0.15, "cm"),
      axis.text.x = element_text(size = 11, angle = 90, vjust = 0.5, hjust = 0),
      axis.text.y = element_text(size = 10)
    ) +
    scale_x_discrete(labels = function(x) gsub("positive", "", x, ignore.case = TRUE)) +
    labs(x = NULL, y = ylab_txt, title = NULL)
}

```

## MiFish
Run function
```{r}
mifish_posctl <- setNames(vector("list", length(mifish_years)), as.character(mifish_years))

for (yr in mifish_years) {
  ps <- phylo_mifish[[as.character(yr)]]
  if (is.null(ps)) next

  mifish_posctl[[as.character(yr)]] <- make_posctl_objects(ps)
}

```

Plot and save
```{r}
mifish_posctl_plots <- list()

for (yr_chr in names(mifish_posctl)) {
  ps_pos    <- mifish_posctl[[yr_chr]]$ps_pos
  ps_pos_ra <- mifish_posctl[[yr_chr]]$ps_pos_ra

  if (is.null(ps_pos) || is.null(ps_pos_ra) || nsamples(ps_pos) == 0) next

  yr <- as.integer(yr_chr)

  p_reads <- make_posctl_barplot(ps_pos, "Read Abundance")
  p_ra    <- make_posctl_barplot(ps_pos_ra, "Relative Abundance")


  mifish_posctl_plots[[yr_chr]] <- list(reads = p_reads, ra = p_ra)

  ggsave(plot = p_reads,
         filename = file.path(out_dir, sprintf("mifish_posctl_barplot_%d.jpg", yr)),
         width = 7, height = 4, units = "in", dpi = 300)

  ggsave(plot = p_ra,
         filename = file.path(out_dir, sprintf("mifish_posctl_barplot_ra_%d.jpg", yr)),
         width = 7, height = 4, units = "in", dpi = 300)
}

```


## Elas
Run
Not every year has elasmo pos ctls, so skip those that don't exist
```{r}
elas_posctl <- list()

for (yr in elas_years) {
  ps <- phylo_elas[[as.character(yr)]]
  if (is.null(ps)) next

  obj <- make_posctl_objects(ps)

  if (is.null(obj)) {
    message("Skipping ", yr, " (no positive controls or no taxa detected)")
    next
  }

  elas_posctl[[as.character(yr)]] <- obj
}


```


Plot and save
```{r}
elas_posctl_plots <- list()

for (yr_chr in names(elas_posctl)) {
  ps_pos    <- elas_posctl[[yr_chr]]$ps_pos
  ps_pos_ra <- elas_posctl[[yr_chr]]$ps_pos_ra

  if (is.null(ps_pos) || is.null(ps_pos_ra) || nsamples(ps_pos) == 0) next

  yr <- as.integer(yr_chr)

  p_reads <- make_posctl_barplot(ps_pos, "Read Abundance")
  p_ra    <- make_posctl_barplot(ps_pos_ra, "Relative Abundance")


  elas_posctl_plots[[yr_chr]] <- list(reads = p_reads, ra = p_ra)

  ggsave(plot = p_reads,
         filename = file.path(out_dir, sprintf("elas_posctl_barplot_%d.jpg", yr)),
         width = 7, height = 4, units = "in", dpi = 300)

  ggsave(plot = p_ra,
         filename = file.path(out_dir, sprintf("elas_posctl_barplot_ra_%d.jpg", yr)),
         width = 7, height = 4, units = "in", dpi = 300)
}


```




## CO1
Run 
```{r}
co1_posctl <- setNames(vector("list", length(co1_years)), as.character(co1_years))

for (yr in co1_years) {
  ps <- phylo_co1[[as.character(yr)]]
  if (is.null(ps)) next

  co1_posctl[[as.character(yr)]] <- make_posctl_objects(ps)
}
```


Plot and save
```{r}
co1_posctl_plots <- list()

for (yr_chr in names(co1_posctl)) {
  ps_pos    <- co1_posctl[[yr_chr]]$ps_pos
  ps_pos_ra <- co1_posctl[[yr_chr]]$ps_pos_ra

  if (is.null(ps_pos) || is.null(ps_pos_ra) || nsamples(ps_pos) == 0) next

  yr <- as.integer(yr_chr)

  p_reads <- make_posctl_barplot(ps_pos, "Read Abundance")
  p_ra    <- make_posctl_barplot(ps_pos_ra, "Relative Abundance")

  co1_posctl_plots[[yr_chr]] <- list(reads = p_reads, ra = p_ra)

  ggsave(plot = p_reads,
         filename = file.path(out_dir, sprintf("co1_posctl_barplot_%d.jpg", yr)),
         width = 7, height = 4, units = "in", dpi = 300)

  ggsave(plot = p_ra,
         filename = file.path(out_dir, sprintf("co1_posctl_barplot_ra_%d.jpg", yr)),
         width = 7, height = 4, units = "in", dpi = 300)
}

```

```{r}
co1_total_posctl <- setNames(vector("list", length(co1_years)), as.character(co1_years))

for (yr in co1_years) {
  ps <- phylo_co1_total[[as.character(yr)]]
  if (is.null(ps)) next
  co1_total_posctl[[as.character(yr)]] <- make_posctl_objects(ps)
}

co1_total_posctl_plots <- list()

for (yr_chr in names(co1_total_posctl)) {
  ps_pos    <- co1_total_posctl[[yr_chr]]$ps_pos
  ps_pos_ra <- co1_total_posctl[[yr_chr]]$ps_pos_ra

  if (is.null(ps_pos) || is.null(ps_pos_ra) || nsamples(ps_pos) == 0) next

  yr <- as.integer(yr_chr)

  p_reads <- make_posctl_barplot(ps_pos, "Read Abundance")
  p_ra    <- make_posctl_barplot(ps_pos_ra, "Relative Abundance")

  co1_total_posctl_plots[[yr_chr]] <- list(reads = p_reads, ra = p_ra)

  ggsave(plot = p_reads,
         filename = file.path(out_dir, sprintf("co1_total_posctl_barplot_%d.jpg", yr)),
         width = 7, height = 4, units = "in", dpi = 300)

  ggsave(plot = p_ra,
         filename = file.path(out_dir, sprintf("co1_total_posctl_barplot_ra_%d.jpg", yr)),
         width = 7, height = 4, units = "in", dpi = 300)
}

# co1_nottrawl_posctl_plots <- list()
# 
# for (yr_chr in names(co1_nottrawl_sample_dfs)) {
# 
#   ps_pos_obj <- try(make_posctl_objects(co1_nottrawl_sample_dfs[[yr_chr]]$ps_raw), silent = TRUE)
#   if (inherits(ps_pos_obj, "try-error") || is.null(ps_pos_obj)) next
# 
#   ps_pos    <- ps_pos_obj$ps_pos
#   ps_pos_ra <- ps_pos_obj$ps_pos_ra
# 
#   if (is.null(ps_pos) || is.null(ps_pos_ra) || nsamples(ps_pos) == 0) next
# 
#   yr <- as.integer(yr_chr)
# 
#   p_reads <- make_posctl_barplot(ps_pos, "Read Abundance")
#   p_ra    <- make_posctl_barplot(ps_pos_ra, "Relative Abundance")
# 
#   co1_nottrawl_posctl_plots[[yr_chr]] <- list(reads = p_reads, ra = p_ra)
# 
#   ggsave(plot = p_reads,
#          filename = file.path(out_dir, sprintf("co1_nottrawl_posctl_barplot_%d.jpg", yr)),
#          width = 7, height = 4, units = "in", dpi = 300)
# 
#   ggsave(plot = p_ra,
#          filename = file.path(out_dir, sprintf("co1_nottrawl_posctl_barplot_ra_%d.jpg", yr)),
#          width = 7, height = 4, units = "in", dpi = 300)
# }

```


# Bubble Plots
## Prepare functions and make dfs
Function to remove controls, sort by station number in sites, order Bayside by West/ East (instead of alphabetically East/West), and order the months in a sensible order, and convert to dataframe
```{r}
prep_psmelt_raw_and_ra <- function(ps,
                                  tax_rank   = "Taxon.CommonName",
                                  order_rank = tax_rank,
                                  control_col = "controls",
                                  site_col    = "sites",
                                  bayside_col = "Bayside",
                                  month_col   = "Month") {

  # helpers
  make_site_levels <- function(x) {
    x_chr <- as.character(x)
    x_num <- suppressWarnings(as.numeric(x_chr))
    if (all(is.na(x_num))) {
      # fallback: character sort
      sort(unique(x_chr))
    } else {
      # numeric sort, keep original string labels (e.g., "01", "1")
      x_chr[order(x_num)][!duplicated(x_num[order(x_num)])] |> unique()
    }
  }

  make_month_levels <- function(x) {
    x_chr <- as.character(x)

    # numeric months (1-12, "01"-"12")
    x_num <- suppressWarnings(as.integer(x_chr))
    if (!all(is.na(x_num))) {
      lev <- sort(unique(x_num))
      return(as.character(lev))
    }

    # month names
    month_full  <- tolower(month.name)   # january..december
    month_abbr  <- tolower(month.abb)    # jan..dec
    x_low <- tolower(x_chr)

    if (all(x_low %in% month_full)) {
      return(month.name[match(sort(unique(x_low), index.return = TRUE)$x, month_full)])
    }
    if (all(x_low %in% month_abbr)) {
      return(month.abb[match(sort(unique(x_low), index.return = TRUE)$x, month_abbr)])
    }

    # fallback: leave in sorted unique order
    sort(unique(x_chr))
  }

  apply_ordering <- function(df) {

    # 1) Taxon ordering 
    if (!is.null(order_rank) && order_rank %in% names(df)) {
      df <- df %>%
        mutate(.tax_order = as.character(.data[[order_rank]])) %>%
        arrange(.tax_order)
    }

    # 2) Sites factor
    if (site_col %in% names(df)) {
      site_levels <- make_site_levels(df[[site_col]])
      df <- df %>%
        mutate(
          sites_f = factor(as.character(.data[[site_col]]), levels = site_levels)
        ) %>%
        arrange(sites_f, .by_group = FALSE)
    }

    # 3) Bayside factor
    if (bayside_col %in% names(df)) {
      df <- df %>%
        mutate(
          Bayside_std = case_when(
            toupper(as.character(.data[[bayside_col]])) %in% c("W","WEST") ~ "WEST",
            toupper(as.character(.data[[bayside_col]])) %in% c("E","EAST") ~ "EAST",
            TRUE ~ NA_character_
          ),
          Bayside_f = factor(Bayside_std, levels = c("WEST","EAST"))
        )
    }

    # 4) Month factor
    if (month_col %in% names(df)) {
      month_levels <- make_month_levels(df[[month_col]])
      df <- df %>%
        mutate(
          Month_f = factor(as.character(.data[[month_col]]), levels = month_levels)
        )
    }

    # optional: drop helper column so it doesn't leak into plots
    if (".tax_order" %in% names(df)) {
      df <- df %>% select(-.tax_order)
    }

    df
  }


  # remove controls (keep only NA)
  md <- as.data.frame(sample_data(ps))
  has_controls <- control_col %in% names(md)

  ps_raw <- ps
  if (has_controls) {
    keep_ids <- rownames(md)[is.na(md[[control_col]])]
    ps_raw <- prune_samples(keep_ids, ps_raw)
  }

  # melt raw
  df_raw <- psmelt(ps_raw)
  df_raw[df_raw == 0] <- NA
  df_raw <- df_raw %>% filter(!is.na(Abundance))
  df_raw <- apply_ordering(df_raw)

  # ---- FIX: match your current make_taxon_display(tc, sp, cn) signature ----
  df_raw$Taxon.Display <- make_taxon_display(
    tc = if ("Taxon.CommonName" %in% names(df_raw)) df_raw$Taxon.CommonName else NA_character_,
    sp = if ("Species" %in% names(df_raw))          df_raw$Species          else NA_character_,
    cn = if ("CommonName" %in% names(df_raw))       df_raw$CommonName       else NA_character_
  )


  # rel abun (tax_glom + compositional)
  if (!tax_rank %in% rank_names(ps)) {
    stop(
      "Tax rank '", tax_rank, "' not found in tax_table(ps). Available: ",
      paste(rank_names(ps), collapse = ", ")
    )
  }

  ps_ra <- tax_glom(ps, taxrank = tax_rank)

  if (has_controls) {
    md2 <- as.data.frame(sample_data(ps_ra))
    keep_ids2 <- rownames(md2)[is.na(md2[[control_col]])]
    ps_ra <- prune_samples(keep_ids2, ps_ra)
  }

  ps_ra <- microbiome::transform(ps_ra, transform = "compositional")

  df_ra <- psmelt(ps_ra)
  df_ra[df_ra == 0] <- NA
  df_ra <- df_ra %>% filter(!is.na(Abundance))
  df_ra <- apply_ordering(df_ra)

  # ---- FIX: match your current make_taxon_display(tc, sp, cn) signature ----
  df_ra$Taxon.Display <- make_taxon_display(
    tc = if ("Taxon.CommonName" %in% names(df_ra)) df_ra$Taxon.CommonName else NA_character_,
    sp = if ("Species" %in% names(df_ra))          df_ra$Species          else NA_character_,
    cn = if ("CommonName" %in% names(df_ra))       df_ra$CommonName       else NA_character_
  )

  list(
    ps_raw = ps_raw,
    df_raw = df_raw,
    ps_ra  = ps_ra,
    df_ra  = df_ra
  )
}


# force the "display" column to be Species if Taxon.CommonName is missing
force_taxlabel_species <- function(x) {
  # x is the list returned by prep_psmelt_raw_and_ra()
  if (!is.null(x$df_raw) && nrow(x$df_raw) > 0 && "Species" %in% names(x$df_raw)) {
    x$df_raw$`Taxon.CommonName` <- as.character(x$df_raw$Species)
  }
  if (!is.null(x$df_ra) && nrow(x$df_ra) > 0 && "Species" %in% names(x$df_ra)) {
    x$df_ra$`Taxon.CommonName` <- as.character(x$df_ra$Species)
  }
  x
}

# glom/RA at Species, then set Taxon.CommonName=Species
prep_psmelt_raw_and_ra_co1_species <- function(ps, ...) {
  x <- prep_psmelt_raw_and_ra(ps, tax_rank = "Species", ...)
  force_taxlabel_species(x)
}

```

Generate tables
```{r}
mifish_sample_dfs <- lapply(phylo_mifish, prep_psmelt_raw_and_ra)
elas_sample_dfs   <- lapply(phylo_elas,   prep_psmelt_raw_and_ra)

# CO1 trawl-only (uses Taxon.CommonName)
co1_sample_dfs <- lapply(
  phylo_co1,
  prep_psmelt_raw_and_ra,
  tax_rank   = "Taxon.CommonName",
  order_rank = "Taxon.CommonName"
)

# CO1_TOTAL (glom at Species, THEN copy Species -> Taxon.CommonName)
co1_total_sample_dfs <- lapply(
  phylo_co1_total,
  prep_psmelt_raw_and_ra_co1_species,
  order_rank = "Species"
)

# CO1_NOTTRAWL = CO1_TOTAL taxa minus taxa present in CO1 trawl-only dataset (per year)
co1_nottrawl_sample_dfs <- purrr::imap(co1_total_sample_dfs, function(x_total, yr_chr) {

  if (is.null(x_total) || is.null(x_total$df_raw) || nrow(x_total$df_raw) == 0) return(x_total)

  x_trawl <- co1_sample_dfs[[yr_chr]]
  if (is.null(x_trawl) || is.null(x_trawl$df_raw) || nrow(x_trawl$df_raw) == 0) {
    return(x_total)  # nothing to subtract
  }

  # ---- Use Taxon.CommonName as the subtraction key (exists in both) ----
  if (!("Taxon.CommonName" %in% names(x_total$df_raw)) || !("Taxon.CommonName" %in% names(x_trawl$df_raw))) {
    stop("Taxon.CommonName missing in one of the CO1 dfs for year ", yr_chr)
  }

  trawl_taxa <- unique(as.character(x_trawl$df_raw$Taxon.CommonName))
  trawl_taxa <- trawl_taxa[!is.na(trawl_taxa) & trawl_taxa != ""]

  # Filter dfs (NOTTRAWL = TOTAL minus trawl taxa)
  x_total$df_raw <- dplyr::filter(x_total$df_raw, !(.data$Taxon.CommonName %in% trawl_taxa))
  x_total$df_ra  <- dplyr::filter(x_total$df_ra,  !(.data$Taxon.CommonName %in% trawl_taxa))

  # Prune phyloseq by Taxon.CommonName (safe; returns NULL if would go to 0 taxa)
  keep_taxa <- unique(as.character(x_total$df_raw$Taxon.CommonName))
  keep_taxa <- keep_taxa[!is.na(keep_taxa) & keep_taxa != ""]

  x_total$ps_raw <- prune_ps_to_taxon_commonname(x_total$ps_raw, keep_taxa, tax_rank = "Taxon.CommonName")
  x_total$ps_ra  <- prune_ps_to_taxon_commonname(x_total$ps_ra,  keep_taxa, tax_rank = "Taxon.CommonName")

  x_total
})


```


Clean up tables a bit
- check site levels are ordered
- average replicates
- remove pos ctl organisms from 2025 data (guinea pig, python)

```{r}
# remove positive-control organisms (taxa) everywhere
posctl_patterns <- c(
  "Cavia",
  "Python"
  # add more strings here as needed (case-insensitive partial match)
)

drop_posctl_taxa_ps <- function(ps, patterns = posctl_patterns,
                                cols = c("Taxon.CommonName", "Species", "CommonName")) {
  tt <- as.data.frame(tax_table(ps))

  cols <- intersect(cols, colnames(tt))
  if (length(cols) == 0) return(ps)

  hit <- Reduce(
    `|`,
    lapply(patterns, function(p) {
      Reduce(`|`, lapply(cols, function(cc) grepl(p, tt[[cc]], ignore.case = TRUE)))
    })
  )

  prune_taxa(!hit, ps)
}

drop_posctl_taxa_df <- function(df, patterns = posctl_patterns,
                                cols = c("Taxon.CommonName", "Species", "CommonName")) {
  cols <- intersect(cols, names(df))
  if (length(cols) == 0) return(df)

  hit <- Reduce(
    `|`,
    lapply(patterns, function(p) {
      Reduce(`|`, lapply(cols, function(cc) grepl(p, df[[cc]], ignore.case = TRUE)))
    })
  )

  dplyr::filter(df, !hit)
}

mifish_sample_dfs <- lapply(mifish_sample_dfs, \(x) { x$df_raw <- drop_posctl_taxa_df(x$df_raw); x$df_ra <- drop_posctl_taxa_df(x$df_ra); x })
elas_sample_dfs   <- lapply(elas_sample_dfs,   \(x) { x$df_raw <- drop_posctl_taxa_df(x$df_raw); x$df_ra <- drop_posctl_taxa_df(x$df_ra); x })

co1_sample_dfs        <- lapply(co1_sample_dfs,        \(x) { x$df_raw <- drop_posctl_taxa_df(x$df_raw); x$df_ra <- drop_posctl_taxa_df(x$df_ra); x })
co1_total_sample_dfs  <- lapply(co1_total_sample_dfs,  \(x) { x$df_raw <- drop_posctl_taxa_df(x$df_raw); x$df_ra <- drop_posctl_taxa_df(x$df_ra); x })
co1_nottrawl_sample_dfs <- lapply(co1_nottrawl_sample_dfs, \(x) { x$df_raw <- drop_posctl_taxa_df(x$df_raw); x$df_ra <- drop_posctl_taxa_df(x$df_ra); x })

# ensure sitelevels are in numeric order for x axis
get_sitelevels <- function(df, site_col = "sites") {
  if ("sites_f" %in% names(df)) {
    return(levels(df$sites_f))
  }

  x <- as.character(df[[site_col]])
  x_num <- suppressWarnings(as.numeric(x))

  if (all(is.na(x_num))) sort(unique(x)) else as.character(sort(unique(x_num)))
}

# average replicates
average_replicates <- function(df, facet_col) {
  grp_cols <- c("sites", facet_col, "CommonName", "Taxon.CommonName", "Taxon.Display")
  grp_cols <- grp_cols[grp_cols %in% names(df)]

  df %>%
    group_by(across(all_of(grp_cols))) %>%
    summarize(Abundance = mean(Abundance), .groups = "drop")
}
```

Single bubble plot function
```{r}
make_bubbleplot <- function(df,
                            facet_col,
                            title = NULL,
                            sitelevels = NULL,
                            size_breaks = c(1000, 10000, 100000, 1000000, 10000000),
                            facet_type = c("grid", "wrap"),
                            nrow_wrap = 1) {

  facet_type <- match.arg(facet_type)
  if (is.null(sitelevels)) sitelevels <- get_sitelevels(df)

  fill_vals <- make_fill_values(df$Taxon.Display)

  p <- ggplot(
    df,
    aes(x = factor(sites, levels = sitelevels),
        y = fct_rev(`Taxon.Display`))
  ) +
    geom_point(
      aes(size = ifelse(Abundance == 0, NA, Abundance),
          fill = `Taxon.Display`),
      color = "black", pch = 21
    ) +
    scale_size_area(breaks = size_breaks) +
    theme_bw() +
    scale_fill_manual(values = fill_vals) +
    labs(x = "", y = "", size = "Read Abundance", fill = "", title = title) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x  = element_text(size = 12),
      axis.text.y  = element_text(size = 10),
      legend.position = "bottom"
    ) +
    guides(fill = "none")

  if (facet_type == "grid") {
    p <- p + facet_grid(as.formula(paste0("~", facet_col)),
                        scales = "free", space = "free", drop = TRUE)
  } else {
    p <- p + facet_wrap(as.formula(paste0("~", facet_col)),
                        drop = TRUE, nrow = nrow_wrap)
  }

  p
}


```

Function to make all bubble plots for one year (by bayside, by month, by habitat and month, using above)
```{r}
make_bubbleplots_for_year <- function(df_raw,
                                      lib = "mifish",
                                      year = 2024,
                                      out_dir = "figures-timeseries",
                                      trawl_sites = 1:13,
                                      habitats_to_plot = c("EELGRASS", "CLAM SANCTUARY", "OYSTER REEF")) {

  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

  # compute sitelevels
  sitelevels <- get_sitelevels(df_raw)

  plots <- list()

  # 1) By Bayside
  if ("Bayside_f" %in% names(df_raw)) {
    df_bayside <- average_replicates(df_raw, facet_col = "Bayside_f")

    plots$bayside <- make_bubbleplot(
      df_bayside,
      facet_col = "Bayside_f",
      sitelevels = sitelevels,
      facet_type = "grid"
    )

    ggsave(
      plots$bayside,
      filename = file.path(out_dir, sprintf("bubbleplot_%s_Bayside_%d.jpg", lib, year)),
      width = 9, height = 8, units = "in", dpi = 300
    )
  }

  # 2) By Month (all sites)
  if ("Month_f" %in% names(df_raw)) {
    df_month <- average_replicates(df_raw, facet_col = "Month_f")

    plots$month <- make_bubbleplot(
      df_month,
      facet_col = "Month_f",
      sitelevels = sitelevels,
      facet_type = "wrap",
      nrow_wrap = 1
    )

    ggsave(
      plots$month,
      filename = file.path(out_dir, sprintf("bubbleplot_%s_Month_%d.jpg", lib, year)),
      width = 22, height = 8, units = "in", dpi = 300
    )

    # 3) By Month (trawl sites only)
    df_month_trawl <- df_raw %>%
      mutate(.site_num = suppressWarnings(as.integer(as.character(sites)))) %>%
      filter(.site_num %in% trawl_sites) %>%
      average_replicates(facet_col = "Month_f")

    plots$month_trawl <- make_bubbleplot(
      df_month_trawl,
      facet_col = "Month_f",
      sitelevels = sitelevels,
      facet_type = "wrap",
      nrow_wrap = 1
    )

    ggsave(
      plots$month_trawl,
      filename = file.path(out_dir, sprintf("bubbleplot_%s_Month_trawlonly_%d.jpg", lib, year)),
      width = 12, height = 8, units = "in", dpi = 300
    )

    # 4) By Month within each habitat
    if ("Habitat" %in% names(df_raw)) {
      for (hab in habitats_to_plot) {
        df_hab <- df_raw %>%
          filter(Habitat == hab) %>%
          average_replicates(facet_col = "Month_f")

        # skip empty habitat-years
        if (nrow(df_hab) == 0) next

        plots[[paste0("month_hab_", gsub(" ", "", tolower(hab)))]] <- make_bubbleplot(
          df_hab,
          facet_col = "Month_f",
          sitelevels = sitelevels,
          facet_type = "grid"
        )

        ggsave(
          plots[[paste0("month_hab_", gsub(" ", "", tolower(hab)))]],
          filename = file.path(out_dir, sprintf("bubbleplot_%s_Month_%s_%d.jpg", lib, gsub(" ", "", tolower(hab)), year)),
          width = 9, height = 5, units = "in", dpi = 300
        )
      }
    }
  }

  plots
}

```


## Make plots
```{r}
bubbleplots_all <- list()

# MiFish
bubbleplots_all$mifish <- lapply(names(mifish_sample_dfs), function(yr_chr) {
  yr <- as.integer(yr_chr)
  df <- mifish_sample_dfs[[yr_chr]]$df_raw
  make_bubbleplots_for_year(df, lib = "mifish", year = yr, out_dir = out_dir)
})
names(bubbleplots_all$mifish) <- names(mifish_sample_dfs)

# Elas02
bubbleplots_all$elas <- lapply(names(elas_sample_dfs), function(yr_chr) {
  yr <- as.integer(yr_chr)
  df <- elas_sample_dfs[[yr_chr]]$df_raw
  make_bubbleplots_for_year(df, lib = "elas", year = yr, out_dir = out_dir)
})
names(bubbleplots_all$elas) <- names(elas_sample_dfs)

# CO1
bubbleplots_all$co1 <- lapply(names(co1_sample_dfs), function(yr_chr) {
  yr <- as.integer(yr_chr)
  df <- co1_sample_dfs[[yr_chr]]$df_raw
  make_bubbleplots_for_year(df, lib = "co1", year = yr, out_dir = out_dir)
})
names(bubbleplots_all$co1) <- names(co1_sample_dfs)

# CO1_NOTTRAWL
bubbleplots_all$co1_nottrawl <- lapply(names(co1_nottrawl_sample_dfs), function(yr_chr) {
  yr <- as.integer(yr_chr)
  df <- co1_nottrawl_sample_dfs[[yr_chr]]$df_raw
  make_bubbleplots_for_year(df, lib = "co1_nottrawl", year = yr, out_dir = out_dir)
})


```







# Seasonality - Time Series
Pull top species, generate a seasonal time series plot, then plot each year on top of each other to assess interannual variability in seasonality

Function to grab top species
```{r}
make_seasonal_overlay_df <- function(df_list,
                                     trawl_sites = as.character(1:13),
                                     site_col = c("site_altname", "sites"),
                                     month_col = "Month_f",
                                     tax_col = "Taxon.CommonName",
                                     abund_col = "Abundance",
                                     top_n = 5,
                                     use_ra = TRUE,
                                     keep_months = NULL) {

  if (is.character(use_ra)) {
    use_ra <- tolower(use_ra) %in% c("true", "t", "ra", "df_ra", "relative", "relabund")
  }
  use_ra <- isTRUE(use_ra)

  df_key <- if (use_ra) "df_ra" else "df_raw"
  ps_key <- if (use_ra) "ps_ra" else "ps_raw"

  # pick a site column from first df that exists
  first_df <- df_list[[1]][[df_key]]
  site_col <- site_col[site_col %in% names(first_df)][1] %||% "sites"
  trawl_sites_key <- as.character(as.integer(trawl_sites))

  # sampled months from SAMPLE METADATA (not psmelt df) ----
  sampled_ym <- purrr::imap_dfr(df_list, function(x, yr) {
    ps <- x[[ps_key]]
    if (is.null(ps) || phyloseq::nsamples(ps) == 0) return(NULL)

    md <- data.frame(phyloseq::sample_data(ps), check.names = FALSE, stringsAsFactors = FALSE)

    # choose site column in metadata too (try same, else fallback)
    site_md_col <- if (site_col %in% names(md)) site_col else {
      c("site_altname", "sites")[c("site_altname", "sites") %in% names(md)][1]
    }

    # choose month column in metadata:
    month_md_col <- if (month_col %in% names(md)) month_col else if ("Month" %in% names(md)) "Month" else month_col

    out <- md %>%
      mutate(
        Year = as.integer(yr),
        .site = as.character(.data[[site_md_col]]),
        .month = as.character(.data[[month_md_col]])
      ) %>%
      filter(.site %in% trawl_sites_key) %>%
      distinct(Year, .month) %>%
      mutate(.sampled = TRUE)

    if (!is.null(keep_months)) out <- out %>% filter(.month %in% keep_months)
    out
  })

  if (is.null(sampled_ym) || nrow(sampled_ym) == 0) {
    return(list(df = tibble::tibble(), top_taxa = character()))
  }

  # abundances from psmelt df (df_ra or df_raw) ----
  all_df <- purrr::imap_dfr(df_list, function(x, yr) {
    d <- x[[df_key]]
    if (is.null(d) || nrow(d) == 0) return(NULL)

    d <- d %>%
      mutate(
        Year = as.integer(yr),
        "{site_col}" := as.character(.data[[site_col]]),
        "{month_col}" := as.character(.data[[month_col]])
      )

    if (!is.null(keep_months)) d <- d %>% filter(.data[[month_col]] %in% keep_months)
    d
  })

  all_df <- all_df %>%
    filter(.data[[site_col]] %in% trawl_sites_key)

  # number of samples per Year-Month
  n_samp <- all_df %>%
    distinct(Year, .month = .data[[month_col]], Sample) %>%
    count(Year, .month, name = "n_samples")
  
  # sum of RA per Year-Month-Taxon (absent taxa contribute 0)
  sum_tax <- all_df %>%
    group_by(Year, .month = .data[[month_col]], .data[[tax_col]]) %>%
    summarize(sum_abund = sum(.data[[abund_col]], na.rm = TRUE), .groups = "drop")
  
  # mean across ALL samples in that Year-Month (including zeros for non-detections)
  df_month <- sum_tax %>%
    left_join(n_samp, by = c("Year", ".month")) %>%
    mutate(MeanAbundance = sum_abund / n_samples) %>%
    select(Year, .month, !!rlang::sym(tax_col), MeanAbundance)

  top_taxa <- df_month %>%
    group_by(.data[[tax_col]]) %>%
    summarize(TotalAbundance = sum(MeanAbundance, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(TotalAbundance)) %>%
    slice_head(n = top_n) %>%
    pull(.data[[tax_col]])

  years <- sort(unique(sampled_ym$Year))
  month_levels <- if (!is.null(keep_months)) unique(as.character(keep_months)) else sort(unique(sampled_ym$.month))

  grid <- tidyr::expand_grid(
    Year = years,
    .month = month_levels,
    !!rlang::sym(tax_col) := top_taxa
  ) %>%
    left_join(sampled_ym, by = c("Year", ".month")) %>%
    left_join(df_month,  by = c("Year", ".month", tax_col)) %>%
    mutate(
      MeanAbundance = dplyr::if_else(
        !is.na(.sampled),
        dplyr::coalesce(MeanAbundance, 0),
        NA_real_
      ),
      Year = factor(Year),
      FacetLabel = stringr::str_wrap(.data[[tax_col]], width = 20)
    )

  list(df = grid, top_taxa = top_taxa)
}

```

Plot function
- keeps species colors and overlays years
```{r}
plot_seasonal_overlay <- function(df_top,
                                  month_col = ".month",
                                  tax_col = "Taxon.CommonName",
                                  colors = myColors_Sci_comname,
                                  title = NULL) {

  year_levels <- sort(unique(as.integer(as.character(df_top$Year))))
  alpha_vals  <- seq(0.25, 1, length.out = length(year_levels))
  names(alpha_vals) <- as.character(year_levels)

  month_levels <- if (is.factor(df_top[[month_col]])) levels(df_top[[month_col]]) else unique(as.character(df_top[[month_col]]))

  df_top <- df_top %>%
    mutate(
  .tax_color = get_db_color(.data[[tax_col]], default = "grey60"),
      Year = factor(as.character(Year), levels = as.character(year_levels)),
      .month = factor(as.character(.data[[month_col]]), levels = month_levels, ordered = TRUE)
    )

  # keep only non-NA points for line drawing, and only groups with >=2 points
  df_lines <- df_top %>%
    filter(!is.na(MeanAbundance)) %>%
    group_by(FacetLabel, Year) %>%
    filter(n() >= 2) %>%
    ungroup()


  ggplot(df_top, aes(
    x = .month,
    y = MeanAbundance,
    group = Year,
    alpha = Year
  )) +
    geom_line(
      data = df_lines,
      aes(color = .tax_color),
      linewidth = 1
    ) +
    geom_point(
      aes(color = .tax_color),
      size = 1.6,
      na.rm = TRUE
    ) +
    facet_wrap(~FacetLabel, scales = "free_y") +
    scale_color_identity() +
    scale_alpha_manual(
      name = "Year\n(lighter = earlier)",
      values = alpha_vals,
      drop = FALSE
    ) +
    labs(x = "", y = "Relative abundance (avg across trawl sites)", title = title) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      legend.box = "vertical",
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 10)
    ) +
    guides(
      alpha = guide_legend(
        override.aes = list(color = "black", linewidth = 1.2),
        nrow = 1
      )
    )
}

```

Generate
```{r}
# modify for which months should be plotted (numeric codes only)
keep_months <- c("5","6","7","8","9","10")

run_library_overlay <- function(df_list, lib_name, top_n = 5, use_ra = TRUE, keep_months = NULL) {

  tmp <- make_seasonal_overlay_df(
    df_list,
    top_n = top_n,
    use_ra = use_ra,
    keep_months = keep_months
  )

  # if nothing comes back, return a placeholder 
  if (is.null(tmp$df) || !is.data.frame(tmp$df) || nrow(tmp$df) == 0 || !(".month" %in% names(tmp$df))) {
    p_empty <- ggplot() +
      theme_void() +
      labs(title = paste0(lib_name, " top ", top_n, " taxa: seasonal overlay")) +
      annotate("text", x = 0, y = 0,
               label = "No data after filtering (check trawl sites / keep_months / Month column).",
               size = 4)

    return(list(
      plot = p_empty,
      df = tmp$df,
      top_taxa = tmp$top_taxa %||% character(),
      filename = sprintf("%s_seasonal_overlay_top%d.jpg", lib_name, top_n)
    ))
  }

  df_top <- tmp$df %>%
    mutate(.month = as.character(.month))

  # provided keep_months, enforce them (recommended: numeric only)
  if (!is.null(keep_months)) {
    keep_months <- as.character(keep_months)

    df_top <- df_top %>%
      filter(.month %in% keep_months)

    # Relabel numeric month codes to names if they are exactly 5-10
    if (all(keep_months %in% c("5","6","7","8","9","10"))) {
      df_top <- df_top %>%
        mutate(.month = factor(
          .month,
          levels = c("5","6","7","8","9","10"),
          labels = c("May","Jun","Jul","Aug","Sep","Oct"),
          ordered = TRUE
        ))
    } else {
      # Otherwise just keep the provided order
      df_top <- df_top %>%
        mutate(.month = factor(.month, levels = keep_months, ordered = TRUE))
    }
  } else {
    # If keep_months is NULL, just preserve whatever order appears
    df_top <- df_top %>%
      mutate(.month = factor(.month, levels = unique(.month), ordered = TRUE))
  }

  p <- plot_seasonal_overlay(
    df_top,
    month_col = ".month",
    title = paste0(lib_name, " top ", top_n, " taxa: seasonal overlay")
  )

  list(
    plot = p,
    df = df_top,
    top_taxa = tmp$top_taxa,
    filename = sprintf("%s_seasonal_overlay_top%d.jpg", lib_name, top_n)
  )
}



```

Visualize
```{r}
# run all three
mifish_overlay_top10 <- run_library_overlay(mifish_sample_dfs, "MiFish", top_n = 10, use_ra = TRUE, keep_months = keep_months)
elas_overlay_top5    <- run_library_overlay(elas_sample_dfs,   "Elas02", top_n = 5,  use_ra = TRUE, keep_months = keep_months)
co1_overlay_top10    <- run_library_overlay(co1_sample_dfs,    "CO1",    top_n = 10, use_ra = TRUE, keep_months = keep_months)
co1_nottrawl_overlay_top10 <- run_library_overlay(co1_nottrawl_sample_dfs, "CO1_NOTTRAWL", top_n = 10, use_ra = TRUE, keep_months = keep_months)


mifish_overlay_top10$top_taxa
elas_overlay_top5$top_taxa
co1_overlay_top10$top_taxa
co1_nottrawl_overlay_top10$top_taxa

# show plot in notebook
mifish_overlay_top10$plot
elas_overlay_top5$plot
co1_overlay_top10$plot
co1_nottrawl_overlay_top10$plot # not working but move on for now

```

Export
```{r}
save_overlay <- function(obj, out_dir, width = 11, height = 8, units = "in", dpi = 300) {
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  ggsave(
    filename = file.path(out_dir, obj$filename),
    plot = obj$plot,
    width = width,
    height = height,
    units = units,
    dpi = dpi
  )
}

# Export
save_overlay(mifish_overlay_top10, out_dir, width = 11, height = 8)
save_overlay(elas_overlay_top5,    out_dir, width = 11, height = 8)
save_overlay(co1_overlay_top10,    out_dir, width = 11, height = 8)
save_overlay(co1_nottrawl_overlay_top10, out_dir, width = 11, height = 8)

```

# Seasonality- Maps
## First remake map of Shinnecock Bay
[GSHHG map from NOAA](https://www.soest.hawaii.edu/pwessel/gshhg/).  
Following [tutorial](https://rpubs.com/MRufino/coastlines).   
Also had to add in [fix](https://gis.stackexchange.com/questions/404385/r-sf-some-edges-are-crossing-in-a-multipolygon-how-to-make-it-valid-when-using).  
```{r}
# lets get a box to make a smaller sized objects and facilitate the examples:
geo.box <- c(xmin=-72.61, xmax=-72.4, ymin=40.79, ymax=40.9)

sf_use_s2(FALSE)

# Shoreline data for us detail
gshhg.l1 <- sf::read_sf("raw_data/gshhg-shp-2.3.7/GSHHS_shp/f/GSHHS_f_L1.shp") %>% st_crop(geo.box)
plot(gshhg.l1["id"])

# make ggplot object
shinnbay_map <- ggplot(data = gshhg.l1) +
  geom_sf(aes(geometry = geometry)) +
    scale_x_continuous(breaks = seq(-72.6, -72.4, 0.1)) +
    scale_y_continuous(breaks = seq(40.80, 40.9, 0.05))

shinnbay_map

```

## Species Maps

Functions to map a given species for a given library/year over season (every sampling event) 
```{r}
# choose the first column name that exists
pick_col <- function(df, candidates, required = TRUE) {
  out <- candidates[candidates %in% names(df)][1]
  if (length(out) == 0 || is.na(out)) {
    if (required) stop("None of these columns exist: ", paste(candidates, collapse = ", "))
    return(NULL)
  }
  out
}



# 1) Display candidate taxa for a given library + year (df_ra only)
candidate_taxa <- function(df_list,
                           year,
                           tax_col = "Taxon.CommonName",
                           sample_col = "Sample",
                           ra_col = "Abundance",
                           top_n = 30,
                           min_prevalence = 0) {

  year <- as.character(year)
  if (!year %in% names(df_list)) stop("Year not found in df_list: ", year)

  d <- df_list[[year]]$df_ra
  if (is.null(d) || nrow(d) == 0) stop("df_ra is empty for year ", year)

  tax_col    <- pick_col(d, c(tax_col, "Taxon.CommonName", "Taxon", "taxon"))
  sample_col <- pick_col(d, c(sample_col, "Sample", "sample", "SampleID"))
  ra_col     <- pick_col(d, c(ra_col, "Abundance", "RelAbundance", "RA"))

  # collapse multiple ASVs per sample->taxon (sum RA within sample)
  d_sum <- d %>%
    mutate(
      .tax   = as.character(.data[[tax_col]]),
      .samp  = as.character(.data[[sample_col]]),
      .ra    = as.numeric(.data[[ra_col]])
    ) %>%
    group_by(.samp, .tax) %>%
    summarize(ra = sum(.ra, na.rm = TRUE), .groups = "drop")

  n_samples_total <- d_sum %>% distinct(.samp) %>% nrow()

  out <- d_sum %>%
    group_by(.tax) %>%
    summarize(
      n_samples_present = sum(ra > 0, na.rm = TRUE),
      prevalence = n_samples_present / n_samples_total,
      mean_ra = mean(ra, na.rm = TRUE),
      median_ra = median(ra, na.rm = TRUE),
      max_ra = max(ra, na.rm = TRUE),
      total_ra = sum(ra, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    filter(prevalence >= min_prevalence) %>%
    arrange(desc(total_ra)) %>%
    mutate(rank = row_number()) %>%
    select(rank, taxon = .tax, everything()) %>%
    slice_head(n = top_n)

  out
}

# 2) Map one chosen taxon’s per-sample RA across all timepoints in a year
map_taxon_year <- function(df_list,
                           year,
                           taxon,
                           base_map,                 # your shinnbay_map
                           bbox = c(xmin=-72.61, xmax=-72.4, ymin=40.79, ymax=40.9),
                           colors = myColors_Sci_comname,
                           tax_col = "Taxon.CommonName",
                           sample_col = "Sample",
                           ra_col = "Abundance",
                           lat_col = c("lat", "Latitude", "Lat"),
                           lon_col = c("long", "lon", "Longitude", "Long"),
                           datecode_col = c("Datecode", "datecode"),
                           date_label_col = c("Date", "date"),
                           facet_ncol = 2,
                           point_size_range = c(1.2, 6),
                           alpha_points = 0.85) {

  year <- as.character(year)
  if (!year %in% names(df_list)) stop("Year not found in df_list: ", year)

  d <- df_list[[year]]$df_ra
  if (is.null(d) || nrow(d) == 0) stop("df_ra is empty for year ", year)

  pick_col <- function(df, candidates, required = TRUE) {
    out <- candidates[candidates %in% names(df)][1]
    if (length(out) == 0 || is.na(out)) {
      if (required) stop("None of these columns exist: ", paste(candidates, collapse = ", "))
      return(NULL)
    }
    out
  }

  tax_col        <- pick_col(d, c(tax_col, "Taxon.CommonName", "Taxon", "taxon"))
  sample_col     <- pick_col(d, c(sample_col, "Sample", "sample", "SampleID"))
  ra_col         <- pick_col(d, c(ra_col, "Abundance", "RelAbundance", "RA"))
  lat_col        <- pick_col(d, lat_col)
  lon_col        <- pick_col(d, lon_col)
  datecode_col   <- pick_col(d, datecode_col)
  date_label_col <- pick_col(d, date_label_col, required = FALSE)

  # per-sample RA for this taxon (sum within sample in case multiple OTUs map to same taxon)
  d_tax <- d %>%
    transmute(
      .tax  = as.character(.data[[tax_col]]),
      .samp = as.character(.data[[sample_col]]),
      .ra   = as.numeric(.data[[ra_col]]),
      .lat  = as.numeric(.data[[lat_col]]),
      .lon  = as.numeric(.data[[lon_col]]),
      .dc   = as.integer(.data[[datecode_col]]),
      .date = if (!is.null(date_label_col) && date_label_col %in% names(d)) as.character(.data[[date_label_col]]) else NA_character_
    ) %>%
    filter(.tax == taxon) %>%
    group_by(.samp, .dc, .lat, .lon, .tax) %>%
    summarize(ra = sum(.ra, na.rm = TRUE), .date = dplyr::first(.date), .groups = "drop")

  if (nrow(d_tax) == 0) stop("No rows found for taxon = '", taxon, "' in year ", year)

  # facet VARIABLE is Datecode
  dc_levels <- sort(unique(d_tax$.dc))
  d_tax <- d_tax %>%
    mutate(.dc_f = factor(.dc, levels = dc_levels))

  # build labels (Date if present, else YYYY-MM-DD from Datecode)
  dc_to_label <- setNames(
    ifelse(
      is.na(tapply(d_tax$.date, d_tax$.dc, function(x) x[!is.na(x)][1] %||% NA_character_)),
      paste0(substr(dc_levels, 1, 4), "-", substr(dc_levels, 5, 6), "-", substr(dc_levels, 7, 8)),
      as.character(tapply(d_tax$.date, d_tax$.dc, function(x) x[!is.na(x)][1] %||% NA_character_))
    ),
    as.character(dc_levels)
  )

  # bbox filter (catches accidental +72 longitudes etc.)
  before_n <- nrow(d_tax)
  d_tax <- d_tax %>%
    filter(
      .lon >= bbox["xmin"], .lon <= bbox["xmax"],
      .lat >= bbox["ymin"], .lat <= bbox["ymax"]
    )
  after_n <- nrow(d_tax)
  if (after_n == 0) {
    stop("All points fell outside bbox. Check longitude/latitude sign/values in df_ra for ", year)
  }
  if (after_n < before_n) {
    message("Dropped ", before_n - after_n, " rows outside bbox (check lon/lat sign issues).")
  }

  # constant color per taxon
  tax_color <- get_db_color(taxon, default = "grey60")


  base_map +
    geom_point(
      data = d_tax,
      aes(x = .lon, y = .lat, size = ra),
      color = tax_color,
      alpha = alpha_points
    ) +
    facet_wrap(
      ~ .dc_f,
      ncol = facet_ncol,
      labeller = as_labeller(function(x) dc_to_label[as.character(x)])
    ) +
    scale_size_continuous(range = point_size_range) +
    coord_sf(
      xlim = c(bbox["xmin"], bbox["xmax"]),
      ylim = c(bbox["ymin"], bbox["ymax"]),
      expand = FALSE
    ) +
    labs(
      title = paste0("Relative abundance: ", taxon, " (", year, ")"),
      size = "Rel. abundance"
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      strip.text = element_text(size = 10),
      panel.spacing = unit(0.9, "lines"),
      plot.title = element_text(face = "bold"),
      axis.title.x = element_blank(),
      axis.title.y = element_blank()
    )
}

```

Make dfs of candidate species
```{r}
make_candidate_list <- function(df_list,
                                lib_name = "LIB",
                                top_n = 30,
                                min_prevalence = 0,
                                tax_col = "Taxon.CommonName",
                                sample_col = "Sample",
                                ra_col = "Abundance") {

  yrs <- names(df_list)
  if (is.null(yrs) || length(yrs) == 0) stop("df_list has no names (years).")

  # ensure years are in chronological order when possible
  yrs_num <- suppressWarnings(as.integer(yrs))
  if (!all(is.na(yrs_num))) yrs <- yrs[order(yrs_num)]

  cand_list <- purrr::map(yrs, function(y) {
    message("Candidate taxa: ", lib_name, " ", y)
    candidate_taxa(
      df_list = df_list,
      year = y,
      top_n = top_n,
      min_prevalence = min_prevalence,
      tax_col = tax_col,
      sample_col = sample_col,
      ra_col = ra_col
    )
  })
  names(cand_list) <- paste0("cand_", yrs, "_", lib_name)

  cand_list
}
```

Build them for each library in one go 
```{r}
cand_mifish <- make_candidate_list(mifish_sample_dfs, lib_name = "mifish", top_n = 100)
cand_elas   <- make_candidate_list(elas_sample_dfs,   lib_name = "elas",   top_n = 100)
cand_co1    <- make_candidate_list(co1_sample_dfs,    lib_name = "co1",    top_n = 100,  tax_col = "Taxon.Display")
cand_co1_nottrawl <- make_candidate_list(co1_nottrawl_sample_dfs, lib_name = "co1_nottrawl", top_n = 1000, tax_col = "Taxon.Display")

```

Map a candidate to test
```{r}
chosen <-cand_co1$cand_2024_co1$taxon[5]

hc_map_2024 <- map_taxon_year(
  co1_sample_dfs,
  year = 2024,
  taxon = chosen,
  base_map = shinnbay_map,
  tax_col = "Taxon.Display",
  bbox = c(xmin=-72.61, xmax=-72.4, ymin=40.79, ymax=40.9),
  facet_ncol = 2
)


hc_map_2024

chosen <- cand_co1$cand_2025_co1$taxon[7]

hc_map_2025 <- map_taxon_year(
  co1_sample_dfs,
  year = 2025,
  taxon = chosen,
  base_map = shinnbay_map,
  bbox = c(xmin=-72.61, xmax=-72.4, ymin=40.79, ymax=40.9),
  facet_ncol = 2
)

hc_map_2025

chosen_nt <- cand_co1_nottrawl$cand_2024_co1_nottrawl$taxon[81]

nt_map_2024 <- map_taxon_year(
  co1_nottrawl_sample_dfs,
  year = 2024,
  taxon = chosen_nt,
  base_map = shinnbay_map,
  bbox = c(xmin=-72.61, xmax=-72.4, ymin=40.79, ymax=40.9),
  facet_ncol = 2
)

nt_map_2024


```
Save all species from a single library/year as one pdf
Function:
```{r}
save_species_maps_pdf <- function(df_list,
                                  cand_df,
                                  year,
                                  lib_name,
                                  base_map,
                                  facet_ncol = 2,
                                  out_dir,
                                  width = 11,
                                  height = 8.5) {

  stopifnot("taxon" %in% names(cand_df))

  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

  pdf_file <- file.path(
    out_dir,
    paste0(lib_name, "_", year, "_species_maps.pdf")
  )

  taxa <- unique(as.character(cand_df$taxon))
  taxa <- taxa[!is.na(taxa) & taxa != ""]

  pdf(pdf_file, width = width, height = height, onefile = TRUE)
  on.exit(dev.off(), add = TRUE)

  for (tx in taxa) {
    message("Plotting: ", lib_name, " ", year, " | ", tx)

    p <- map_taxon_year(
      df_list  = df_list,
      year     = year,
      taxon    = tx,
      base_map = base_map,
      facet_ncol = facet_ncol
    )

    print(p)
  }

  message("Saved: ", pdf_file)
  invisible(pdf_file)
}

```

Run across all years/libraries

Function-
```{r}
# helpers to standardize names
normalize_cand_names <- function(cand_list, lib_label) {
  out <- cand_list

  names(out) <- gsub(
    pattern = paste0("^cand_(\\d{4})_", lib_label, "$"),
    replacement = paste0(lib_label, "_\\1"),
    x = names(out)
  )

  out
}

cand_mifish_norm <- normalize_cand_names(cand_mifish, "mifish")
cand_elas_norm   <- normalize_cand_names(cand_elas,   "elas")
cand_co1_norm    <- normalize_cand_names(cand_co1,    "co1")
cand_co1_nottrawl_norm <- normalize_cand_names(cand_co1_nottrawl, "co1_nottrawl")

names(cand_mifish_norm) <- sub("^mifish_", "MiFish_", names(cand_mifish_norm))
names(cand_elas_norm)   <- sub("^elas_",   "Elas02_", names(cand_elas_norm))
names(cand_co1_norm)    <- sub("^co1_",    "CO1_",    names(cand_co1_norm))
names(cand_co1_nottrawl_norm) <- sub("^co1_nottrawl_", "CO1_NOTTRAWL_", names(cand_co1_nottrawl_norm))

cand_list <- c(
  cand_co1_norm,
  cand_co1_nottrawl_norm,
  cand_mifish_norm,
  cand_elas_norm
)

lib_lists <- list(
  CO1         = co1_sample_dfs,
  CO1_NOTTRAWL = co1_nottrawl_sample_dfs,
  MiFish      = mifish_sample_dfs,
  Elas02      = elas_sample_dfs
)


# helper
parse_lib_year <- function(nm) {
  parts <- strsplit(nm, "_")[[1]]
  if (length(parts) < 2) return(NULL)

  year <- tail(parts, 1)
  lib  <- paste(parts[-length(parts)], collapse = "_")

  list(lib = lib, year = year)
}


# function
run_all_species_pdfs <- function(lib_lists, cand_list, base_map, out_dir, facet_ncol = 2,
                                 width = 11, height = 8.5) {

  for (nm in names(cand_list)) {
    py <- parse_lib_year(nm)
    if (is.null(py)) {
      message("Skipping cand_list item with bad name: ", nm)
      next
      }
    
    lib  <- py$lib
    year <- py$year

    if (!lib %in% names(lib_lists)) {
      message("Skipping (no df_list for lib): ", lib, " [", nm, "]")
      next
    }

    df_list <- lib_lists[[lib]]

    if (!as.character(year) %in% names(df_list)) {
      message("Skipping (year missing in df_list): ", lib, " ", year)
      next
    }

    cand_df <- cand_list[[nm]]
    if (!is.data.frame(cand_df) || !"taxon" %in% names(cand_df)) {
      message("Skipping (cand_df missing taxon column): ", nm)
      next
    }

    message("=== Exporting PDF: ", lib, " ", year, " (", nrow(cand_df), " taxa) ===")

    try(
      save_species_maps_pdf(
        df_list   = df_list,
        cand_df   = cand_df,
        year      = year,
        lib_name  = lib,
        base_map  = base_map,
        facet_ncol = facet_ncol,
        out_dir   = out_dir,
        width     = width,
        height    = height
      ),
      silent = FALSE
    )
  }

  invisible(TRUE)
}
```

Run-
```{r}
run_all_species_pdfs(
  lib_lists = lib_lists,
  cand_list = cand_list,
  base_map  = shinnbay_map,
  out_dir   = out_dir,
  facet_ncol = 2,
  width = 11,
  height = 8.5
)


```
**Getting some plotting errors bc don't have coordinates for Canal and Sedge sites from 2021/ 2022**



# Avg temporal and avg spatial trend
For each species, the average abund in time for whole bay alongside a map of avg abund at all stations
Functions:
```{r}
# helper
to_num <- function(x) as.numeric(as.character(x))

# 1) Extract per-sample-per-date-per-station RA
get_taxon_year_points <- function(df_list,
                                  year,
                                  taxon,
                                  tax_col = "Taxon.CommonName",
                                  sample_col = "Sample",
                                  ra_col = "Abundance",
                                  lat_col = c("lat", "Latitude", "Lat"),
                                  lon_col = c("long", "lon", "Longitude", "Long"),
                                  datecode_col = c("Datecode", "datecode"),
                                  date_label_col = c("Date", "date")) {

  year <- as.character(year)
  if (!year %in% names(df_list)) stop("Year not found in df_list: ", year)

  d <- df_list[[year]]$df_ra
  if (is.null(d) || nrow(d) == 0) stop("df_ra is empty for year ", year)

  tax_col        <- pick_col(d, c(tax_col, "Taxon.CommonName", "Taxon", "taxon"))
  sample_col     <- pick_col(d, c(sample_col, "Sample", "sample", "SampleID"))
  ra_col         <- pick_col(d, c(ra_col, "Abundance", "RelAbundance", "RA"))
  lat_col        <- pick_col(d, lat_col)
  lon_col        <- pick_col(d, lon_col)
  datecode_col   <- pick_col(d, datecode_col)
  date_label_col <- pick_col(d, date_label_col, required = FALSE)

  # (A) Universe of ALL samples (one row per Sample x Datecode x Station)
  d_samples <- d %>%
    transmute(
      .samp = as.character(.data[[sample_col]]),
      .dc   = as.integer(.data[[datecode_col]]),
      .lat  = to_num(.data[[lat_col]]),
      .lon  = to_num(.data[[lon_col]]),
      .date = if (!is.null(date_label_col) && date_label_col %in% names(d)) {
        as.character(.data[[date_label_col]])
      } else {
        NA_character_
      }
    ) %>%
    distinct(.samp, .dc, .lat, .lon, .date)

  # (B) Taxon RA per sample/date/station (sum across OTUs mapping to same taxon)
  d_tax_present <- d %>%
    transmute(
      .tax  = as.character(.data[[tax_col]]),
      .samp = as.character(.data[[sample_col]]),
      .dc   = as.integer(.data[[datecode_col]]),
      ra    = to_num(.data[[ra_col]])
    ) %>%
    filter(.tax == taxon) %>%
    group_by(.samp, .dc, .tax) %>%
    summarize(ra = sum(ra, na.rm = TRUE), .groups = "drop")

  # (C) Left join onto the sample universe; missing taxon rows become 0
  d_tax <- d_samples %>%
    left_join(d_tax_present, by = c(".samp", ".dc")) %>%
    mutate(
      .tax = taxon,
      ra   = dplyr::coalesce(ra, 0)
    )

  # If taxon truly never appears in the entire year, keep zeros but warn
  if (all(d_tax$ra == 0, na.rm = TRUE)) {
    message("Note: taxon '", taxon, "' has RA=0 for all samples in year ", year, ".")
  }

  # real Date from Datecode (correct ordering)
  d_tax <- d_tax %>%
    mutate(
      .datecode_str = sprintf("%08d", .dc),
      .date_real    = as.Date(.datecode_str, format = "%Y%m%d")
    )

  d_tax
}

# 2) Left panel: average RA across ALL samples for each sampling date
#    (includes zeros because d_tax has one row per sample/date)
plot_taxon_time_trend <- function(d_tax,
                                  taxon,
                                  year,
                                  colors = myColors_Sci_comname,
                                  aspect_ratio = 1) {

  tax_color <- get_db_color(taxon, default = "grey60")

  d_time <- d_tax %>%
    group_by(.date_real, .dc) %>%
    summarize(
      mean_ra = mean(ra, na.rm = TRUE),     # zeros included
      n_samples = dplyr::n(),
      .groups = "drop"
    ) %>%
    arrange(.date_real)

  ggplot(d_time, aes(x = .date_real, y = mean_ra)) +
    geom_line(color = tax_color, linewidth = 1) +
    geom_point(color = tax_color, size = 2) +
    scale_x_date(date_breaks = "1 month", date_labels = "%b") +
    labs(
      title = "Mean relative abundance per sampling date",
      x = NULL,
      y = "Mean RA"
    ) +
    theme_minimal() +
    theme(
      plot.title   = element_text(face = "bold", size = 12),
      axis.text.x  = element_text(vjust = 0.5),
      aspect.ratio = aspect_ratio
    )
}

# 3) Right panel: annual mean RA per station (lon/lat)
plot_taxon_annual_map <- function(d_tax,
                                  taxon,
                                  year,
                                  base_map,
                                  bbox = c(xmin=-72.61, xmax=-72.4, ymin=40.79, ymax=40.9),
                                  colors = myColors_Sci_comname,
                                  point_size_range = c(1.5, 7),
                                  alpha_points = 0.9,
                                  map_min_ra = 1e-6) {  # <- threshold to hide zeros

  tax_color <- get_db_color(taxon, default = "grey60")


  d_station <- d_tax %>%
    filter(!is.na(.lon), !is.na(.lat)) %>%
    group_by(.lon, .lat) %>%
    summarize(
      mean_ra = mean(ra, na.rm = TRUE),  # includes zeros across samples (correct)
      n = dplyr::n(),
      .groups = "drop"
    ) %>%
    # clip to bbox
    filter(
      .lon >= bbox["xmin"], .lon <= bbox["xmax"],
      .lat >= bbox["ymin"], .lat <= bbox["ymax"]
    ) %>%
    filter(!is.na(mean_ra), mean_ra > map_min_ra)

  base_map +
    geom_point(
      data = d_station,
      aes(x = .lon, y = .lat, size = mean_ra),
      color = tax_color,
      alpha = alpha_points
    ) +
    scale_size_continuous(range = point_size_range, name = "Mean RA") +
    coord_sf(
      xlim = c(bbox["xmin"], bbox["xmax"]),
      ylim = c(bbox["ymin"], bbox["ymax"]),
      expand = FALSE
    ) +
    labs(title = "Mean relative abundance per station") +
    guides(
      size = guide_legend(
        title.position = "top",
        title.hjust = 0.5,
        override.aes = list(color = "black", alpha = 1)
      )
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 12),
      legend.position = "bottom",
      axis.title = element_blank(),
      axis.text  = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
}


# 4) Dual panel: time trend | annual map
plot_species_year_dual <- function(df_list,
                                   year,
                                   taxon,
                                   base_map,
                                   bbox = c(xmin=-72.61, xmax=-72.4, ymin=40.79, ymax=40.9),
                                   colors = myColors_Sci_comname,
                                   tax_col = "Taxon.Display") {

  d_tax <- get_taxon_year_points(df_list, year, taxon, tax_col = tax_col)
  
  p_left <- plot_taxon_time_trend(
    d_tax, taxon, year,
    colors = colors,
    aspect_ratio = 0.5
  )

  p_right <- plot_taxon_annual_map(
    d_tax, taxon, year, base_map,
    bbox = bbox, colors = colors
  )

  p_right_wrapped <- patchwork::wrap_elements(full = p_right)

  (p_left | p_right_wrapped) +
    patchwork::plot_layout(widths = c(1.4, 1)) +
    patchwork::plot_annotation(
      title = paste0(taxon, " — ", year),
      theme = theme(plot.title = element_text(face = "bold", size = 14))
    )
}

```

Run example
```{r, fig.width=16, fig.height=9}
plot_species_year_dual(
  df_list  = co1_sample_dfs,
  year     = 2024,
  taxon    = cand_co1$cand_2024_co1$taxon[1],
  base_map = shinnbay_map,
  tax_col="Taxon.Display"
)

```



Generate pdfs for all years/libraries with a species on every page
Function-
```{r}
# 1) Build candidate taxa tables automatically for each library/year
make_all_candidate_lists <- function(lib_lists,
                                     top_n_by_lib = list(MiFish = 100, Elas02 = 15, CO1 = 15),
                                     min_prev_by_lib = list(MiFish = 0, Elas02 = 0, CO1 = 0)) {

  cand_list <- list()

  for (lib in names(lib_lists)) {
    df_list <- lib_lists[[lib]]
    yrs <- names(df_list)

    top_n <- top_n_by_lib[[lib]] %||% 50
    min_prev <- min_prev_by_lib[[lib]] %||% 0

    for (yr in yrs) {
      message("Candidates: ", lib, " ", yr)
      cand_df <- try(
        candidate_taxa(df_list, year = yr, top_n = top_n, min_prevalence = min_prev),
        silent = TRUE
      )
      if (inherits(cand_df, "try-error")) {
        message("  -> skipped (candidate_taxa failed): ", lib, " ", yr)
        next
      }
      cand_list[[paste0(lib, "_", yr)]] <- cand_df
    }
  }

  cand_list
}


# 2) Save ONE PDF for one library/year: one species per page 
save_dual_species_pdf <- function(df_list,
                                  cand_df,
                                  year,
                                  lib_name,
                                  base_map,
                                  out_dir,
                                  bbox = c(xmin=-72.61, xmax=-72.4, ymin=40.79, ymax=40.9),
                                  width = 16,
                                  height = 9,
                                  max_taxa = Inf) {

  stopifnot(is.data.frame(cand_df), "taxon" %in% names(cand_df))
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

  year_chr <- as.character(year)
  pdf_file <- file.path(out_dir, paste0(lib_name, "_", year_chr, "_dualpanel_species.pdf"))

  taxa <- unique(as.character(cand_df$taxon))
  taxa <- taxa[!is.na(taxa) & taxa != ""]
  if (is.finite(max_taxa)) taxa <- head(taxa, max_taxa)

  pdf(pdf_file, width = width, height = height, onefile = TRUE)
  on.exit(dev.off(), add = TRUE)

  for (tx in taxa) {
    message("Plotting: ", lib_name, " ", year_chr, " | ", tx)

    p <- try(
      plot_species_year_dual(
        df_list  = df_list,
        year     = year_chr,
        taxon    = tx,
        base_map = base_map,
        bbox     = bbox
      ),
      silent = TRUE
    )

    if (inherits(p, "try-error")) {
      # Don't fail the whole PDF—write a simple page explaining what broke
      plot.new()
      title(main = paste0("FAILED: ", lib_name, " ", year_chr, "\n", tx),
            sub = as.character(p))
      next
    }

    print(p)
  }

  message("Saved: ", pdf_file)
  invisible(pdf_file)
}


# 3) Run across ALL libraries/years -
run_all_dual_species_pdfs <- function(lib_lists,
                                      cand_list,
                                      base_map,
                                      out_dir,
                                      bbox = c(xmin=-72.61, xmax=-72.4, ymin=40.79, ymax=40.9),
                                      width = 16,
                                      height = 9,
                                      max_taxa = Inf) {

  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

  for (nm in names(cand_list)) {
    py <- parse_lib_year(nm)
    if (is.null(py)) {
      message("Skipping cand_list item with bad name: ", nm)
      next
      }
    
    lib  <- py$lib
    year <- py$year

    if (!lib %in% names(lib_lists)) {
      message("Skipping (no df_list for lib): ", lib, " [", nm, "]")
      next
    }

    df_list <- lib_lists[[lib]]
    if (!as.character(year) %in% names(df_list)) {
      message("Skipping (year missing in df_list): ", lib, " ", year)
      next
    }

    cand_df <- cand_list[[nm]]
    if (!is.data.frame(cand_df) || !"taxon" %in% names(cand_df)) {
      message("Skipping (cand_df missing taxon column): ", nm)
      next
    }

    message("=== Exporting DUAL PDF: ", lib, " ", year, " (", nrow(cand_df), " taxa) ===")

    save_dual_species_pdf(
      df_list   = df_list,
      cand_df   = cand_df,
      year      = year,
      lib_name  = lib,
      base_map  = base_map,
      out_dir   = out_dir,
      bbox      = bbox,
      width     = width,
      height    = height,
      max_taxa  = max_taxa
    )
  }

  invisible(TRUE)
}
```

Run
```{r}
# set up library lists

lib_lists <- list(
  MiFish       = mifish_sample_dfs,
  Elas02       = elas_sample_dfs,
  CO1          = co1_sample_dfs,
  CO1_NOTTRAWL = co1_nottrawl_sample_dfs
)


# Build candidate lists automatically (uses years embedded in each df_list)
cand_list <- make_all_candidate_lists(
  lib_lists = lib_lists,
  top_n_by_lib = list(MiFish = 1e9, Elas02 = 1e9, CO1 = 1e9, CO1_NOTTRAWL = 1e9)
)

# Export PDFs (one per library-year), one species per page
run_all_dual_species_pdfs(
  lib_lists = lib_lists,
  cand_list = cand_list,
  base_map  = shinnbay_map,
  out_dir   = out_dir,
  bbox      = c(xmin=-72.61, xmax=-72.4, ymin=40.79, ymax=40.9),
  width     = 16,
  height    = 9,
  max_taxa  = Inf
)

```

# Diversity
## Richness
First re-generate plot (from Jia) showing species richness vs # of reads
```{r}
# helper: remove controls (keeps only samples where controls is NA)
drop_controls <- function(ps, control_col = "controls") {
  if (is.null(ps) || nsamples(ps) == 0) return(ps)
  md <- as.data.frame(sample_data(ps))
  if (!control_col %in% names(md)) return(ps)

  keep_ids <- rownames(md)[is.na(md[[control_col]])]
  prune_samples(keep_ids, ps)
}

# helper: per-sample summary (reads + richness)
summarize_reads_taxa_by_sample <- function(ps,
                                          year = NA,
                                          drop_ctrl = TRUE,
                                          control_col = "controls") {
  if (is.null(ps) || nsamples(ps) == 0 || ntaxa(ps) == 0) {
    return(tibble(
      year = as.character(year),
      Sample = character(0),
      total_reads = numeric(0),
      n_taxa = integer(0)
    ))
  }

  if (drop_ctrl) ps <- drop_controls(ps, control_col = control_col)
  if (is.null(ps) || nsamples(ps) == 0 || ntaxa(ps) == 0) {
    return(tibble(
      year = as.character(year),
      Sample = character(0),
      total_reads = numeric(0),
      n_taxa = integer(0)
    ))
  }

  otu <- as(otu_table(ps), "matrix")
  if (!taxa_are_rows(ps)) otu <- t(otu)   # taxa x samples

  reads_by_sample <- colSums(otu, na.rm = TRUE)
  taxa_by_sample  <- colSums(otu > 0, na.rm = TRUE)

  tibble(
    year = as.character(year),
    Sample = names(reads_by_sample),
    total_reads = as.numeric(reads_by_sample),
    n_taxa = as.integer(taxa_by_sample)
  )
}

# make a per-library per-sample table from a named list of phyloseq objects
summarize_library_by_sample <- function(phylo_list,
                                        library_name = "library",
                                        drop_ctrl = TRUE,
                                        control_col = "controls") {
  yrs <- names(phylo_list)

  purrr::map_dfr(yrs, function(yr) {
    summarize_reads_taxa_by_sample(
      ps = phylo_list[[yr]],
      year = yr,
      drop_ctrl = drop_ctrl,
      control_col = control_col
    )
  }) %>%
    mutate(
      library = library_name,
      year = factor(year, levels = sort(unique(year)))
    ) %>%
    filter(!is.na(total_reads), !is.na(n_taxa))
}

# plotting: one plot per library, points = samples, color = year
plot_reads_vs_taxa_by_sample <- function(df,
                                        title = NULL,
                                        log_x = TRUE,
                                        alpha = 0.8) {
  p <- ggplot(df, aes(x = total_reads, y = n_taxa, color = year)) +
    geom_point(size = 2.4, alpha = alpha) +
    theme_minimal() +
    labs(
      title = title %||% unique(df$library),
      x = "Total sequence reads per sample",
      y = "Species richness (unique taxa per sample)",
      color = "Year"
    )

  if (log_x) p <- p + scale_x_log10()
  p
}

`%||%` <- function(a, b) if (!is.null(a)) a else b

# Provide year levels explicitly so ggplot's default palette maps consistently.
year_levels_all <- function(...) {
  yrs <- sort(unique(unlist(list(...))))
  as.character(yrs[!is.na(yrs)])
}

set_year_factor <- function(df, year_levels) {
  df %>%
    dplyr::mutate(
      year = as.character(year),
      year = factor(year, levels = year_levels)
    )
}

```

Plot
```{r}
# build per-sample tables
df_mifish <- summarize_library_by_sample(phylo_mifish, "MiFish", drop_ctrl = TRUE)
df_elas   <- summarize_library_by_sample(phylo_elas,   "Elas02", drop_ctrl = TRUE)
df_co1    <- summarize_library_by_sample(phylo_co1,    "CO1 trawl-only", drop_ctrl = TRUE)
df_co1tot <- summarize_library_by_sample(phylo_co1_total, "CO1 total", drop_ctrl = TRUE)

# plots (one per library; points are samples; color is year)
p_mifish <- plot_reads_vs_taxa_by_sample(df_mifish, title = "MiFish: richness vs reads (by sample)")
p_elas   <- plot_reads_vs_taxa_by_sample(df_elas,   title = "Elas02: richness vs reads (by sample)")
p_co1    <- plot_reads_vs_taxa_by_sample(df_co1,    title = "CO1 (trawl-only): richness vs reads (by sample)")
p_co1tot <- plot_reads_vs_taxa_by_sample(df_co1tot, title = "CO1 (total): richness vs reads (by sample)")

p_mifish
p_elas
p_co1
p_co1tot


```

--> Correlation between read depth and richness HOWEVER at the same sequencing depth, there does seem to be more richness in later years...


Save
```{r}
ggsave(file.path(out_dir, "reads_vs_taxa_mifish.jpg"), p_mifish, width = 6.5, height = 4.5, dpi = 300)
ggsave(file.path(out_dir, "reads_vs_taxa_elas02.jpg"), p_elas,   width = 6.5, height = 4.5, dpi = 300)
ggsave(file.path(out_dir, "reads_vs_taxa_co1_trawl.jpg"), p_co1, width = 6.5, height = 4.5, dpi = 300)
ggsave(file.path(out_dir, "reads_vs_taxa_co1_total.jpg"), p_co1tot, width = 6.5, height = 4.5, dpi = 300)

```



### Richness after rarefy correction

Function to implement rrarefy in vegan
```{r}
rarefy_richness_montecarlo <- function(mat,
                                       depth,
                                       n_iter = 1000,
                                       seed = 123,
                                       drop_ctrl = FALSE) {
  # mat: samples × taxa raw count matrix
  # depth: rarefaction depth
  # n_iter: number of Monte Carlo draws
  # seed: random seed

  if (is.null(mat) || nrow(mat) == 0) return(NULL)
  
  # original (unrarefied) metrics
  orig_depth    <- rowSums(mat)
  orig_richness <- rowSums(mat > 0)

  # ensure we have enough reads to rarefy
  keep_samples <- orig_depth >= depth
  mat <- mat[keep_samples, , drop = FALSE]
  if (nrow(mat) == 0) return(NULL)

  set.seed(seed)

  # Container for richness results
  richness_mat <- matrix(
    NA_integer_,
    nrow = nrow(mat),
    ncol = n_iter,
    dimnames = list(rownames(mat), NULL)
  )

  # Rarefy once per iteration for all samples
  for (i in seq_len(n_iter)) {
    # vegan::rrarefy rarefies all rows at once
    rar_mat <- vegan::rrarefy(mat, sample = depth)  
    richness_mat[, i] <- rowSums(rar_mat > 0)
  }

  # Summaries
  tibble(
    Sample              = rownames(richness_mat),
    orig_depth          = orig_depth[keep_samples],
    orig_richness       = orig_richness[keep_samples],
    richness_rarefied   = rowMeans(richness_mat),
    richness_sd         = apply(richness_mat, 1, sd),
    richness_q025       = apply(richness_mat, 1, quantile, probs = 0.025),
    richness_q975       = apply(richness_mat, 1, quantile, probs = 0.975)
    )

}


iterate_library_year <- function(phylo_list,
                                 library_name,
                                 depth = 1e5,
                                 n_iter = 1000,
                                 seed = 123,
                                 control_col = "controls") {

  out <- list()

  for (yr in names(phylo_list)) {

    ps <- phylo_list[[yr]]
    if (is.null(ps)) next

    # drop controls
    ps_clean <- drop_controls(ps, control_col = control_col)
    if (is.null(ps_clean) || nsamples(ps_clean) == 0) next

    # extract sample x taxa matrix
    mat <- get_sample_taxa_matrix(ps_clean)
    if (is.null(mat) || nrow(mat) == 0) next

    # rarefy
    df_rich <- rarefy_richness_montecarlo(
      mat,
      depth  = depth,
      n_iter = n_iter,
      seed   = seed
    )

    if (is.null(df_rich) || nrow(df_rich) == 0) next

    # minimal, safe metadata
    md <- as.data.frame(sample_data(ps_clean))
    md$Sample <- rownames(md)
    keep_cols <- intersect(
      colnames(md),
      c("Sample", "sites", "Station", "Month", "Bayside")
      )
    
    md <- md[, keep_cols, drop = FALSE]
    md <- tibble::as_tibble(md)


    # join + annotate
    df_rich <- df_rich %>%
      dplyr::left_join(md, by = "Sample") %>%
      dplyr::mutate(
        library = library_name,
        year    = as.integer(yr)
      )

    out[[yr]] <- df_rich
  }

  dplyr::bind_rows(out)
}





```




Run
(Skipping the filtered CO1/ trawl species only dataset bc those reads are all less than 500 each since they were dominated by phytoplankton etc and those all were filted out)


```{r}
# Run for your libraries
df_mifish_rarefied <- iterate_library_year(
  phylo_mifish, "MiFish",
  depth = 1e5, n_iter = 100, seed = 123
)

df_elas_rarefied <- iterate_library_year(
  phylo_elas, "Elas02",
  depth = 1e5, n_iter = 100, seed = 123
)

df_co1tot_rarefied <- iterate_library_year(
  phylo_co1_total, "CO1 total",
  depth = 1e5, n_iter = 100, seed = 123
)

df_rarefied_all <- bind_rows(
  df_mifish_rarefied,
  df_elas_rarefied,
  df_co1tot_rarefied
)
```

Warning is ok- this is just vegan flagging that the min abundance (greater than 0) was not 1, but 500, 1300, etc. That's because these are pre-filtered datasets


Plot raw seq depth richness vs rarefaction correction fichness
```{r}
#single year order across ALL libraries (to matcfh earlier plot)
year_levels <- year_levels_all(df_mifish$year, df_elas$year, df_co1tot$year)

# prep rarefied df with consistent year factor
df_rarefied_all2 <- df_rarefied_all %>%
  dplyr::mutate(
    library = as.character(library),
    year    = as.character(year)
  ) %>%
  set_year_factor(year_levels)

# plotting function: old vs rarefied richness
plot_old_vs_rarefied <- function(df, title = NULL, alpha = 0.65) {
  ggplot(df, aes(x = orig_richness, y = richness_rarefied, color = year)) +
    geom_point(alpha = alpha, size = 2.2) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    theme_minimal() +
    labs(
      title = title %||% unique(df$library),
      x = "Original richness (unrarefied)",
      y = "Rarefied richness (mean over iterations)",
      color = "Year"
    )
}

# make 3 separate plots
p_oldnew_mifish <- df_rarefied_all2 %>%
  dplyr::filter(library == "MiFish") %>%
  plot_old_vs_rarefied(title = "MiFish: original vs rarefied richness")

p_oldnew_elas <- df_rarefied_all2 %>%
  dplyr::filter(library == "Elas02") %>%
  plot_old_vs_rarefied(title = "Elas02: original vs rarefied richness")

p_oldnew_co1tot <- df_rarefied_all2 %>%
  dplyr::filter(library == "CO1 total") %>%
  plot_old_vs_rarefied(title = "CO1 total: original vs rarefied richness")

p_oldnew_mifish
p_oldnew_elas
p_oldnew_co1tot

```

Export plots
```{r}
# rarefaction comparison plots
ggsave(
  filename = file.path(out_dir, "richness_original_vs_rarefied_mifish.jpg"),
  plot = p_oldnew_mifish,
  width = 6.5, height = 4.5, units = "in", dpi = 300
)

ggsave(
  filename = file.path(out_dir, "richness_original_vs_rarefied_elas02.jpg"),
  plot = p_oldnew_elas,
  width = 6.5, height = 4.5, units = "in", dpi = 300
)

ggsave(
  filename = file.path(out_dir, "richness_original_vs_rarefied_co1_total.jpg"),
  plot = p_oldnew_co1tot,
  width = 6.5, height = 4.5, units = "in", dpi = 300
)

```

Export table
```{r}
readr::write_csv(
  df_rarefied_all2,
  file.path(out_dir, "rarefied_richness_per_sample.csv")
)

```


## CONTINUE HERE 1/21/26
# NEED TO DO RAREFACTION CORRECTION BEFORE DIVERSITY BELOW. WHAT ABOUT ABOVE?
make plots explaining rarefaction correction too, comparing diversity before and after
# FIGURE OUT CRANGON- phylo trees?





Species richness by month and habitat types
- In June, eel grass stations were not sampled but all other habitat types were. So any zeroes in June are real for all habitat types except eel grass---
```{r}
# First: define all months and habitats you'd expect
all_months <- unique(c(mifish_df$Month, elas_df$Month))
all_habitats <- unique(c(mifish_df$Habitat, elas_df$Habitat))

# MiFish
species_counts_2024_mifish <- mifish_df %>%
  group_by(Habitat, Month) %>%
  summarise(UniqueSpeciesCount = n_distinct(`Taxon.CommonName`), .groups = "drop") %>%
  complete(Habitat = all_habitats, Month = all_months, fill = list(UniqueSpeciesCount = 0)) %>%
  filter(!(Habitat == "EELGRASS" & Month == "6"))


# Elas
species_counts_2024_elas <- elas_df %>%
  group_by(Habitat, Month) %>%
  summarise(UniqueSpeciesCount = n_distinct(`Taxon.CommonName`), .groups = "drop") %>%
  complete(Habitat = all_habitats, Month = all_months, fill = list(UniqueSpeciesCount = 0)) %>%
  filter(!(Habitat == "EELGRASS" & Month == "6"))

species_counts_2024_mifish
species_counts_2024_elas

```

#### Mifish
```{r}

sp_richnesss_habitat_month_mifish_plot <- ggplot(species_counts_2024_mifish, aes(x = Month, y = UniqueSpeciesCount, group = Habitat, color = Habitat)) +
  geom_line(size = 1, alpha = 0.8) +
  geom_point(size = 2, alpha = 0.8) +
  labs(x = "Month", y = "Species Richness", title = "2024 Species Richness; MiFish") +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    legend.title = element_blank()
  ) +
  scale_color_manual(values = c("azure2", "chartreuse4", "cadetblue1", "darkgrey"))

sp_richnesss_habitat_month_mifish_plot

# save
ggsave(plot = sp_richnesss_habitat_month_mifish_plot, filename = "figures-2024eDNA/sp_richnesss_habitat_month_mifish_plot.jpg", width = 5, height = 3, units = "in")


```



#### Elas02
```{r}
sp_richnesss_habitat_month_elas_plot <- ggplot(species_counts_2024_elas, aes(x = Month, y = UniqueSpeciesCount, group = Habitat, color = Habitat)) +
  geom_line(size = 1, alpha = 0.8) +
  geom_point(size = 2, alpha = 0.8) +
  labs(x = "Month", y = "Species Richness", title = "2024 Species Richness; Elas02") +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    legend.title = element_blank()
  ) +
  scale_color_manual(values = c("azure2", "chartreuse4", "cadetblue1", "darkgrey"))

sp_richnesss_habitat_month_elas_plot

# save
ggsave(plot = sp_richnesss_habitat_month_elas_plot, filename = "figures-2024eDNA/sp_richnesss_habitat_month_elas_plot.jpg", width = 5, height = 3, units = "in")


```






