---
title: "ShiRP eDNA Time Series- Mifish Elas02 CO1 plots"
output: html_notebook
---


Notebook for making plots from eDNA data (trawl and manual samples only- not expedition)

Import data tables generated with [/MiFish-Elas-CO1-handsamples-2020-2025.nb.html](MiFish-Elas-CO1-handsamples-2020-2025 notebook)

# Load packages

```{r}
# rm(list = ls())
library(tidyverse)
library(phyloseq)
library(vegan)
library(RColorBrewer)
library(microbiome)
library(metagMisc)
library(fantaxtic)
library(ggrepel)
library(paletteer)
library(pals)
library(compositions)
library(hms)
library(ggvegan)
```

#  Import cleand up data tables
## Mifish-
Function that imports cleaned up data tables and turns into phyloseq object for all mifish data, any year:
```{r}
read_matrix_first_col_as_rownames <- function(path) {
  df <- read.csv(path, header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)

  # Use first column as rownames (whatever its name is)
  rn <- df[[1]]
  df <- df[, -1, drop = FALSE]
  rownames(df) <- rn

  as.matrix(df)
}

read_sampledata_first_col_as_rownames <- function(path) {
  df <- read.csv(path, header = TRUE, stringsAsFactors = TRUE, check.names = FALSE)

  rn <- df[[1]]
  df <- df[, -1, drop = FALSE]
  rownames(df) <- rn

  df
}

make_phyloseq_mifish <- function(year, data_dir = "filtered_data") {

  asv_path  <- file.path(data_dir, sprintf("mifish_%d_glommed_asv_table.csv", year))
  tax_path  <- file.path(data_dir, sprintf("mifish_%d_glommed_tax_table.csv", year))
  samp_path <- file.path(data_dir, sprintf("mifish_%d_sample_data.csv", year))

  asv_mat <- read_matrix_first_col_as_rownames(asv_path)

  # Tax table is different: rownames column might not be first; handle by name if present
  tax_df <- read.csv(tax_path, header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
  if ("X1stRepresentative_ASV" %in% names(tax_df)) {
    rownames(tax_df) <- tax_df$X1stRepresentative_ASV
    tax_df <- tax_df[, setdiff(names(tax_df), "X1stRepresentative_ASV"), drop = FALSE]
  } else {
    # fallback: assume first column contains rownames
    rn <- tax_df[[1]]
    tax_df <- tax_df[, -1, drop = FALSE]
    rownames(tax_df) <- rn
  }
  tax_mat <- as.matrix(tax_df)

  samp_df <- read_sampledata_first_col_as_rownames(samp_path)

  phyloseq(
    otu_table(asv_mat, taxa_are_rows = TRUE),
    tax_table(tax_mat),
    sample_data(samp_df)
  )
}

```

Run
```{r}
years <- 2020:2025 #update years as needed
phylo_mifish <- setNames(lapply(years, make_phyloseq_mifish), as.character(years))
```



## Elas02
Function
```{r}
make_phyloseq_elas <- function(year, data_dir = "filtered_data") {

  asv_path  <- file.path(data_dir, sprintf("elas_%d_glommed_asv_table.csv", year))
  tax_path  <- file.path(data_dir, sprintf("elas_%d_glommed_tax_table.csv", year))
  samp_path <- file.path(data_dir, sprintf("elas_%d_sample_data.csv", year))

  # ASV table: first column = rownames (robust to X / ...1 / blank header)
  asv_mat <- read_matrix_first_col_as_rownames(asv_path)

  # Tax table: prefer REVAMP-style column name, otherwise fall back to first column
  tax_df <- read.csv(tax_path, header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
  if ("X1stRepresentative_ASV" %in% names(tax_df)) {
    rownames(tax_df) <- tax_df$X1stRepresentative_ASV
    tax_df <- tax_df[, setdiff(names(tax_df), "X1stRepresentative_ASV"), drop = FALSE]
  } else {
    rn <- tax_df[[1]]
    tax_df <- tax_df[, -1, drop = FALSE]
    rownames(tax_df) <- rn
  }
  tax_mat <- as.matrix(tax_df)

  # Sample metadata: first column = rownames (robust)
  samp_df <- read_sampledata_first_col_as_rownames(samp_path)

  phyloseq(
    otu_table(asv_mat, taxa_are_rows = TRUE),
    tax_table(tax_mat),
    sample_data(samp_df)
  )
}

```

Run
```{r}
elas_years <- c(2021, 2023, 2024, 2025)

phylo_elas <- setNames(
  lapply(elas_years, make_phyloseq_elas),
  as.character(elas_years)
)
```

## CO1
Keep in mind this is filtered to only those species expected to be caughy by trawl (eg. pelagic/ swimming species plus all fish. removed many planktonic and benthic invertebrates)
```{r}
make_phyloseq_co1 <- function(year, data_dir = "filtered_data") {

  asv_path  <- file.path(data_dir, sprintf("co1_%d_glommed_asv_table_fortrawlcomparisononly.csv", year))
  tax_path  <- file.path(data_dir, sprintf("co1_%d_glommed_tax_table_fortrawlcomparisononly.csv", year))
  samp_path <- file.path(data_dir, sprintf("co1_%d_sample_data_fortrawlcomparisononly.csv", year))

  # ASV table (robust to X / ...1 / blank header)
  asv_mat <- read_matrix_first_col_as_rownames(asv_path)

  # Tax table (prefer REVAMP column name; otherwise fall back to first col)
  tax_df <- read.csv(tax_path, header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
  if ("X1stRepresentative_ASV" %in% names(tax_df)) {
    rownames(tax_df) <- tax_df$X1stRepresentative_ASV
    tax_df <- tax_df[, setdiff(names(tax_df), "X1stRepresentative_ASV"), drop = FALSE]
  } else {
    rn <- tax_df[[1]]
    tax_df <- tax_df[, -1, drop = FALSE]
    rownames(tax_df) <- rn
  }
  tax_mat <- as.matrix(tax_df)

  # Sample metadata (robust)
  samp_df <- read_sampledata_first_col_as_rownames(samp_path)

  phyloseq(
    otu_table(asv_mat, taxa_are_rows = TRUE),
    tax_table(tax_mat),
    sample_data(samp_df)
  )
}

```

Run
```{r}
co1_years <- c(2024, 2025)

phylo_co1 <- setNames(
  lapply(co1_years, make_phyloseq_co1),
  as.character(co1_years)
)
```


## Common names
Import common names database and set color scheme
```{r}
commonnames <- read_csv(file = "../eDNA-databases/commonnames.csv")

myColors_sciname <- setNames(commonnames$Color, commonnames$Taxon)
myColors_comname <- setNames(commonnames$Color, commonnames$CommonName)
myColors_Sci_comname <- setNames(commonnames$Color, commonnames$Taxon.CommonName)


head(myColors_sciname)
head(myColors_comname)
head(myColors_Sci_comname)
```

# CONTINUE HERE 1/7/25- KEEP AUTOMATING ANALYSIS. GET TO PLOTTING CO1 BIVALVES AND CRUSTACEANS- MAPS AND TIME SERIES
# Plot positive controls
Note- there are no Elas-2 positive controls, just MiFish

Filter them out into new ps object:
```{r}
# Make new df with positive controls and remove ASVs that are not represented
ps2024_mifish_posctl <- subset_samples(phylo_mifish$`2024`, controls=="positive")

# Use metagMisc to remove ASV rows with 0 abundance for all positive samples
ps2024_mifish_posctl <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps2024_mifish_posctl,
  minabund = 0,
  relabund = FALSE)

# Make a rel abun ps object also
ps2024_mifish_posctl_ra <- microbiome::transform(ps2024_mifish_posctl, transform = "compositional")
```

Plot rel abun in pos ctls
2024
```{r}
mifish_posctl_barplot <- plot_bar(ps2024_mifish_posctl, fill = "Taxon.CommonName", x = "replicates", title = "2024 Manual Samples: Positive Controls") + 
  scale_y_continuous(name="Read Abundance")+
  theme_minimal() +
  scale_fill_manual(values = myColors_Sci_comname) +
  guides(fill = guide_legend(ncol = 2)) +  
  theme(legend.title = element_blank(), axis.text.x = element_text(size = 11), axis.text.y = element_text(size = 10)) +
  scale_x_discrete(labels = function(x) gsub("positive", "", x, ignore.case = TRUE)) +
  labs(x = NULL, y = "Read Abundance", title = NULL)

mifish_posctl_barplot

  

mifish_posctl_barplot_ra <- plot_bar(ps2024_mifish_posctl_ra, fill = "Taxon.CommonName", x = "replicates", title = "2024 Manual Samples: Positive Controls") + 
  scale_y_continuous(name="Relative Abundance")+
  theme_minimal() +
  scale_fill_manual(values = myColors_Sci_comname) +
  guides(fill = guide_legend(ncol = 2)) +  
  theme(legend.title = element_blank(), axis.text.x = element_text(size = 11), axis.text.y = element_text(size = 10)) +
  scale_x_discrete(labels = function(x) gsub("positive", "", x, ignore.case = TRUE)) +
  labs(x = NULL, y = "Relative Abundance", title = NULL)

mifish_posctl_barplot_ra


#ggsave(plot = mifish_posctl_barplot_ra, filename = "figures-2024eDNA/mifish_posctl_barplot_ra.jpg", width = 7, height = 4, units = "in")

```
2025
```{r}
# Make new df with positive controls and remove ASVs that are not represented
ps2025_mifish_posctl <- subset_samples(phylo_mifish$`2025`, controls=="positive")

# Use metagMisc to remove ASV rows with 0 abundance for all positive samples
ps2025_mifish_posctl <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps2025_mifish_posctl,
  minabund = 0,
  relabund = FALSE)

# Make a rel abun ps object also
ps2025_mifish_posctl_ra <- microbiome::transform(ps2025_mifish_posctl, transform = "compositional")
```

```{r}
mifish_posctl_barplot <- plot_bar(ps2025_mifish_posctl, fill = "Taxon.CommonName", x = "replicates", title = "2025 Manual Samples: Positive Controls") + 
  scale_y_continuous(name="Read Abundance")+
  theme_minimal() +
  scale_fill_manual(values = myColors_Sci_comname) +
  guides(fill = guide_legend(ncol = 2)) +  
  theme(legend.title = element_blank(), axis.text.x = element_text(size = 11), axis.text.y = element_text(size = 10)) +
  scale_x_discrete(labels = function(x) gsub("positive", "", x, ignore.case = TRUE)) +
  labs(x = NULL, y = "Read Abundance", title = NULL)

mifish_posctl_barplot

  

mifish_posctl_barplot_ra <- plot_bar(ps2025_mifish_posctl_ra, fill = "Taxon.CommonName", x = "replicates", title = "2025 Manual Samples: Positive Controls") + 
  scale_y_continuous(name="Relative Abundance")+
  theme_minimal() +
  scale_fill_manual(values = myColors_Sci_comname) +
  guides(fill = guide_legend(ncol = 2)) +  
  theme(legend.title = element_blank(), axis.text.x = element_text(size = 11), axis.text.y = element_text(size = 10)) +
  scale_x_discrete(labels = function(x) gsub("positive", "", x, ignore.case = TRUE)) +
  labs(x = NULL, y = "Relative Abundance", title = NULL)

mifish_posctl_barplot_ra


#ggsave(plot = mifish_posctl_barplot_ra, filename = "figures-2025eDNA/mifish_posctl_barplot_ra.jpg", width = 7, height = 4, units = "in")

```



## Plots
### Prepare data
Remove pos controls, glom by Species.CommonName, and convert to dataframe
#### Mifish
```{r}
phylo_2024_mifish_glommed <- tax_glom(phylo_2024_mifish, "Species.CommonName")
mifish_df <- subset_samples(phylo_2024_mifish_glommed, !controls %in% c("positive"))
mifish_df <- psmelt(mifish_df)

#relative abund ps object
phylo_2024_mifish_glommed_ra <- tax_glom(phylo_2024_mifish, "Species.CommonName")
mifish_df_ra <- subset_samples(phylo_2024_mifish_glommed_ra, !controls %in% c("positive")) 
mifish_df_ra <- microbiome::transform(mifish_df_ra, transform = "compositional") 
mifish_df_ra <- psmelt(mifish_df_ra)

# replace zeroes in the table with NA
mifish_df[mifish_df == 0] <- NA
mifish_df_ra[mifish_df_ra == 0] <- NA

# and remove rows with NAs (so they don't appear as small dots in plot)
mifish_df <-  filter(mifish_df, !is.na(Abundance))
mifish_df_ra <-  filter(mifish_df_ra, !is.na(Abundance))


# and view
mifish_df
mifish_df_ra
```


#### Elas
```{r}
phylo_2024_elas_glommed <- tax_glom(phylo_2024_elas, "Species.CommonName")
elas_df <- subset_samples(phylo_2024_elas_glommed, !controls %in% c("positive"))
elas_df <- psmelt(elas_df)

#relative abund ps object
phylo_2024_elas_glommed_ra <- tax_glom(phylo_2024_elas, "Species.CommonName")
elas_df_ra <- subset_samples(phylo_2024_elas_glommed_ra, !controls %in% c("positive")) 
elas_df_ra <- microbiome::transform(elas_df_ra, transform = "compositional") 
elas_df_ra <- psmelt(elas_df_ra)

# replace zeroes in the table with NA
elas_df[elas_df == 0] <- NA
elas_df_ra[elas_df_ra == 0] <- NA

# and remove rows with NAs (so they don't appear as small dots in plot)
elas_df <-  filter(elas_df, !is.na(Abundance))
elas_df_ra <-  filter(elas_df_ra, !is.na(Abundance))


# and view
elas_df
elas_df_ra
```

### Prepare for plot

#### Remove controls from dataframes
```{r}
mifish_df <- mifish_df %>% filter(is.na(controls))
mifish_df_ra <- mifish_df_ra %>% filter(is.na(controls))

elas_df <- elas_df %>% filter(is.na(controls))
elas_df_ra <- elas_df_ra %>% filter(is.na(controls))
```


#### Sort levels

```{r}
# sort station numbers
sitelevels <- as.character(sort(as.numeric(unique(c(mifish_df$sites)))))
sitelevels

# put east/west in geographic/sensible order (not alphabetic)
mifish_df$Bayside_f= factor(mifish_df$Bayside, levels=c('WEST','EAST'))
mifish_df_ra$Bayside_f= factor(mifish_df_ra$Bayside, levels=c('WEST','EAST'))
elas_df$Bayside_f= factor(elas_df$Bayside, levels=c('WEST','EAST'))
elas_df_ra$Bayside_f= factor(elas_df_ra$Bayside, levels=c('WEST','EAST'))

# put months in time/sensible order (not alphabetic)
mifish_df$Month_f= factor(mifish_df$Month, levels=sort(unique(mifish_df$Month)))
mifish_df_ra$Month_f= factor(mifish_df_ra$Month, levels=sort(unique(mifish_df$Month)))
elas_df$Month_f= factor(elas_df$Month, levels=sort(unique(mifish_df$Month)))
elas_df_ra$Month_f= factor(elas_df_ra$Month, levels=sort(unique(mifish_df$Month)))

```

### Mifish Bubble Plots
#### By bayside
```{r}
# First average replicates at same sites
mifish_df_bayside <- mifish_df %>%
    group_by(sites, Bayside_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
mifish_df_bayside


bubbleplot_mifish_Bayside <- ggplot(mifish_df_bayside, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 11), 
        axis.text.y = element_text(size = 11), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Bayside_f, scales = "free", space = "free", drop= TRUE)
bubbleplot_mifish_Bayside
  
# save
ggsave(plot = bubbleplot_mifish_Bayside, filename = "figures-2024eDNA/bubbleplot_mifish_Bayside.jpg", width = 9, height = 8, units = "in")
```

#### By month
```{r}
# First average replicates from same month
mifish_df_month <- mifish_df %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
mifish_df_month


bubbleplot_mifish_Month <- ggplot(mifish_df_month, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_wrap(~Month_f, drop= TRUE, nrow = 1)  
bubbleplot_mifish_Month
  
# save
ggsave(plot = bubbleplot_mifish_Month, filename = "figures-2024eDNA/bubbleplot_mifish_Month.jpg", width = 22, height = 8, units = "in")
```

#### By month, trawl sites only

```{r}
# First average replicates from same month
mifish_df_month_trawl <- mifish_df %>%
  filter(sites %in% c(1,2,3,4,5,6,7,8,9,10,11,12,13)) %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
mifish_df_month_trawl


bubbleplot_mifish_Month_trawlonly <- ggplot(mifish_df_month_trawl, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_wrap(~Month_f, drop= TRUE, nrow = 1)  
bubbleplot_mifish_Month_trawlonly
  
# save
ggsave(plot = bubbleplot_mifish_Month_trawlonly, filename = "figures-2024eDNA/bubbleplot_mifish_Month_trawlonly.jpg", width = 12, height = 8, units = "in")
```



#### By month, other sites

First checked labels and "SEA GRASS" AND "EEL GRASS" should be the same. Fix
```{r}
#check before
unique(mifish_df$Habitat)

mifish_df <- mifish_df %>%
  mutate(Habitat = case_when(
    Habitat == "SEAGRASS" ~ "EELGRASS",
    TRUE ~ Habitat  # Keep all other values unchanged
  ))

#and after
unique(mifish_df$Habitat)
```

Plot by othet habitat types:
Eelgrass
```{r}
# First average replicates from same month
mifish_df_month_eelgrass <- mifish_df %>%
  filter(Habitat == "EELGRASS") %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
mifish_df_month_eelgrass


bubbleplot_mifish_Month_eelgrass <- ggplot(mifish_df_month_eelgrass, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Month_f, drop = TRUE, scales = "free", space='free')
bubbleplot_mifish_Month_eelgrass

# save
ggsave(plot = bubbleplot_mifish_Month_eelgrass, filename = "figures-2024eDNA/bubbleplot_mifish_Month_eelgrass.jpg", width = 9, height = 5, units = "in")
```


Clam Sanctuary
```{r}
# First average replicates from same month
mifish_df_month_clamsanct <- mifish_df %>%
  filter(Habitat == "CLAM SANCTUARY") %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
mifish_df_month_clamsanct


bubbleplot_mifish_Month_clamsanct <- ggplot(mifish_df_month_clamsanct, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Month_f, drop = TRUE, scales = "free", space='free')
bubbleplot_mifish_Month_clamsanct

# save
ggsave(plot = bubbleplot_mifish_Month_clamsanct, filename = "figures-2024eDNA/bubbleplot_mifish_Month_clamsanct.jpg", width = 7, height = 3.5, units = "in")
```


Oyster reef
```{r}
# First average replicates from same month
mifish_df_month_oysterreef <- mifish_df %>%
  filter(Habitat == "OYSTER REEF") %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
mifish_df_month_oysterreef


bubbleplot_mifish_Month_oysterreef <- ggplot(mifish_df_month_oysterreef, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Month_f, drop = TRUE, scales = "free")
bubbleplot_mifish_Month_oysterreef

# save
ggsave(plot = bubbleplot_mifish_Month_oysterreef, filename = "figures-2024eDNA/bubbleplot_mifish_Month_oysterreef.jpg", width = 9, height = 4.5, units = "in")
```



### MiFish- Species Time Series

#### Trawl sites

```{r}
# First average replicates from same month and across all trawl sites
mifish_df_month_trawls <- mifish_df %>%
  filter(site_altname %in% c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13)) %>%
  group_by(Month_f, CommonName, `Species.CommonName`) %>%
  dplyr::summarize(MeanAbundance = mean(Abundance)) %>%
  mutate(Month_f = factor(Month_f, levels = 6:9, labels = c("Jun", "Jul", "Aug", "Sep"), ordered = TRUE))
mifish_df_month_trawls

# take only top species
top20_species <- mifish_df_month_trawls %>%
  group_by(`Species.CommonName`) %>%
  summarise(TotalAbundance = sum(MeanAbundance, na.rm = TRUE)) %>%
  arrange(desc(TotalAbundance)) %>%
  slice_head(n = 20) %>%
  pull(`Species.CommonName`)


mifish_df_month_top20 <- mifish_df_month_trawls %>%
  filter(`Species.CommonName` %in% top20_species) %>%
  ungroup() %>%  
  complete(
    Month_f,
    nesting(`Species.CommonName`, CommonName),
    fill = list(MeanAbundance = 0)
  ) %>%
  mutate(FacetLabel = str_wrap(`Species.CommonName`, width = 20))

timeseries_mifish_Month <- 
ggplot(mifish_df_month_top20) + 
  geom_area(aes(x = Month_f, y = MeanAbundance, fill = `Species.CommonName`, group = 1), alpha = 0.7) +
  facet_wrap(~FacetLabel, scales = "free_y") +
  xlab("") +
  scale_fill_manual(values = myColors_Sci_comname) +
  theme_minimal() +
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10),
        legend.position = "none") 

timeseries_mifish_Month
  
# save
ggsave(plot = timeseries_mifish_Month, filename = "figures-2024eDNA/mifish_topspecies_timeseries.jpg", width = 11, height = 8, units = "in")
```

#### Clam Sites
```{r}
# First average replicates from same month and across all trawl sites
mifish_df_month_clamsites <- mifish_df %>%
  filter(sites %in% c(23, 24)) %>%
  group_by(Month_f, CommonName, `Species.CommonName`) %>%
  dplyr::summarize(MeanAbundance = mean(Abundance)) %>%
  mutate(Month_f = factor(Month_f, levels = 6:9, labels = c("Jun", "Jul", "Aug", "Sep"), ordered = TRUE))
mifish_df_month_clamsites

# take only top species
top12_species <- mifish_df_month_clamsites %>%
  group_by(`Species.CommonName`) %>%
  summarise(TotalAbundance = sum(MeanAbundance, na.rm = TRUE)) %>%
  arrange(desc(TotalAbundance)) %>%
  slice_head(n = 12) %>%
  pull(`Species.CommonName`)


mifish_df_month_top12 <- mifish_df_month_clamsites %>%
  filter(`Species.CommonName` %in% top12_species) %>%
  ungroup() %>%  
  complete(
    Month_f,
    nesting(`Species.CommonName`, CommonName),
    fill = list(MeanAbundance = 0)
  ) %>%
  mutate(FacetLabel = str_wrap(`Species.CommonName`, width = 20))

timeseries_mifish_Month_clamsites <- 
ggplot(mifish_df_month_top12) + 
  geom_area(aes(x = Month_f, y = MeanAbundance, fill = `Species.CommonName`, group = 1), alpha = 0.7) +
  facet_wrap(~FacetLabel, scales = "free_y") +
  xlab("") +
  scale_fill_manual(values = myColors_Sci_comname) +
  theme_minimal() +
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10),
        legend.position = "none") 

timeseries_mifish_Month_clamsites
  
# save
ggsave(plot = timeseries_mifish_Month_clamsites, filename = "figures-2024eDNA/mifish_topspecies_timeseries_clamsites.jpg", width = 8.8, height = 6, units = "in")
```

#### Oyster Sites
```{r}
# First average replicates from same month and across all trawl sites
mifish_df_month_oystersites <- mifish_df %>%
  filter(sites %in% c(25, 26, 31, 32, 33, 34)) %>%
  group_by(Month_f, CommonName, `Species.CommonName`) %>%
  dplyr::summarize(MeanAbundance = mean(Abundance)) %>%
  mutate(Month_f = factor(Month_f, levels = 6:9, labels = c("Jun", "Jul", "Aug", "Sep"), ordered = TRUE))
mifish_df_month_oystersites

# take only top species
top12_species <- mifish_df_month_oystersites %>%
  group_by(`Species.CommonName`) %>%
  summarise(TotalAbundance = sum(MeanAbundance, na.rm = TRUE)) %>%
  arrange(desc(TotalAbundance)) %>%
  slice_head(n = 12) %>%
  pull(`Species.CommonName`)


mifish_df_month_top12 <- mifish_df_month_oystersites %>%
  filter(`Species.CommonName` %in% top12_species) %>%
  ungroup() %>%  
  complete(
    Month_f,
    nesting(`Species.CommonName`, CommonName),
    fill = list(MeanAbundance = 0)
  ) %>%
  mutate(FacetLabel = str_wrap(`Species.CommonName`, width = 20))

timeseries_mifish_Month_oystersites <- 
ggplot(mifish_df_month_top12) + 
  geom_area(aes(x = Month_f, y = MeanAbundance, fill = `Species.CommonName`, group = 1), alpha = 0.7) +
  facet_wrap(~FacetLabel, scales = "free_y") +
  xlab("") +
  scale_fill_manual(values = myColors_Sci_comname) +
  theme_minimal() +
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10),
        legend.position = "none") 

timeseries_mifish_Month_oystersites
  
# save
ggsave(plot = timeseries_mifish_Month_oystersites, filename = "figures-2024eDNA/mifish_topspecies_timeseries_oystersites.jpg", width = 8.8, height = 6, units = "in")
```

#### Eelgrass Sites
```{r}
# First average replicates from same month and across all trawl sites
mifish_df_month_eelgrasssites <- mifish_df %>%
  filter(sites %in% c(27, 28, 29, 30)) %>%
  group_by(Month_f, CommonName, `Species.CommonName`) %>%
  dplyr::summarize(MeanAbundance = mean(Abundance)) %>%
  mutate(Month_f = factor(Month_f, levels = 6:9, labels = c("Jun", "Jul", "Aug", "Sep"), ordered = TRUE))
mifish_df_month_eelgrasssites

# take only top species
top12_species <- mifish_df_month_eelgrasssites %>%
  group_by(`Species.CommonName`) %>%
  summarise(TotalAbundance = sum(MeanAbundance, na.rm = TRUE)) %>%
  arrange(desc(TotalAbundance)) %>%
  slice_head(n = 12) %>%
  pull(`Species.CommonName`)


mifish_df_month_top12 <- mifish_df_month_eelgrasssites %>%
  filter(`Species.CommonName` %in% top12_species) %>%
  ungroup() %>%  
  complete(
    Month_f,
    nesting(`Species.CommonName`, CommonName),
    fill = list(MeanAbundance = 0)
  ) %>%
  mutate(FacetLabel = str_wrap(`Species.CommonName`, width = 20))

timeseries_mifish_Month_eelgrasssites <- 
ggplot(mifish_df_month_top12) + 
  geom_area(aes(x = Month_f, y = MeanAbundance, fill = `Species.CommonName`, group = 1), alpha = 0.7) +
  facet_wrap(~FacetLabel, scales = "free_y") +
  xlab("") +
  scale_fill_manual(values = myColors_Sci_comname) +
  theme_minimal() +
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10),
        legend.position = "none") 

timeseries_mifish_Month_eelgrasssites
  
# save
ggsave(plot = timeseries_mifish_Month_eelgrasssites, filename = "figures-2024eDNA/mifish_topspecies_timeseries_eelgrasssites.jpg", width = 8.8, height = 6, units = "in")
```

### Elas02 Bubble Plots
#### By bayside
```{r}
# First average replicates at same sites
elas_df_bayside <- elas_df %>%
    group_by(sites, Bayside_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
elas_df_bayside


bubbleplot_elas_Bayside <- ggplot(elas_df_bayside, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 9), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Bayside_f, scales = "free", space = "free", drop= TRUE)
bubbleplot_elas_Bayside
  
# save
ggsave(plot = bubbleplot_elas_Bayside, filename = "figures-2024eDNA/bubbleplot_elas_Bayside.jpg", width = 10, height = 3, units = "in")
```


#### By month
```{r}
# First average replicates from same month
elas_df_month <- elas_df %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
elas_df_month


bubbleplot_elas_Month <- ggplot(elas_df_month, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_wrap(~Month_f, drop= TRUE, nrow = 1)  
bubbleplot_elas_Month
  
# save
ggsave(plot = bubbleplot_elas_Month, filename = "figures-2024eDNA/bubbleplot_elas_Month.jpg", width = 22, height = 3, units = "in")
```

#### By month, trawl sites only

```{r}
# First average replicates from same month
elas_df_month_trawl <- elas_df %>%
  filter(sites %in% c(1,2,3,4,5,6,7,8,9,10,11,12,13)) %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
elas_df_month_trawl


bubbleplot_elas_Month_trawlonly <- ggplot(elas_df_month_trawl, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_wrap(~Month_f, drop= TRUE, nrow = 1)  
bubbleplot_elas_Month_trawlonly
  
# save
ggsave(plot = bubbleplot_elas_Month_trawlonly, filename = "figures-2024eDNA/bubbleplot_elas_Month_trawlonly.jpg", width = 11, height = 2.5, units = "in")
```



#### By month, other sites

First checked labels and "SEA GRASS" AND "EEL GRASS" should be the same. Fix
```{r}
#check before
unique(elas_df$Habitat)

elas_df <- elas_df %>%
  mutate(Habitat = case_when(
    Habitat == "SEAGRASS" ~ "EELGRASS",
    TRUE ~ Habitat  # Keep all other values unchanged
  ))

#and after
unique(elas_df$Habitat)
```

Plot by othet habitat types:
Eelgrass
```{r}
# First average replicates from same month
elas_df_month_eelgrass <- elas_df %>%
  filter(Habitat == "EELGRASS") %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
elas_df_month_eelgrass


bubbleplot_elas_Month_eelgrass <- ggplot(elas_df_month_eelgrass, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Month_f, drop = TRUE, scales = "free", space = "free")
bubbleplot_elas_Month_eelgrass

# save
ggsave(plot = bubbleplot_elas_Month_eelgrass, filename = "figures-2024eDNA/bubbleplot_elas_Month_eelgrass.jpg", width = 9, height = 2.5, units = "in")
```


Clam Sanctuary
```{r}
# First average replicates from same month
elas_df_month_clamsanct <- elas_df %>%
  filter(Habitat == "CLAM SANCTUARY") %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
elas_df_month_clamsanct


bubbleplot_elas_Month_clamsanct <- ggplot(elas_df_month_clamsanct, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Month_f, drop = TRUE, scales = "free", space = "free")
bubbleplot_elas_Month_clamsanct

# save
ggsave(plot = bubbleplot_elas_Month_clamsanct, filename = "figures-2024eDNA/bubbleplot_elas_Month_clamsanct.jpg", width = 7, height = 2.5, units = "in")
```


Oyster reef
```{r}
# First average replicates from same month
elas_df_month_oysterreef <- elas_df %>%
  filter(Habitat == "OYSTER REEF") %>%
    group_by(sites, Month_f, CommonName, `Species.CommonName`) %>%
    dplyr::summarize(Abundance = mean(Abundance))
elas_df_month_oysterreef


bubbleplot_elas_Month_oysterreef <- ggplot(elas_df_month_oysterreef, aes(x = factor(sites, levels = sitelevels), y = fct_rev(`Species.CommonName`))) + 
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = `Species.CommonName`), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_Sci_comname)+
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 10), 
        legend.position = "bottom") +
  guides(fill="none") +
  facet_grid(~Month_f, drop = TRUE, scales = "free", space = "free")
bubbleplot_elas_Month_oysterreef

# save
ggsave(plot = bubbleplot_elas_Month_oysterreef, filename = "figures-2024eDNA/bubbleplot_elas_Month_oysterreef.jpg", width = 9, height = 3, units = "in")
```



### Elas- Species Time Series - TBD

### Diversity

Species richness by month and habitat types
- In June, eel grass stations were not sampled but all other habitat types were. So any zeroes in June are real for all habitat types except eel grass---
```{r}
# First: define all months and habitats you'd expect
all_months <- unique(c(mifish_df$Month, elas_df$Month))
all_habitats <- unique(c(mifish_df$Habitat, elas_df$Habitat))

# MiFish
species_counts_2024_mifish <- mifish_df %>%
  group_by(Habitat, Month) %>%
  summarise(UniqueSpeciesCount = n_distinct(`Species.CommonName`), .groups = "drop") %>%
  complete(Habitat = all_habitats, Month = all_months, fill = list(UniqueSpeciesCount = 0)) %>%
  filter(!(Habitat == "EELGRASS" & Month == "6"))


# Elas
species_counts_2024_elas <- elas_df %>%
  group_by(Habitat, Month) %>%
  summarise(UniqueSpeciesCount = n_distinct(`Species.CommonName`), .groups = "drop") %>%
  complete(Habitat = all_habitats, Month = all_months, fill = list(UniqueSpeciesCount = 0)) %>%
  filter(!(Habitat == "EELGRASS" & Month == "6"))

species_counts_2024_mifish
species_counts_2024_elas

```

#### Mifish
```{r}

sp_richnesss_habitat_month_mifish_plot <- ggplot(species_counts_2024_mifish, aes(x = Month, y = UniqueSpeciesCount, group = Habitat, color = Habitat)) +
  geom_line(size = 1, alpha = 0.8) +
  geom_point(size = 2, alpha = 0.8) +
  labs(x = "Month", y = "Species Richness", title = "2024 Species Richness; MiFish") +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    legend.title = element_blank()
  ) +
  scale_color_manual(values = c("azure2", "chartreuse4", "cadetblue1", "darkgrey"))

sp_richnesss_habitat_month_mifish_plot

# save
ggsave(plot = sp_richnesss_habitat_month_mifish_plot, filename = "figures-2024eDNA/sp_richnesss_habitat_month_mifish_plot.jpg", width = 5, height = 3, units = "in")


```



#### Elas02
```{r}
sp_richnesss_habitat_month_elas_plot <- ggplot(species_counts_2024_elas, aes(x = Month, y = UniqueSpeciesCount, group = Habitat, color = Habitat)) +
  geom_line(size = 1, alpha = 0.8) +
  geom_point(size = 2, alpha = 0.8) +
  labs(x = "Month", y = "Species Richness", title = "2024 Species Richness; Elas02") +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    legend.title = element_blank()
  ) +
  scale_color_manual(values = c("azure2", "chartreuse4", "cadetblue1", "darkgrey"))

sp_richnesss_habitat_month_elas_plot

# save
ggsave(plot = sp_richnesss_habitat_month_elas_plot, filename = "figures-2024eDNA/sp_richnesss_habitat_month_elas_plot.jpg", width = 5, height = 3, units = "in")


```






