---
title: "Plots and Analysis for Salzman et al. Elasmobranch eDNA paper"
output: html_notebook
---


<br/>

# Table of Contents
- [Load packages](#load-packages)
- [Import and prepare data](#import-and-prepare-data)
- [QC and filtering](#qc-and-filtering)
- [Visualizations](#visualizations)

      
<br/>



# Load packages

```{r}
# rm(list = ls())
library(tidyverse)
library(phyloseq)
library(vegan)
library(RColorBrewer)
library(microbiome)
library(metagMisc)
library(fantaxtic)
library(ggrepel)
library(paletteer)

```

#  Import and prepare data

Import-
I am creating the equivalent of "phylo1" from REVAMP, the dataset is the ASV-based, raw reads, not qual filtered dataset; "qual filtered" means the unknown ASVs have been removed, & low effort samples removed. The only filtering that has been done is the unknown ASVs have been removed.

Side note- I checked some of the high abundance unknowns that were thrown out (ASV_4 from 2021 and ASV_2 and _3 from 2023) and they are diatoms. They are coming up as unknown bc they are not in my Elas02 database. 

**2021 data**
```{r}
asv_count <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2021-Elas02/ASV2Taxonomy/ASVs_counts_NOUNKNOWNS.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count) <- asv_count$x
asv_count <- asv_count %>% select(-x)
asv_count_mat <- as.matrix(asv_count)

asv_taxonomy <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2021-Elas02/ASV2Taxonomy/results-revamp-2021-Elas02_asvTaxonomyTable_NOUNKNOWNS.txt", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_taxonomy) <- asv_taxonomy$ASV
asv_taxonomy <- asv_taxonomy %>% select(-ASV)
asv_taxonomy_mat <- as.matrix(asv_taxonomy)

sample_metadata <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2021-Elas02/sample_metadata_forR.txt", header=TRUE, stringsAsFactors=TRUE)
row.names(sample_metadata) <- sample_metadata$Sample
sample_metadata <- sample_metadata %>% select(-Sample)

ASV_2021 = otu_table(asv_count_mat, taxa_are_rows = TRUE)
TAX_2021 = tax_table(asv_taxonomy_mat)
samples_2021 = sample_data(sample_metadata)

phylo_2021 <- phyloseq(ASV_2021, TAX_2021, samples_2021)

```


**2023 data**
```{r}
asv_count <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2023-Elas02/ASV2Taxonomy/ASVs_counts_NOUNKNOWNS.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count) <- asv_count$x
asv_count <- asv_count %>% select(-x)
asv_count_mat <- as.matrix(asv_count)

asv_taxonomy <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2023-Elas02/ASV2Taxonomy/results-revamp-2023-Elas02_asvTaxonomyTable_NOUNKNOWNS.txt", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_taxonomy) <- asv_taxonomy$ASV
asv_taxonomy <- asv_taxonomy %>% select(-ASV)
asv_taxonomy_mat <- as.matrix(asv_taxonomy)

sample_metadata <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2023-Elas02/sample_metadata_forR.txt", header=TRUE, stringsAsFactors=TRUE)
row.names(sample_metadata) <- sample_metadata$Sample
sample_metadata <- sample_metadata %>% select(-Sample)

ASV_2023 = otu_table(asv_count_mat, taxa_are_rows = TRUE)
TAX_2023 = tax_table(asv_taxonomy_mat)
samples_2023 = sample_data(sample_metadata)

phylo_2023 <- phyloseq(ASV_2023, TAX_2023, samples_2023)

```



Generate some metadata for manuscript
How many ASVs and reads were thrown out as unknowns?

**2021**
```{r}
asv_count_unknowns_2021 <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2021-Elas02/processed_tables/ASVs_counts_controlsRemoved.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count_unknowns_2021) <- asv_count_unknowns_2021$x
asv_count_unknowns_2021 <- asv_count_unknowns_2021 %>% select(-x)

asv_count_nounknowns_2021 <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2021-Elas02/processed_tables/ASVs_counts_NOUNKNOWNS_controlsRemoved.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count_nounknowns_2021) <- asv_count_nounknowns_2021$x
asv_count_nounknowns_2021 <- asv_count_nounknowns_2021 %>% select(-x)


# number of ASVs retained after removal of unknowns
count(asv_count_nounknowns_2021)/count(asv_count_unknowns_2021)


# number of reads retained after removal of unknowns
sum(rowSums(asv_count_nounknowns_2021))/sum(rowSums(asv_count_unknowns_2021))



```



**2023**
```{r}
asv_count_unknowns_2023 <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2023-Elas02/processed_tables/ASVs_counts_controlsRemoved.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count_unknowns_2023) <- asv_count_unknowns_2023$x
asv_count_unknowns_2023 <- asv_count_unknowns_2023 %>% select(-x)

asv_count_nounknowns_2023 <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2023-Elas02/processed_tables/ASVs_counts_NOUNKNOWNS_controlsRemoved.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count_nounknowns_2023) <- asv_count_nounknowns_2023$x
asv_count_nounknowns_2023 <- asv_count_nounknowns_2023 %>% select(-x)


# number of ASVs retained after removal of unknowns
count(asv_count_nounknowns_2023)/count(asv_count_unknowns_2023)


# number of reads retained after removal of unknowns
sum(rowSums(asv_count_nounknowns_2023))/sum(rowSums(asv_count_unknowns_2023))



```



# QC and filtering


## Rarefaction curves

```{r}
# output from phyloseq needs to be converted to matrix and transposed
table1 <- otu_table(phylo_2021)
class(table1) <- "matrix" 
table1 <- t(table1) 

rarecurve(table1, step=50, cex=0.5)

```
A few don't look great (T9S7, T5S6, T3Wessuck, T1Tiana, T7S7) but most are good!


```{r}
# output from phyloseq needs to be converted to matrix and transposed
table1 <- otu_table(phylo_2023)
class(table1) <- "matrix" 
table1 <- t(table1) 

rarecurve(table1, step=50, cex=0.5)

```
Keep an eye on SC5, SS9, SS3, T3S2, T5S1, T3S7, T3WC, and a few others

## Curating

For quick reference, scan through Krona plots:
- 2021 dataset [here](results-revamp-2021-Elas02/ASV2Taxonomy/KRONA_plots/results-revamp-2021-Elas02_samplesSummedKRONA.html) and 
- 2023 dataset [here](results-revamp-2023-Elas02/ASV2Taxonomy/KRONA_plots/results-revamp-2023-Elas02_samplesSummedKRONA.html))
- Note, unknowns in Krona plots were already removed from phylo objects

**2021**
Check some features of the phyloseq object
```{r}
ranks <- rank_names(phylo_2021)
ranks

unique(tax_table(phylo_2021)[, ranks[1]])
unique(tax_table(phylo_2021)[, ranks[2]])
unique(tax_table(phylo_2021)[, ranks[3]])
```

Remove the birds, mammals, and Actinopteri- even if some of these have exact match to primers, they are not really subject of study.
```{r}
nrow(tax_table(phylo_2021)) # number of taxa before filtering
ps2021 <- subset_taxa(phylo_2021, !Class == 'Aves' & !Class == 'Mammalia' & !Class == 'Actinopteri')
nrow(tax_table(ps2021)) # number of taxa left
unique(tax_table(ps2021)[, ranks[3]])
```
**2023**
Check some features of the phyloseq object
```{r}
ranks <- rank_names(phylo_2023)
ranks

unique(tax_table(phylo_2023)[, ranks[1]])
unique(tax_table(phylo_2023)[, ranks[2]])
unique(tax_table(phylo_2023)[, ranks[3]])
```

Remove the birds, mammals, and Actinopteri- even if some of these have exact match to primers, they are not really subject of study.
```{r}
nrow(tax_table(phylo_2023)) # number of taxa before filtering
ps2023 <- subset_taxa(phylo_2023, !Class == 'Aves' & !Class == 'Mammalia' & !Class == 'Actinopteri')
nrow(tax_table(ps2023)) # number of taxa left
unique(tax_table(ps2023)[, ranks[3]])
```



Remove low abundant reads
- Scan REVAMP output:`results-revamp-2021-Elas02_asvTaxonomyTable.txt` and `results-revamp-2023-Elas02_asvTaxonomyTable.txt` have taxonomy of all ASVs. 
  - In 2021, there are a few that don't make sense. There were a lot of pos controls that year and it seems that some crossover occurred and pos control reads were detected in other samples:
    - leopard shark (Triakis semifasciata: ASV_123, ASV_1406, ASV_22, ASV_293, ASV_308, ASV_556, ASV_659, ASV_801, ASV_940]
    - chain catshark [Scyliorhinus retifer: ASV_1030, ASV_1254, ASV_1495, ASV_1496, ASV_220, ASV_280, ASV_397, ASV_584, ASV_631, ASV_7, ASV_715, ASV_797]
    - Arabian smooth-hound [Mustelus mosis: ASV_78]
    - brown smooth-hound [Mustelus henlei: ASV_616, ASV_619]
    - Japanese wobbegong [Orectolobus japonicus, ASV_234]
    - and nurse shark [Ginglymostoma cirratum: ASV129, ASV_588]) 

From Libby's thesis:
*Subcutaneous muscle tissue samples were collected for two local shark species: spiny dogfish (Squalus acanthius) and smooth dogfish (Musculus canis). A tissue sample was taken from each species. Each specimen was caught in Long Island waters and stored at -20°C as part of an alternate study. Water samples were taken from various tanks containing sharks at the Long Island Aquarium in October 2021. These tanks contained known species of elasmobranchs, including sand tiger sharks (Charcharias taurus), an ornate wobbegong (Orectolobus japonicus), and nurse sharks (Ginglymostoma cirratum). Additional water samples were taken from a shark tank at Cornell Cooperative Extension in Babylon, NY in October of 2021. This tank contained only chain catshark (Scyliorhinus retifer) but had a shared sump system with other elasmobranchs and used bay water. All positive controls were processed following the same protocol as the experimental samples.*




Check some of these weird annotations:
```{r}
unique(tax_table(ps2021)[, ranks[7]])
unique(tax_table(ps2023)[, ranks[7]])


weird_taxa <- c("Triakis semifasciata", "Scyliorhinus retifer", "Mustelus mosis", "Mustelus henlei", "Orectolobus japonicus", "Ginglymostoma cirratum", "Mustelus griseus")
```

Check their read abundance from 2021:
```{r}
weird_taxa_ps <- subset_taxa(phylo_2021, Species %in% weird_taxa)
data.frame(otu_table(weird_taxa_ps)[,11:92]) # skip first 10 columns which are pos controls. These are expected to have some of these strange taxa. 
```

- Most are zeroes. Where detected, most of these are less than 1000 reads except for ASV_78 (Mustelus mosis), in T5S11, which has over 8000 reads. I blasted this ASV and its second blast hit is Mustelus canis so I think this is a misannotation.
- All others seem like bleed over from positive controls


Check their read abundance from 2023:
```{r}
weird_taxa_ps <- subset_taxa(phylo_2023, Species %in% weird_taxa)
data.frame(otu_table(weird_taxa_ps)) 
```
Again mostly zeroes. Highest read abundance is ~650 reads.

Remove ASVs with less than 1000 reads in a sample. 
```{r}
ps2021 <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps2021,
  minabund = 1000,
  relabund = FALSE)

data.frame(otu_table(ps2021))
```

```{r}
ps2023 <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps2023,
  minabund = 1000,
  relabund = FALSE)

data.frame(otu_table(ps2023))
```


### Manual curation

There are some abundant taxa resolved only to the genus level. For example:
- In 2021, Squalus makes up 8% of the Chondricthyes
- In 2023, Bathytosia makes up 19% of the Chondricthyes

I don't want to lose these  when I group by species. Can use NArm = FALSE option in `tax_glom` but then they will plot as NA and I will lose all tax information in the species plot... 

  - Found a cool package for replacing NAs in the tax table with next lowest known rank, [fantaxtic](https://github.com/gmteunisse/fantaxtic)

  - There are only [3 species of Bathytosia](https://www.fishbase.us/Nomenclature/ScientificNameSearchList.php?crit1_value=Bathytoshia) and only one is from the North Atlantic, *Bathytoshia centroura*, the roughtail stingray. So these are likely those. Replace the Species rank name, NA, with "Bathytoshia sp."

  - There are [many more Squalus species](https://www.fishbase.us/Nomenclature/ScientificNameSearchList.php?) so harder to say but these are likely Squalus acanthias, spiny dogfish, which are common regionally. Still the blast hits are not really conclusive so can replace with *Squalus sp.*

  - There are some Rhinoptera only identified to the genus level. Their blast hits are kind of all over the genus so replace with *Rhinoptera sp.*

  - The genus Pteroplatytrygon was also identified at low abundance. There is only one species in this genus, *Pteroplatytrygon violacea*, so it is very likely this species... I don't know why REVAMP didn't pull out the spp name in this case. Replace with *Pteroplatytrygon violacea*

- **Remember to discuss all this in MS.**
  
```{r}
# Check what they look like before replacement
data.frame(tax_table(subset_taxa(ps2021, is.na(Species))))
data.frame(tax_table(subset_taxa(ps2023, is.na(Species))))
```

```{r}
# Replace NA in Species with "Name-of-genus sp."
tax_table(ps2021) <- tax_table(fantaxtic::name_na_taxa(ps2021, include_rank = F, na_label = "<tax> sp."))
tax_table(ps2023) <- tax_table(fantaxtic::name_na_taxa(ps2023, include_rank = F, na_label = "<tax> sp."))

```

And lastly replace the Mustelus mosis ASV (ASV_78 from 2021) with Mustelus sp., based on discussion above and replace Pteroplatytrygon sp. with Pteroplatytrygon violacea.

```{r}
tax_table(ps2021) <- gsub(tax_table(ps2021), pattern="Mustelus mosis", replacement="Mustelus sp.")
tax_table(ps2023) <- gsub(tax_table(ps2023), pattern="Pteroplatytrygon sp.", replacement="Pteroplatytrygon violacea")

```

```{r}
# Check after replacement
data.frame(tax_table(ps2021))
data.frame(tax_table(ps2023))
```
Fix some labeling issues with 2021 positive controls:
Remove SA2 and MC2- something went wrong with these
```{r}
ps2021 = subset_samples(ps2021, replicates != "MC2" & replicates != "SA2")
ps2021
```
Also SA1 and MC1 labels were swapped. Fixed this directly in metadata table under "replicates" column


## Agglomerate by taxonomy
Generate new otu tables after grouping ASVs by annotated species name
```{r}
ps2021glommed <- tax_glom(ps2021, "Species")

# plot total reads
plot_bar(ps2021glommed, x = "Sample")

# view tax_table
data.frame(tax_table(ps2021glommed))
```



```{r}
ps2023glommed <- tax_glom(ps2023, "Species")

# plot total reads
plot_bar(ps2023glommed, x = "Sample")

# view tax_table
data.frame(tax_table(ps2023glommed))
```

## Export taxonomy-based tables for collaborator

**2021**
```{r}
elas02_2021_asv_table_taxbased <- data.frame(otu_table(ps2021glommed))
elas02_2021_tax_table_taxbased <- rownames_to_column(data.frame(tax_table(ps2021glommed)), var = "1stRepresentative_ASV")
elas02_2021_sample_data_taxbased <- data.frame(sample_data(ps2021glommed))


# add on common names to tax_table
commonnames <- read_csv(file = "../eDNA-databases/commonnames.csv")
elas02_2021_tax_table_taxbased <- left_join(elas02_2021_tax_table_taxbased, commonnames, by = "Species")

# view
elas02_2021_asv_table_taxbased
elas02_2021_tax_table_taxbased
elas02_2021_sample_data_taxbased

# export
write.csv(elas02_2021_asv_table_taxbased,"figures-salzman-etal-elasmos/elas02_2021_asv_table_taxbased.csv", row.names = TRUE)
write.csv(elas02_2021_tax_table_taxbased,"figures-salzman-etal-elasmos/elas02_2021_tax_table_taxbased.csv", row.names = FALSE)
write.csv(elas02_2021_sample_data_taxbased,"figures-salzman-etal-elasmos/elas02_2021_sample_data_taxbased.csv", row.names = TRUE)
```


**2023**
```{r}
elas02_2023_asv_table_taxbased <- data.frame(otu_table(ps2023glommed))
elas02_2023_tax_table_taxbased <- rownames_to_column(data.frame(tax_table(ps2023glommed)), var = "1stRepresentative_ASV")
elas02_2023_sample_data_taxbased <- data.frame(sample_data(ps2023glommed))

# add on common names to tax_table
commonnames <- read_csv(file = "../eDNA-databases/commonnames.csv")
elas02_2023_tax_table_taxbased <- left_join(elas02_2023_tax_table_taxbased, commonnames, by = "Species")

# view
elas02_2023_asv_table_taxbased
elas02_2023_tax_table_taxbased
elas02_2023_sample_data_taxbased

# export
write.csv(elas02_2023_asv_table_taxbased,"figures-salzman-etal-elasmos/elas02_2023_asv_table_taxbased.csv", row.names = TRUE)
write.csv(elas02_2023_tax_table_taxbased,"figures-salzman-etal-elasmos/elas02_2023_tax_table_taxbased.csv", row.names = FALSE)
write.csv(elas02_2023_sample_data_taxbased,"figures-salzman-etal-elasmos/elas02_2023_sample_data_taxbased.csv", row.names = TRUE)
```



Also calculate *relative abundances* (# of ASV sequences/sum total sequences in sample), for some plotting

```{r}
ps2021_ra <- microbiome::transform(ps2021, transform = "compositional")
ps2021glommed_ra <- microbiome::transform(ps2021glommed, transform = "compositional")

ps2023_ra <- microbiome::transform(ps2023, transform = "compositional")
ps2023glommed_ra <- microbiome::transform(ps2023glommed, transform = "compositional")

```

# Visualizations



## Simple barplots

2021, raw reads
```{r}
ps2021_barplot <- plot_bar(ps2021, x = "Sample", fill = "Species")
ps2021_barplot
```

2021, relatve abundances
```{r}
ps2021_ra_barplot <- plot_bar(ps2021_ra, x = "Sample", fill = "Species")
ps2021_ra_barplot
```


2023, raw reads
```{r}
ps2023_barplot <- plot_bar(ps2023, x = "Sample", fill = "Species")
ps2023_barplot
```

2023, relatve abundances
```{r}
ps2023_ra_barplot <- plot_bar(ps2023_ra, x = "Sample", fill = "Species")
ps2023_ra_barplot
```

## Examine controls 

Plot positive controls

```{r}


ps2021_posctl_barplot <- plot_bar(subset_samples(ps2021glommed, controls=="positive"), x = "replicates", fill = "Species", title = "2021 Positive Controls: Read Abundance") + 
  scale_y_continuous(name="Read Abundance")+
  theme(axis.title.x=element_blank())+
  scale_fill_paletteer_d("MetBrewer::Signac")
ps2021_posctl_barplot

ps2021_posctl_ra_barplot <- plot_bar(subset_samples(ps2021glommed_ra, controls=="positive"), x = "replicates", fill = "Species", title = "2021 Positive Controls: Relative Abundance") + 
  scale_y_continuous(name="Relative Abundance")+
  theme(axis.title.x=element_blank())+
    scale_fill_paletteer_d("MetBrewer::Signac")
ps2021_posctl_ra_barplot





# save
ggsave(plot = ps2021_posctl_barplot, filename = "figures-salzman-etal-elasmos/2021_posctl_barplot.eps", width = 7, height = 5, units = "in")
ggsave(plot = ps2021_posctl_ra_barplot, filename = "figures-salzman-etal-elasmos/2021_posctl_ra_barplot.eps", width = 7, height = 5, units = "in")

# export data of pos ctls for text of manuscript
ps2021_posctls <- data.frame(otu_table(subset_samples(ps2021glommed, controls == "positive")))
ps2021_posctls_ra <- data.frame(otu_table(subset_samples(ps2021glommed_ra, controls == "positive")))
ps2021_posctls_taxa <- data.frame(tax_table(subset_samples(ps2021glommed, controls == "positive")))

ps2021_posctls
ps2021_posctls_ra
ps2021_posctls_taxa

write.csv(ps2021_posctls, file = "figures-salzman-etal-elasmos/ps2021_posctls.csv")
write.csv(ps2021_posctls_ra, file = "figures-salzman-etal-elasmos/ps2021_posctls_ra.csv")
write.csv(ps2021_posctls_taxa, file = "figures-salzman-etal-elasmos/ps2021_posctls_taxa.csv")


```



Remove Pos and Neg  Controls from all 8 phyloseq objects:

ASV-based, reads:
ps2021
ps2023

ASV-based, relative abundance:
ps2021_ra
ps2023_ra

Taxonomy (species)- based, reads:
ps2021glommed
ps2023glommed

Taxonomy (species)- based, relative abundance:
ps2021glommed_ra
ps2023glommed_ra

```{r}
ps2021 <- subset_samples(ps2021, is.na(controls))
ps2021_ra <- subset_samples(ps2021_ra, is.na(controls))
ps2021glommed <- subset_samples(ps2021glommed, is.na(controls))
ps2021glommed_ra <- subset_samples(ps2021glommed_ra, is.na(controls))


ps2023 <- subset_samples(ps2023, is.na(controls))
ps2023_ra <- subset_samples(ps2023_ra, is.na(controls))
ps2023glommed <- subset_samples(ps2023glommed, is.na(controls))
ps2023glommed_ra <- subset_samples(ps2023glommed_ra, is.na(controls))

```



## Bubble plots

Based on my previous [scripts](https://github.com/lizsuter/Cariaco_Euk)

Work with tax-based dataframes (rather than ASV-based):

Taxonomy (species)- based, reads:
ps2021glommed
ps2023glommed

Taxonomy (species)- based, relative abundance:
ps2021glommed_ra
ps2023glommed_ra

```{r}
# convert ps object to dataframe using phyloseq's psmelt
ps2021glommed_df <- psmelt(ps2021glommed)
ps2023glommed_df <- psmelt(ps2023glommed)
ps2021glommed_ra_df <- psmelt(ps2021glommed_ra)
ps2023glommed_ra_df <- psmelt(ps2023glommed_ra)

# replace zeroes in the table with NA
ps2021glommed_df[ps2021glommed_df == 0] <- NA
ps2023glommed_df[ps2023glommed_df == 0] <- NA
ps2021glommed_ra_df[ps2021glommed_ra_df == 0] <- NA
ps2023glommed_ra_df[ps2023glommed_ra_df == 0] <- NA

# and remove rows with NAs (so they don't appear as small dots in plot)
ps2021glommed_df <-  filter(ps2021glommed_df, !is.na(Abundance))
ps2023glommed_df <-  filter(ps2023glommed_df, !is.na(Abundance))
ps2021glommed_ra_df <-  filter(ps2021glommed_ra_df, !is.na(Abundance))
ps2023glommed_ra_df <-  filter(ps2023glommed_ra_df, !is.na(Abundance))

# add in common names
ps2021glommed_df <- left_join(ps2021glommed_df, commonnames, by = "Species")
ps2023glommed_df <- left_join(ps2023glommed_df, commonnames, by = "Species")
ps2021glommed_ra_df <- left_join(ps2021glommed_ra_df, commonnames, by = "Species")
ps2023glommed_ra_df <- left_join(ps2023glommed_ra_df, commonnames, by = "Species")

# and view
ps2021glommed_df
ps2023glommed_df
ps2021glommed_ra_df
ps2023glommed_ra_df
```



Plot by species, scientific name
**2021**
```{r}
# ps2021glommed_df
# ps2023glommed_df

# pull out and order site names
levels(unique(c(ps2021glommed_df$sites, ps2023glommed_df$sites)))
# reorder to
sitelevels <- c("1","2","3","4","5","6","7","8","9","10","11","12","13","TI","WS", "C4", "C5", "S1", "S3","S6", "S9")

# set colors so Com name and Sp name are consistent across plots
# have 16 unique species
myColors_sciname <- setNames(c("steelblue4","#DF536B","#61D04F","#2297E6","#28E2E5","#CD0BBC", "#F5C710","brown","#6A3D9A","#FF7F00","gray","orchid1","#FDBF6F","gold1","#FB9A99","khaki2"), commonnames$Species[which(commonnames$Species %in% unique(c(ps2021glommed_df$Species, ps2023glommed_df$Species)))])
myColors_comname <- setNames(c("steelblue4","#DF536B","#61D04F","#2297E6","#28E2E5","#CD0BBC", "#F5C710","brown","#6A3D9A","#FF7F00","gray","orchid1","#FDBF6F","gold1","#FB9A99","khaki2"),commonnames$CommonName[which(commonnames$Species %in% unique(c(ps2021glommed_df$Species, ps2023glommed_df$Species)))])

# and check they are listed in same order
myColors_sciname
myColors_comname


# 2021, read abundance, tax-based, common name
bubbleplot_comname_2021 <- ggplot(ps2021glommed_df, aes(x = factor(sites, levels = sitelevels), y = fct_rev(CommonName))) + # the fancy stuff around y (CommonName) helps to present it in reverse order in the plot (from top to btm alphabetically)
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = CommonName), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_comname) +
  theme(axis.title.x=element_blank()) +
  facet_grid(Month~BAYSIDE, scales = "free", space = "free", drop= TRUE)

# 2021, read abundance, tax-based, sci name
bubbleplot_sciname_2021 <- ggplot(ps2021glommed_df, aes(x = factor(sites, levels = sitelevels), y = fct_rev(Species))) +
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = Species), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_sciname) +
  theme(axis.title.x=element_blank()) +
  facet_grid(Month~BAYSIDE, scales = "free", space = "free", drop= TRUE)


# 2021, relative abundance, tax-based, common name
bubbleplot_comname_2021_ra <- ggplot(ps2021glommed_ra_df, aes(x = factor(sites, levels = sitelevels), y = fct_rev(CommonName))) +
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = CommonName), color = "black", pch = 21) +
  scale_size_area(breaks = c(.25,.5,.75,1), max_size = 6)+
  xlab("")+
  ylab("")+
  labs(size="Relative Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_comname) +
  theme(axis.title.x=element_blank()) +
  facet_grid(Month~BAYSIDE, scales = "free", space = "free", drop= TRUE)


# 2021, relative abundance, tax-based, sci name
bubbleplot_sciname_2021_ra <- ggplot(ps2021glommed_ra_df, aes(x = factor(sites, levels = sitelevels), y = fct_rev(Species))) +
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = Species), color = "black", pch = 21) +
  scale_size_area(breaks = c(.25,.5,.75,1), max_size = 6)+
  xlab("")+
  ylab("")+
  labs(size="Relative Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_sciname) +
  theme(axis.title.x=element_blank()) +
  facet_grid(Month~BAYSIDE, scales = "free", space = "free", drop= TRUE)


bubbleplot_comname_2021
bubbleplot_sciname_2021

bubbleplot_comname_2021_ra
bubbleplot_sciname_2021_ra

# save
ggsave(plot = bubbleplot_comname_2021, filename = "figures-salzman-etal-elasmos/bubbleplot_comname_2021.eps", width = 7, height = 7, units = "in")
ggsave(plot = bubbleplot_sciname_2021, filename = "figures-salzman-etal-elasmos/bubbleplot_sciname_2021.eps", width = 6, height = 7, units = "in")
ggsave(plot = bubbleplot_comname_2021_ra, filename = "figures-salzman-etal-elasmos/bubbleplot_comname_2021_ra.eps", width = 7, height = 7, units = "in")
ggsave(plot = bubbleplot_sciname_2021_ra, filename = "figures-salzman-etal-elasmos/bubbleplot_sciname_2021_ra.eps", width = 6, height = 7, units = "in")
```


**2023**
```{r}
# ps2021glommed_ra_df
# ps2023glommed_ra_df

# 2023, read abundance, tax-based, common name
bubbleplot_comname_2023 <- ggplot(ps2023glommed_df, aes(x = factor(sites, levels = sitelevels), y = fct_rev(CommonName))) + # the fancy stuff around y (CommonName) helps to present it in reverse order in the plot (from top to btm alphabetically)
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = CommonName), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_comname) +
  theme(axis.title.x=element_blank()) +
  facet_grid(Month~BAYSIDE, scales = "free", space = "free", drop= TRUE)

# 2023, read abundance, tax-based, sci name
bubbleplot_sciname_2023 <- ggplot(ps2023glommed_df, aes(x = factor(sites, levels = sitelevels), y = fct_rev(Species))) +
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = Species), color = "black", pch = 21) +
  scale_size_area(breaks = c(1000,10000,100000,1000000,10000000))+
  xlab("")+
  ylab("")+
  labs(size="Read Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_sciname) +
  theme(axis.title.x=element_blank()) +
  facet_grid(Month~BAYSIDE, scales = "free", space = "free", drop= TRUE)


# 2023, relative abundance, tax-based, common name
bubbleplot_comname_2023_ra <- ggplot(ps2023glommed_ra_df, aes(x = factor(sites, levels = sitelevels), y = fct_rev(CommonName))) +
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = CommonName), color = "black", pch = 21) +
  scale_size_area(breaks = c(.25,.5,.75,1), max_size = 6)+
  xlab("")+
  ylab("")+
  labs(size="Relative Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_comname) +
  theme(axis.title.x=element_blank()) +
  facet_grid(Month~BAYSIDE, scales = "free", space = "free", drop= TRUE)


# 2023, relative abundance, tax-based, sci name
bubbleplot_sciname_2023_ra <- ggplot(ps2023glommed_ra_df, aes(x = factor(sites, levels = sitelevels), y = fct_rev(Species))) +
  geom_point(aes(size=ifelse(Abundance==0, NA, Abundance), fill = Species), color = "black", pch = 21) +
  scale_size_area(breaks = c(.25,.5,.75,1), max_size = 6)+
  xlab("")+
  ylab("")+
  labs(size="Relative Abundance", fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_sciname) +
  theme(axis.title.x=element_blank()) +
  facet_grid(Month~BAYSIDE, scales = "free", space = "free", drop= TRUE)


bubbleplot_comname_2023
bubbleplot_sciname_2023

bubbleplot_comname_2023_ra
bubbleplot_sciname_2023_ra

# save
ggsave(plot = bubbleplot_comname_2023, filename = "figures-salzman-etal-elasmos/bubbleplot_comname_2023.eps", width = 7, height = 7, units = "in")
ggsave(plot = bubbleplot_sciname_2023, filename = "figures-salzman-etal-elasmos/bubbleplot_sciname_2023.eps", width = 7.5, height = 7, units = "in")
ggsave(plot = bubbleplot_comname_2023_ra, filename = "figures-salzman-etal-elasmos/bubbleplot_comname_2023_ra.eps", width = 7, height = 7, units = "in")
ggsave(plot = bubbleplot_sciname_2023_ra, filename = "figures-salzman-etal-elasmos/bubbleplot_sciname_2023_ra.eps", width = 7.5, height = 7, units = "in")
```



## Pie charts

Summarize fractional abundances across whole dataset-
```{r}
ps2021_whole <- ps2021glommed_df %>%
  group_by(Species) %>%
    summarize(total_reads = sum(Abundance)) %>%
  mutate(rel_abun = total_reads/sum(across(total_reads))) %>%
  mutate(rel_abun_percent = rel_abun*100) %>%
  mutate(csum = rev(cumsum(rev(rel_abun))), 
         pos = rel_abun/2 + lead(csum, 1),
         pos = if_else(is.na(pos), rel_abun/2, pos)) %>%
  left_join(commonnames, by = "Species")

ps2023_whole <- ps2023glommed_df %>%
  group_by(Species) %>%
    summarize(total_reads = sum(Abundance)) %>%
  mutate(rel_abun = total_reads/sum(across(total_reads))) %>%
  mutate(rel_abun_percent = rel_abun*100) %>%
  mutate(csum = rev(cumsum(rev(rel_abun))), 
         pos = rel_abun/2 + lead(csum, 1),
         pos = if_else(is.na(pos), rel_abun/2, pos)) %>%

  left_join(commonnames, by = "Species")

ps2021_whole
ps2023_whole
```

Apparently there is no geom for pie charts but you can [build a stacked geom_bar and make polar coordinates](https://r-graph-gallery.com/piechart-ggplot2.html)
```{r}
ggplot(ps2021_whole, aes(x="", y = rel_abun, fill=CommonName)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  xlab("")+
  ylab("")+
  labs(fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_comname) +
  theme(axis.title.x=element_blank()) +
  geom_text(aes(label = round(rel_abun_percent,1)),
            position = position_stack(vjust = 0.5))

ggplot(ps2023_whole, aes(x="", y = rel_abun, fill=CommonName)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  xlab("")+
  ylab("")+
  labs(fill = "")+
  theme_bw() +
  scale_fill_manual(values = myColors_comname) +
  theme(axis.title.x=element_blank()) +
  geom_text(aes(label = round(rel_abun_percent,1)),
            position = position_stack(vjust = 0.5))

```


## Time Series Plots
Look at total read abundance and rel abundance for different spp per month. Aggregate sites by bayside (E/W) and type (open water/ bivalve habitat)

**2021**
```{r}
# First group total reads by Species, Habitat_type and Month so there is one point per time point on the plot
TimeSeries_df_2021 <- ps2021glommed_df %>%
  group_by(Month, Habitat_type, CommonName) %>%
  summarise(TotalReads = sum(Abundance))
TimeSeries_df_2021
```


```{r}
#First force the order of Habitat type so it plots in desired order
TimeSeries_df_2021$Habitat_type_f = factor(TimeSeries_df_2021$Habitat_type, levels=c("OPEN WATER", "CLAM"))

speciestimeseries_comname_2021 <- ggplot(TimeSeries_df_2021, aes(x = Month, y = log10(TotalReads))) +
  geom_col(aes(fill = CommonName), show.legend = FALSE) +
  scale_fill_manual(values = myColors_comname) +
  ylab("Log 10 (Reads)")+
  theme_bw() +
  theme(axis.title.x=element_blank(), legend.title=element_blank()) +
  scale_x_continuous(breaks=c(5,6,7,8,9), labels = c("M","J","J","A","S")) +
  facet_grid(Habitat_type_f~CommonName)

speciestimeseries_comname_2021
ggsave(speciestimeseries_comname_2021, filename = "figures-salzman-etal-elasmos/speciestimeseries_comname_2021.eps", height = 4, width = 10, units = "in")
```

**2023**
```{r}
# First group total reads by Species, Habitat_type and Month so there is one point per time point on the plot
TimeSeries_df_2023 <- ps2023glommed_df %>%
  group_by(Month, Habitat_type, CommonName) %>%
  summarise(TotalReads = sum(Abundance))
TimeSeries_df_2023
```

```{r}
#First force the order of Habitat type so it plots in desired order
TimeSeries_df_2023$Habitat_type_f = factor(TimeSeries_df_2023$Habitat_type, levels=c("OPEN WATER", "CLAM SANCTUARY", "OYSTER REEF"))

speciestimeseries_comname_2023 <- ggplot(TimeSeries_df_2023, aes(x = Month, y = log10(TotalReads))) +
  geom_col(aes(fill = CommonName), show.legend = FALSE) +
  scale_fill_manual(values = myColors_comname) +
  ylab("Log 10 (Reads)")+
  theme_bw() +
  theme(axis.title.x=element_blank(), legend.title=element_blank()) +
  scale_x_continuous(breaks=c(5,6,7,8,9), labels = c("M","J","J","A","S")) +
  facet_grid(Habitat_type_f~CommonName)

speciestimeseries_comname_2023
ggsave(speciestimeseries_comname_2023, filename = "figures-salzman-etal-elasmos/speciestimeseries_comname_2023.eps", height = 4, width = 12, units = "in")
```
