---
title: "Plots and Analysis and all eDNA Hand Samples: 2020-2024"
output: html_notebook
---



<br/>
<br/>

 

# Load packages

```{r}
# rm(list = ls())
library(tidyverse)
library(phyloseq)
library(decontam)
library(vegan)
library(RColorBrewer)
library(microbiome)
library(metagMisc)
library(fantaxtic)
library(ggrepel)
library(paletteer)
library(pals)
```

#  Import and prepare data
## MiFish
### 2020 MiFish data
```{r}
asv_count <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2020-MiFIsh/ASV2Taxonomy/ASVs_counts_NOUNKNOWNS.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count) <- asv_count$x
asv_count <- asv_count %>% select(-x)
asv_count_mat <- as.matrix(asv_count)

asv_taxonomy <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2020-MiFIsh/ASV2Taxonomy/results-revamp-2020-MiFIsh_asvTaxonomyTable_NOUNKNOWNS.txt", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_taxonomy) <- asv_taxonomy$ASV
asv_taxonomy <- asv_taxonomy %>% select(-ASV)
asv_taxonomy_mat <- as.matrix(asv_taxonomy)

sample_metadata <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2020-MiFIsh/sample_metadata_forR.txt", header=TRUE, stringsAsFactors=TRUE)
row.names(sample_metadata) <- sample_metadata$Sample
sample_metadata <- sample_metadata %>% select(-Sample)

ASV_mifish_2020 = otu_table(asv_count_mat, taxa_are_rows = TRUE)
TAX_mifish_2020 = tax_table(asv_taxonomy_mat)
samples_mifish_2020 = sample_data(sample_metadata)

phylo_mifish_2020 <- phyloseq(ASV_mifish_2020, TAX_mifish_2020, samples_mifish_2020)

```


### 2021 MiFish data
```{r}
asv_count <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2021-MiFIsh/ASV2Taxonomy/ASVs_counts_NOUNKNOWNS.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count) <- asv_count$x
asv_count <- asv_count %>% select(-x)
asv_count_mat <- as.matrix(asv_count)

asv_taxonomy <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2021-MiFIsh/ASV2Taxonomy/results-revamp-2021-MiFIsh_asvTaxonomyTable_NOUNKNOWNS.txt", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_taxonomy) <- asv_taxonomy$ASV
asv_taxonomy <- asv_taxonomy %>% select(-ASV)
asv_taxonomy_mat <- as.matrix(asv_taxonomy)

sample_metadata <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2021-MiFIsh/sample_metadata_forR.txt", header=TRUE, stringsAsFactors=TRUE)
row.names(sample_metadata) <- sample_metadata$Sample
sample_metadata <- sample_metadata %>% select(-Sample)

ASV_mifish_2021 = otu_table(asv_count_mat, taxa_are_rows = TRUE)
TAX_mifish_2021 = tax_table(asv_taxonomy_mat)
samples_mifish_2021 = sample_data(sample_metadata)

phylo_mifish_2021 <- phyloseq(ASV_mifish_2021, TAX_mifish_2021, samples_mifish_2021)

```

### 2022 MiFish data
Analyzed in two batches. Import a and b, then join into one table-
```{r}
asv_counta <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2022-MiFisha/ASV2Taxonomy/ASVs_counts_NOUNKNOWNS.tsv", header=TRUE, stringsAsFactors=FALSE)
asv_countb <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2022-MiFishb/ASV2Taxonomy/ASVs_counts_NOUNKNOWNS.tsv", header=TRUE, stringsAsFactors=FALSE)

# ASV numbers are unique to each of the subsets of data, a and b. Marker into ASV name to indicate whether it's from library a or b
asv_counta <- asv_counta %>% mutate(x = paste0(x, "_a"))
asv_countb <- asv_countb %>% mutate(x = paste0(x, "_b"))

asv_count <- full_join(asv_counta, asv_countb)

row.names(asv_count) <- asv_count$x
asv_count <- asv_count %>% select(-x)
asv_count_mat <- as.matrix(asv_count)

asv_taxonomya <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2022-MiFisha/ASV2Taxonomy/results-revamp-2022-MiFisha_asvTaxonomyTable_NOUNKNOWNS.txt", header=TRUE, stringsAsFactors=FALSE)
asv_taxonomyb <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2022-MiFishb/ASV2Taxonomy/results-revamp-2022-MiFishb_asvTaxonomyTable_NOUNKNOWNS.txt", header=TRUE, stringsAsFactors=FALSE)

# ASV numbers are unique to each of the subsets of data, a and b. Marker into ASV name to indicate whether it's from library a or b
asv_taxonomya <- asv_taxonomya %>% mutate(ASV = paste0(ASV, "_a"))
asv_taxonomyb <- asv_taxonomyb %>% mutate(ASV = paste0(ASV, "_b"))

asv_taxonomy <- full_join(asv_taxonomya, asv_taxonomyb)

row.names(asv_taxonomy) <- asv_taxonomy$ASV
asv_taxonomy <- asv_taxonomy %>% select(-ASV)
asv_taxonomy_mat <- as.matrix(asv_taxonomy)

sample_metadataa <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2022-MiFisha/sample_metadata_forR.txt", header=TRUE, stringsAsFactors=TRUE)
sample_metadatab <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2022-MiFishb/sample_metadata_forR.txt", header=TRUE, stringsAsFactors=TRUE)

sample_metadata <- full_join(sample_metadataa, sample_metadatab)

row.names(sample_metadata) <- sample_metadata$Sample
sample_metadata <- sample_metadata %>% select(-Sample)

ASV_mifish_2022 = otu_table(asv_count_mat, taxa_are_rows = TRUE)
TAX_mifish_2022 = tax_table(asv_taxonomy_mat)
samples_mifish_2022 = sample_data(sample_metadata)

# set NAs to zeroes
ASV_mifish_2022[is.na(ASV_mifish_2022)] <- 0



phylo_mifish_2022 <- phyloseq(ASV_mifish_2022, TAX_mifish_2022, samples_mifish_2022)

```


### 2023 MiFish data
```{r}
asv_count <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2023-MiFIsh/ASV2Taxonomy/ASVs_counts_NOUNKNOWNS.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count) <- asv_count$x
asv_count <- asv_count %>% select(-x)
asv_count_mat <- as.matrix(asv_count)

asv_taxonomy <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2023-MiFIsh/ASV2Taxonomy/results-revamp-2023-MiFIsh_asvTaxonomyTable_NOUNKNOWNS.txt", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_taxonomy) <- asv_taxonomy$ASV
asv_taxonomy <- asv_taxonomy %>% select(-ASV)
asv_taxonomy_mat <- as.matrix(asv_taxonomy)

sample_metadata <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2023-MiFIsh/sample_metadata_forR.txt", header=TRUE, stringsAsFactors=TRUE)
row.names(sample_metadata) <- sample_metadata$Sample
sample_metadata <- sample_metadata %>% select(-Sample)

ASV_mifish_2023 = otu_table(asv_count_mat, taxa_are_rows = TRUE)
TAX_mifish_2023 = tax_table(asv_taxonomy_mat)
samples_mifish_2023 = sample_data(sample_metadata)

phylo_mifish_2023 <- phyloseq(ASV_mifish_2023, TAX_mifish_2023, samples_mifish_2023)

```

### 2024 MiFish data
```{r}
asv_count <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2024-MiFIsh/ASV2Taxonomy/ASVs_counts_NOUNKNOWNS.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count) <- asv_count$x
asv_count <- asv_count %>% select(-x)
asv_count_mat <- as.matrix(asv_count)

asv_taxonomy <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2024-MiFIsh/ASV2Taxonomy/results-revamp-2024-MiFIsh_asvTaxonomyTable_NOUNKNOWNS.txt", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_taxonomy) <- asv_taxonomy$ASV
asv_taxonomy <- asv_taxonomy %>% select(-ASV)
asv_taxonomy_mat <- as.matrix(asv_taxonomy)

sample_metadata <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2024-MiFIsh/sample_metadata_forR.txt", header=TRUE, stringsAsFactors=TRUE)
row.names(sample_metadata) <- sample_metadata$Sample
sample_metadata <- sample_metadata %>% select(-Sample)

ASV_mifish_2024 = otu_table(asv_count_mat, taxa_are_rows = TRUE)
TAX_mifish_2024 = tax_table(asv_taxonomy_mat)
samples_mifish_2024 = sample_data(sample_metadata)

phylo_mifish_2024 <- phyloseq(ASV_mifish_2024, TAX_mifish_2024, samples_mifish_2024)

```

## Elas02
### 2021 data
```{r}
asv_count <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2021-Elas02/ASV2Taxonomy/ASVs_counts_NOUNKNOWNS.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count) <- asv_count$x
asv_count <- asv_count %>% select(-x)
asv_count_mat <- as.matrix(asv_count)

asv_taxonomy <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2021-Elas02/ASV2Taxonomy/results-revamp-2021-Elas02_asvTaxonomyTable_NOUNKNOWNS.txt", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_taxonomy) <- asv_taxonomy$ASV
asv_taxonomy <- asv_taxonomy %>% select(-ASV)
asv_taxonomy_mat <- as.matrix(asv_taxonomy)

sample_metadata <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2021-Elas02/sample_metadata_forR.txt", header=TRUE, stringsAsFactors=TRUE)
row.names(sample_metadata) <- sample_metadata$Sample
sample_metadata <- sample_metadata %>% select(-Sample)

ASV_elas_2021 = otu_table(asv_count_mat, taxa_are_rows = TRUE)
TAX_elas_2021 = tax_table(asv_taxonomy_mat)
samples_elas_2021 = sample_data(sample_metadata)

phylo_elas_2021 <- phyloseq(ASV_elas_2021, TAX_elas_2021, samples_elas_2021)

```

### 2023 data
```{r}
asv_count <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2023-Elas02/ASV2Taxonomy/ASVs_counts_NOUNKNOWNS.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count) <- asv_count$x
asv_count <- asv_count %>% select(-x)
asv_count_mat <- as.matrix(asv_count)

asv_taxonomy <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2023-Elas02/ASV2Taxonomy/results-revamp-2023-Elas02_asvTaxonomyTable_NOUNKNOWNS.txt", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_taxonomy) <- asv_taxonomy$ASV
asv_taxonomy <- asv_taxonomy %>% select(-ASV)
asv_taxonomy_mat <- as.matrix(asv_taxonomy)

sample_metadata <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2023-Elas02/sample_metadata_forR.txt", header=TRUE, stringsAsFactors=TRUE)
row.names(sample_metadata) <- sample_metadata$Sample
sample_metadata <- sample_metadata %>% select(-Sample)

ASV_elas_2023 = otu_table(asv_count_mat, taxa_are_rows = TRUE)
TAX_elas_2023 = tax_table(asv_taxonomy_mat)
samples_elas_2023 = sample_data(sample_metadata)

phylo_elas_2023 <- phyloseq(ASV_elas_2023, TAX_elas_2023, samples_elas_2023)

```


### 2024 data
```{r}
asv_count <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2024-Elas02/ASV2Taxonomy/ASVs_counts_NOUNKNOWNS.tsv", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_count) <- asv_count$x
asv_count <- asv_count %>% select(-x)
asv_count_mat <- as.matrix(asv_count)

asv_taxonomy <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2024-Elas02/ASV2Taxonomy/results-revamp-2024-Elas02_asvTaxonomyTable_NOUNKNOWNS.txt", header=TRUE, stringsAsFactors=FALSE)
row.names(asv_taxonomy) <- asv_taxonomy$ASV
asv_taxonomy <- asv_taxonomy %>% select(-ASV)
asv_taxonomy_mat <- as.matrix(asv_taxonomy)

sample_metadata <- read.delim("/Volumes/easystore/eDNA/shirp-edna/results-revamp-2024-Elas02/sample_metadata_forR.txt", header=TRUE, stringsAsFactors=TRUE)
row.names(sample_metadata) <- sample_metadata$Sample
sample_metadata <- sample_metadata %>% select(-Sample)

ASV_elas_2024 = otu_table(asv_count_mat, taxa_are_rows = TRUE)
TAX_elas_2024 = tax_table(asv_taxonomy_mat)
samples_elas_2024 = sample_data(sample_metadata)

phylo_elas_2024 <- phyloseq(ASV_elas_2024, TAX_elas_2024, samples_elas_2024)

```



# Filtering & QC
## How many taxa before filtering
First change names of phyloseq objects. Retain `phylo_XX_XX`, which is the unfiltered object. Make new objects, `ps_XX_XX`, which will be what remains after the following filtering steps:
```{r}
ps_mifish_2020 <- phylo_mifish_2020
ps_mifish_2021 <- phylo_mifish_2021
ps_mifish_2022 <- phylo_mifish_2022
ps_mifish_2023 <- phylo_mifish_2023
ps_mifish_2024 <- phylo_mifish_2024

ps_elas_2021 <- phylo_elas_2021
ps_elas_2023 <- phylo_elas_2023
ps_elas_2024 <- phylo_elas_2024

# How many taxa before filtering?
nrow(tax_table(phylo_mifish_2020)) 
nrow(tax_table(phylo_mifish_2021)) 
nrow(tax_table(phylo_mifish_2022)) # this number is misleading because I had to do the dada2 step in two subsets, then append the tables, so some of these ASVs very likely should overlap
nrow(tax_table(phylo_mifish_2023)) 
nrow(tax_table(phylo_mifish_2024)) 

nrow(tax_table(phylo_elas_2021)) 
nrow(tax_table(phylo_elas_2023)) 
nrow(tax_table(phylo_elas_2024)) 

```



## Replace NA in taxa list with next highest annotated taxon
```{r}

tax_table(ps_mifish_2020) <- tax_table(fantaxtic::name_na_taxa(ps_mifish_2020, include_rank = T))
tax_table(ps_mifish_2021) <- tax_table(fantaxtic::name_na_taxa(ps_mifish_2021, include_rank = T))
tax_table(ps_mifish_2022) <- tax_table(fantaxtic::name_na_taxa(ps_mifish_2022, include_rank = T))
tax_table(ps_mifish_2023) <- tax_table(fantaxtic::name_na_taxa(ps_mifish_2023, include_rank = T))
tax_table(ps_mifish_2024) <- tax_table(fantaxtic::name_na_taxa(ps_mifish_2024, include_rank = T))


tax_table(ps_elas_2021) <- tax_table(fantaxtic::name_na_taxa(ps_elas_2021, include_rank = T))
tax_table(ps_elas_2023) <- tax_table(fantaxtic::name_na_taxa(ps_elas_2023, include_rank = T))
tax_table(ps_elas_2024) <- tax_table(fantaxtic::name_na_taxa(ps_elas_2024, include_rank = T))



```


## Remove prokaryotes
check
```{r}
unique(tax_table(ps_mifish_2020)[, "Kingdom"])
unique(tax_table(ps_mifish_2021)[, "Kingdom"])
unique(tax_table(ps_mifish_2022)[, "Kingdom"])
unique(tax_table(ps_mifish_2023)[, "Kingdom"])
unique(tax_table(ps_mifish_2024)[, "Kingdom"])

unique(tax_table(ps_elas_2021)[, "Kingdom"])
unique(tax_table(ps_elas_2023)[, "Kingdom"])
unique(tax_table(ps_elas_2024)[, "Kingdom"])
```

remove
```{r}
#ps_mifish_2020 <- subset_taxa(ps_mifish_2020, !Kingdom == 'Bacteria')
ps_mifish_2021 <- subset_taxa(ps_mifish_2021, !Kingdom == 'Bacteria')
ps_mifish_2022 <- subset_taxa(ps_mifish_2022, !(Kingdom %in% c("Bacteria", "Archaea"))) # 2022 also had archaea
ps_mifish_2023 <- subset_taxa(ps_mifish_2023, !Kingdom == 'Bacteria')
ps_mifish_2024 <- subset_taxa(ps_mifish_2024, !Kingdom == 'Bacteria')


# no prokaryotes are in elas objects

```


check
```{r}
unique(tax_table(ps_mifish_2020)[, "Kingdom"])
unique(tax_table(ps_mifish_2021)[, "Kingdom"])
unique(tax_table(ps_mifish_2022)[, "Kingdom"])
unique(tax_table(ps_mifish_2023)[, "Kingdom"])
unique(tax_table(ps_mifish_2024)[, "Kingdom"])

unique(tax_table(ps_elas_2021)[, "Kingdom"])
unique(tax_table(ps_elas_2023)[, "Kingdom"])
unique(tax_table(ps_elas_2024)[, "Kingdom"])
```


## Remove non-marine mammals

check unique classes
```{r}
unique(tax_table(ps_mifish_2020)[, "Class"])
unique(tax_table(ps_mifish_2021)[, "Class"])
unique(tax_table(ps_mifish_2022)[, "Class"])
unique(tax_table(ps_mifish_2023)[, "Class"])
unique(tax_table(ps_mifish_2024)[, "Class"])

unique(tax_table(ps_elas_2021)[, "Class"])
unique(tax_table(ps_elas_2023)[, "Class"])
unique(tax_table(ps_elas_2024)[, "Class"])
```

all ps objects have Mammalia


Check whether these are contamination (eg. human, dog, etc) or something we want to keep (whales, dolphins, etc)
```{r}
unique(tax_table(ps_mifish_2020)[tax_table(ps_mifish_2020)[, "Class"] == "Mammalia", "Species"])
unique(tax_table(ps_mifish_2021)[tax_table(ps_mifish_2021)[, "Class"] == "Mammalia", "Species"])
unique(tax_table(ps_mifish_2022)[tax_table(ps_mifish_2022)[, "Class"] == "Mammalia", "Species"])
unique(tax_table(ps_mifish_2023)[tax_table(ps_mifish_2023)[, "Class"] == "Mammalia", "Species"])
unique(tax_table(ps_mifish_2024)[tax_table(ps_mifish_2024)[, "Class"] == "Mammalia", "Species"])


unique(tax_table(ps_elas_2021)[tax_table(ps_elas_2021)[, "Class"] == "Mammalia", "Species"])
unique(tax_table(ps_elas_2023)[tax_table(ps_elas_2023)[, "Class"] == "Mammalia", "Species"])
unique(tax_table(ps_elas_2024)[tax_table(ps_elas_2024)[, "Class"] == "Mammalia", "Species"])


```

Summary:
MiFish:
2020 - remove all mammals
2021 - remove all mammals (although Puma yagouaroundi is very interesting...)
2022 - retain Tursiops truncatus and Unknown Delphinidae (Family), remove rest of mammals
2023 - remove all mammals
2024 - remove all mammals

Elas02:
remove all mammals from all 3 datasets

```{r}
ps_mifish_2020 <- subset_taxa(ps_mifish_2020, !Class == 'Mammalia')
ps_mifish_2021 <- subset_taxa(ps_mifish_2021, !Class == 'Mammalia')
ps_mifish_2022 <- subset_taxa(ps_mifish_2022, !Species == 'Homo sapiens' & !Species == 'Felis catus' & !Species == 'Bos taurus' & !Species == 'Canis lupus' & !Species == 'Sus scrofa' & !Species == 'Unknown Bovidae (Family)' & !Species == 'Unknown Bos (Genus)' & !Species == 'Unknown Hominidae (Family)' & !Species == 'Unknown Felidae (Family)' & !Species == 'Unknown Primates (Order)')
ps_mifish_2023 <- subset_taxa(ps_mifish_2023, !Class == 'Mammalia')
ps_mifish_2024 <- subset_taxa(ps_mifish_2024, !Class == 'Mammalia')

ps_elas_2021 <- subset_taxa(ps_elas_2021, !Class == 'Mammalia')
ps_elas_2023 <- subset_taxa(ps_elas_2023, !Class == 'Mammalia')
ps_elas_2024 <- subset_taxa(ps_elas_2024, !Class == 'Mammalia')

```

## Remove the classes Aves and Zygnemophyceae

Zygnemophyceae are green algae, cam remove all

Check which Aves were detected
```{r}

#unique(tax_table(ps_mifish_2020)[tax_table(ps_mifish_2020)[, "Class"] == "Aves", "Species"]) # none detected in 2020
unique(tax_table(ps_mifish_2021)[tax_table(ps_mifish_2021)[, "Class"] == "Aves", "Species"])
unique(tax_table(ps_mifish_2022)[tax_table(ps_mifish_2022)[, "Class"] == "Aves", "Species"])
unique(tax_table(ps_mifish_2023)[tax_table(ps_mifish_2023)[, "Class"] == "Aves", "Species"])
unique(tax_table(ps_mifish_2024)[tax_table(ps_mifish_2024)[, "Class"] == "Aves", "Species"])


unique(tax_table(ps_elas_2021)[tax_table(ps_elas_2021)[, "Class"] == "Aves", "Species"])
unique(tax_table(ps_elas_2023)[tax_table(ps_elas_2023)[, "Class"] == "Aves", "Species"])
unique(tax_table(ps_elas_2024)[tax_table(ps_elas_2024)[, "Class"] == "Aves", "Species"])

```

Turkey, cormorant, chicken, gulls, common parakeet, common myna, song sparrow...
Some of these are interesting but also some are possible contamination. Remove all since they are not the target of the study

Also remove Actinopteri from Elas02 libraries

```{r}
ps_mifish_2020 <- subset_taxa(ps_mifish_2020, !Class == 'Aves')
ps_mifish_2021 <- subset_taxa(ps_mifish_2021, !Class == 'Aves')
ps_mifish_2022 <- subset_taxa(ps_mifish_2022, !Class == 'Aves' & !Class == 'Zygnemophyceae')
ps_mifish_2023 <- subset_taxa(ps_mifish_2023, !Class == 'Aves')
ps_mifish_2024 <- subset_taxa(ps_mifish_2024, !Class == 'Aves')

ps_elas_2021 <- subset_taxa(ps_elas_2021, !Class == 'Aves' & !Class == 'Actinopteri')
ps_elas_2023 <- subset_taxa(ps_elas_2023, !Class == 'Aves' & !Class == 'Actinopteri')
ps_elas_2024 <- subset_taxa(ps_elas_2024, !Class == 'Aves' & !Class == 'Actinopteri')
```

check
```{r}
unique(tax_table(ps_mifish_2020)[, "Class"])
unique(tax_table(ps_mifish_2021)[, "Class"])
unique(tax_table(ps_mifish_2022)[, "Class"])
unique(tax_table(ps_mifish_2023)[, "Class"])
unique(tax_table(ps_mifish_2024)[, "Class"])

unique(tax_table(ps_elas_2021)[, "Class"])
unique(tax_table(ps_elas_2023)[, "Class"])
unique(tax_table(ps_elas_2024)[, "Class"])
```



## Remove incorrect annotations
check unique species
```{r}
data.frame(unique(tax_table(ps_mifish_2020)[, "Species"]))
data.frame(unique(tax_table(ps_mifish_2021)[, "Species"]))
data.frame(unique(tax_table(ps_mifish_2022)[, "Species"]))
data.frame(unique(tax_table(ps_mifish_2023)[, "Species"]))
data.frame(unique(tax_table(ps_mifish_2024)[, "Species"]))

data.frame(unique(tax_table(ps_elas_2021)[, "Species"]))
data.frame(unique(tax_table(ps_elas_2023)[, "Species"]))
data.frame(unique(tax_table(ps_elas_2024)[, "Species"]))
```

### Change all menhaden to genus-level annotation

There is something going on with menhaden. The regional species should be Atlantic menhaden (B. tyrannus) but we frequently get Finescale and Gulf menhaden. Retain all at genus level for now and look into phylogenetics later (if it can even be resolved with this short amplicon)

```{r}
tax_table(ps_mifish_2020)[tax_table(ps_mifish_2020)[, "Genus"] == "Brevoortia", "Species"] <- "Unknown Brevoortia (Genus)"
tax_table(ps_mifish_2021)[tax_table(ps_mifish_2021)[, "Genus"] == "Brevoortia", "Species"] <- "Unknown Brevoortia (Genus)"
tax_table(ps_mifish_2022)[tax_table(ps_mifish_2022)[, "Genus"] == "Brevoortia", "Species"] <- "Unknown Brevoortia (Genus)"
tax_table(ps_mifish_2023)[tax_table(ps_mifish_2023)[, "Genus"] == "Brevoortia", "Species"] <- "Unknown Brevoortia (Genus)"
tax_table(ps_mifish_2024)[tax_table(ps_mifish_2024)[, "Genus"] == "Brevoortia", "Species"] <- "Unknown Brevoortia (Genus)"

```

check
```{r}
unique(tax_table(ps_mifish_2020)[tax_table(ps_mifish_2020)[, "Genus"] == "Brevoortia","Species"])
unique(tax_table(ps_mifish_2021)[tax_table(ps_mifish_2021)[, "Genus"] == "Brevoortia","Species"])
unique(tax_table(ps_mifish_2022)[tax_table(ps_mifish_2022)[, "Genus"] == "Brevoortia","Species"])
unique(tax_table(ps_mifish_2023)[tax_table(ps_mifish_2023)[, "Genus"] == "Brevoortia","Species"])
unique(tax_table(ps_mifish_2024)[tax_table(ps_mifish_2024)[, "Genus"] == "Brevoortia","Species"])

```
### Change Mustelus species

check
```{r}
unique(tax_table(ps_elas_2021)[tax_table(ps_elas_2021)[, "Genus"] == "Mustelus","Species"])
unique(tax_table(ps_elas_2023)[tax_table(ps_elas_2023)[, "Genus"] == "Mustelus","Species"])
unique(tax_table(ps_elas_2024)[tax_table(ps_elas_2024)[, "Genus"] == "Mustelus","Species"])

```

Mustelus mosis, Mustelus henlei, and Mustelus griseus are likely incorrect (these are Asian/Arabian/Pacific species). 

Change to Unknown Mustelus (Genus)
```{r}
tax_table(ps_elas_2021)[tax_table(ps_elas_2021)[, "Species"] %in% c("Mustelus mosis", "Mustelus henlei", "Mustelus griseus")] <- "Unknown Mustelus (Genus)"

tax_table(ps_elas_2023)[tax_table(ps_elas_2023)[, "Species"] %in% c("Mustelus mosis", "Mustelus henlei", "Mustelus griseus")] <- "Unknown Mustelus (Genus)"

tax_table(ps_elas_2024)[tax_table(ps_elas_2024)[, "Species"] %in% c("Mustelus mosis", "Mustelus henlei", "Mustelus griseus")] <- "Unknown Mustelus (Genus)"

```

check
```{r}
unique(tax_table(ps_elas_2021)[tax_table(ps_elas_2021)[, "Genus"] == "Mustelus","Species"])
unique(tax_table(ps_elas_2023)[tax_table(ps_elas_2023)[, "Genus"] == "Mustelus","Species"])
unique(tax_table(ps_elas_2024)[tax_table(ps_elas_2024)[, "Genus"] == "Mustelus","Species"])

```

### Change Pteroplatytrygon sp.
check
```{r}
#unique(tax_table(ps_elas_2021)[tax_table(ps_elas_2021)[, "Genus"] == "Pteroplatytrygon","Species"])
unique(tax_table(ps_elas_2023)[tax_table(ps_elas_2023)[, "Genus"] == "Pteroplatytrygon","Species"])
#unique(tax_table(ps_elas_2024)[tax_table(ps_elas_2024)[, "Genus"] == "Pteroplatytrygon","Species"])

```


There is only one possible species, put in regional species name, violacea



```{r}

tax_table(ps_elas_2023)[tax_table(ps_elas_2023)[, "Genus"] == "Pteroplatytrygon", "Species"] <- "Pteroplatytrygon violacea"

```

check
```{r}
unique(tax_table(ps_elas_2023)[tax_table(ps_elas_2023)[, "Genus"] == "Pteroplatytrygon","Species"])

```



## Add in Common Names
Import
```{r}
commonnames <- read_csv(file = "../eDNA-databases/commonnames.csv")

```
Append common names
```{r}
# Extract, update, tax_table- MiFish 2020
taxonomy_mifish_2020 <- as.data.frame(tax_table(ps_mifish_2020)) %>%
  rownames_to_column("ASV") %>%
  left_join(commonnames, by = "Species") 
row.names(taxonomy_mifish_2020) <- taxonomy_mifish_2020$ASV
taxonomy_mifish_2020 <- taxonomy_mifish_2020 %>% select(-ASV)
# Update the phyloseq object
tax_table(ps_mifish_2020) <- as.matrix(taxonomy_mifish_2020)

# 2021
taxonomy_mifish_2021 <- as.data.frame(tax_table(ps_mifish_2021)) %>%
  rownames_to_column("ASV") %>%
  left_join(commonnames, by = "Species") 
row.names(taxonomy_mifish_2021) <- taxonomy_mifish_2021$ASV
taxonomy_mifish_2021 <- taxonomy_mifish_2021 %>% select(-ASV)
tax_table(ps_mifish_2021) <- as.matrix(taxonomy_mifish_2021)

# 2022
taxonomy_mifish_2022 <- as.data.frame(tax_table(ps_mifish_2022)) %>%
  rownames_to_column("ASV") %>%
  left_join(commonnames, by = "Species") 
row.names(taxonomy_mifish_2022) <- taxonomy_mifish_2022$ASV
taxonomy_mifish_2022 <- taxonomy_mifish_2022 %>% select(-ASV)
tax_table(ps_mifish_2022) <- as.matrix(taxonomy_mifish_2022)

# 2023
taxonomy_mifish_2023 <- as.data.frame(tax_table(ps_mifish_2023)) %>%
  rownames_to_column("ASV") %>%
  left_join(commonnames, by = "Species") 
row.names(taxonomy_mifish_2023) <- taxonomy_mifish_2023$ASV
taxonomy_mifish_2023 <- taxonomy_mifish_2023 %>% select(-ASV)
tax_table(ps_mifish_2023) <- as.matrix(taxonomy_mifish_2023)

# 2024
taxonomy_mifish_2024 <- as.data.frame(tax_table(ps_mifish_2024)) %>%
  rownames_to_column("ASV") %>%
  left_join(commonnames, by = "Species") 
row.names(taxonomy_mifish_2024) <- taxonomy_mifish_2024$ASV
taxonomy_mifish_2024 <- taxonomy_mifish_2024 %>% select(-ASV)
tax_table(ps_mifish_2024) <- as.matrix(taxonomy_mifish_2024)

# Elas
#2021
taxonomy_elas_2021 <- as.data.frame(tax_table(ps_elas_2021)) %>%
  rownames_to_column("ASV") %>%
  left_join(commonnames, by = "Species") 
row.names(taxonomy_elas_2021) <- taxonomy_elas_2021$ASV
taxonomy_elas_2021 <- taxonomy_elas_2021 %>% select(-ASV)
tax_table(ps_elas_2021) <- as.matrix(taxonomy_elas_2021)

#2023
taxonomy_elas_2023 <- as.data.frame(tax_table(ps_elas_2023)) %>%
  rownames_to_column("ASV") %>%
  left_join(commonnames, by = "Species") 
row.names(taxonomy_elas_2023) <- taxonomy_elas_2023$ASV
taxonomy_elas_2023 <- taxonomy_elas_2023 %>% select(-ASV)
tax_table(ps_elas_2023) <- as.matrix(taxonomy_elas_2023)

#2024
taxonomy_elas_2024 <- as.data.frame(tax_table(ps_elas_2024)) %>%
  rownames_to_column("ASV") %>%
  left_join(commonnames, by = "Species") 
row.names(taxonomy_elas_2024) <- taxonomy_elas_2024$ASV
taxonomy_elas_2024 <- taxonomy_elas_2024 %>% select(-ASV)
tax_table(ps_elas_2024) <- as.matrix(taxonomy_elas_2024)

```

Check for missing (NA) common names
(If there are, update commonnames database)
```{r}
#tax_table(ps_mifish_2020)[is.na(tax_table(ps_mifish_2020)[, "CommonName"])]
#tax_table(ps_mifish_2021)[is.na(tax_table(ps_mifish_2021)[, "CommonName"])]
#tax_table(ps_mifish_2022)[is.na(tax_table(ps_mifish_2022)[, "CommonName"])]
#tax_table(ps_mifish_2023)[is.na(tax_table(ps_mifish_2023)[, "CommonName"])]
#tax_table(ps_mifish_2024)[is.na(tax_table(ps_mifish_2024)[, "CommonName"])]


#tax_table(ps_elas_2021)[is.na(tax_table(ps_elas_2021)[, "CommonName"])]
#tax_table(ps_elas_2023)[is.na(tax_table(ps_elas_2023)[, "CommonName"])]
#tax_table(ps_elas_2024)[is.na(tax_table(ps_elas_2024)[, "CommonName"])]

```

## Remove Low Abundant ASVs

Based on some analysis in the `Expedition-eDNA-plots.Rmd` notebook, remove ASVs that have <500 reads in a sample
```{r}
data.frame(otu_table(ps_mifish_2020)) # compare before
ps_mifish_2020 <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps_mifish_2020,
  minabund = 500,
  relabund = FALSE)
data.frame(otu_table(ps_mifish_2020)) # and after

data.frame(otu_table(ps_mifish_2021)) # compare before
ps_mifish_2021 <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps_mifish_2021,
  minabund = 500,
  relabund = FALSE)
data.frame(otu_table(ps_mifish_2021)) # and after

data.frame(otu_table(ps_mifish_2022)) # compare before
ps_mifish_2022 <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps_mifish_2022,
  minabund = 500,
  relabund = FALSE)
data.frame(otu_table(ps_mifish_2022)) # and after

data.frame(otu_table(ps_mifish_2023)) # compare before
ps_mifish_2023 <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps_mifish_2023,
  minabund = 500,
  relabund = FALSE)
data.frame(otu_table(ps_mifish_2023)) # and after

data.frame(otu_table(ps_mifish_2024)) # compare before
ps_mifish_2024 <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps_mifish_2024,
  minabund = 500,
  relabund = FALSE)
data.frame(otu_table(ps_mifish_2024)) # and after




data.frame(otu_table(ps_elas_2021)) # compare before
ps_elas_2021 <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps_elas_2021,
  minabund = 500,
  relabund = FALSE)
data.frame(otu_table(ps_elas_2021)) # and after

data.frame(otu_table(ps_elas_2023)) # compare before
ps_elas_2023 <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps_elas_2023,
  minabund = 500,
  relabund = FALSE)
data.frame(otu_table(ps_elas_2023)) # and after

data.frame(otu_table(ps_elas_2024)) # compare before
ps_elas_2024 <- metagMisc::phyloseq_filter_sample_wise_abund_trim(
  ps_elas_2024,
  minabund = 500,
  relabund = FALSE)
data.frame(otu_table(ps_elas_2024)) # and after
```


## Remove Low Effort Samples

Plot bar plots of total sequence reads across all samples
```{r}
plot_bar(ps_mifish_2020)
plot_bar(ps_mifish_2021)
plot_bar(ps_mifish_2022)
plot_bar(ps_mifish_2023)
plot_bar(ps_mifish_2024)


plot_bar(ps_elas_2021)
plot_bar(ps_elas_2023)
plot_bar(ps_elas_2024)

```

Cut out any sample with <15,000 reads from all libraries


```{r}
ps_mifish_2020 <- prune_samples(sample_sums(ps_mifish_2020)>=25000,ps_mifish_2020)
ps_mifish_2021 <- prune_samples(sample_sums(ps_mifish_2021)>=25000,ps_mifish_2021)
ps_mifish_2022 <- prune_samples(sample_sums(ps_mifish_2022)>=25000,ps_mifish_2022)
ps_mifish_2023 <- prune_samples(sample_sums(ps_mifish_2023)>=25000,ps_mifish_2023)
ps_mifish_2024 <- prune_samples(sample_sums(ps_mifish_2024)>=25000,ps_mifish_2024)


ps_elas_2021 <- prune_samples(sample_sums(ps_elas_2021)>=25000,ps_elas_2021)
ps_elas_2023 <- prune_samples(sample_sums(ps_elas_2023)>=25000,ps_elas_2023)
ps_elas_2024 <- prune_samples(sample_sums(ps_elas_2024)>=25000,ps_elas_2024)

```

Plot again to check
```{r}
plot_bar(ps_mifish_2020)
plot_bar(ps_mifish_2021)
plot_bar(ps_mifish_2022)
plot_bar(ps_mifish_2023)
plot_bar(ps_mifish_2024)


plot_bar(ps_elas_2021)
plot_bar(ps_elas_2023)
plot_bar(ps_elas_2024)

```


## Agglomerate 
```{r}
ps_mifish_2020_glommed <- tax_glom(ps_mifish_2020, "Species-CommonName")
ps_mifish_2021_glommed <- tax_glom(ps_mifish_2021, "Species-CommonName")
ps_mifish_2022_glommed <- tax_glom(ps_mifish_2022, "Species-CommonName")
ps_mifish_2023_glommed <- tax_glom(ps_mifish_2023, "Species-CommonName")
ps_mifish_2024_glommed <- tax_glom(ps_mifish_2024, "Species-CommonName")

ps_elas_2021_glommed <- tax_glom(ps_elas_2021, "Species-CommonName")
ps_elas_2023_glommed <- tax_glom(ps_elas_2023, "Species-CommonName")
ps_elas_2024_glommed <- tax_glom(ps_elas_2024, "Species-CommonName")


```

check glommed data tables
```{r}
data.frame(tax_table(ps_mifish_2020_glommed))
data.frame(tax_table(ps_mifish_2021_glommed))
data.frame(tax_table(ps_mifish_2022_glommed))
data.frame(tax_table(ps_mifish_2023_glommed))
data.frame(tax_table(ps_mifish_2024_glommed))

data.frame(tax_table(ps_elas_2021_glommed))
data.frame(tax_table(ps_elas_2023_glommed))
data.frame(tax_table(ps_elas_2024_glommed))


```


## krona plots
MUST PASTE inTO the console:
NOTE- this function doesn't allow you to name Krona files. I went in manually in `krona_files` and names them myself to share with collaborators
```{r}
source('https://raw.githubusercontent.com/markschl/embed_krona/master/embed_krona.R')

mifish_2020_krona <- plot_krona(ps_mifish_2020_glommed) # gets written as krona_files/krona_3.html
mifish_2021_krona <- plot_krona(ps_mifish_2021_glommed) 
mifish_2022_krona <- plot_krona(ps_mifish_2022_glommed) 
mifish_2023_krona <- plot_krona(ps_mifish_2023_glommed) 
mifish_2024_krona <- plot_krona(ps_mifish_2024_glommed) 

elas_2021_krona <- plot_krona(ps_elas_2021_glommed) 
elas_2023_krona <- plot_krona(ps_elas_2023_glommed) 
elas_2024_krona <- plot_krona(ps_elas_2024_glommed) 

```

## Export tables
```{r}
mifish_2020_glommed_asv_table <- data.frame(otu_table(ps_mifish_2020_glommed))
mifish_2020_glommed_tax_table <- rownames_to_column(data.frame(tax_table(ps_mifish_2020_glommed)), var = "1stRepresentative_ASV")
mifish_2020_glommed_sample_data <- data.frame(sample_data(ps_mifish_2020_glommed))
write.csv(mifish_2020_glommed_asv_table,"filtered_data/mifish_2020_glommed_asv_table.csv", row.names = TRUE)
write.csv(mifish_2020_glommed_tax_table,"filtered_data/mifish_2020_glommed_tax_table.csv", row.names = FALSE)
write.csv(mifish_2020_glommed_sample_data,"filtered_data/mifish_2020_glommed_sample_data.csv", row.names = TRUE)

mifish_2021_glommed_asv_table <- data.frame(otu_table(ps_mifish_2021_glommed))
mifish_2021_glommed_tax_table <- rownames_to_column(data.frame(tax_table(ps_mifish_2021_glommed)), var = "1stRepresentative_ASV")
mifish_2021_glommed_sample_data <- data.frame(sample_data(ps_mifish_2021_glommed))
write.csv(mifish_2021_glommed_asv_table,"filtered_data/mifish_2021_glommed_asv_table.csv", row.names = TRUE)
write.csv(mifish_2021_glommed_tax_table,"filtered_data/mifish_2021_glommed_tax_table.csv", row.names = FALSE)
write.csv(mifish_2021_glommed_sample_data,"filtered_data/mifish_2021_glommed_sample_data.csv", row.names = TRUE)



mifish_2022_glommed_asv_table <- data.frame(otu_table(ps_mifish_2022_glommed))
mifish_2022_glommed_tax_table <- rownames_to_column(data.frame(tax_table(ps_mifish_2022_glommed)), var = "1stRepresentative_ASV")
mifish_2022_glommed_sample_data <- data.frame(sample_data(ps_mifish_2022_glommed))
write.csv(mifish_2022_glommed_asv_table,"filtered_data/mifish_2022_glommed_asv_table.csv", row.names = TRUE)
write.csv(mifish_2022_glommed_tax_table,"filtered_data/mifish_2022_glommed_tax_table.csv", row.names = FALSE)
write.csv(mifish_2022_glommed_sample_data,"filtered_data/mifish_2022_glommed_sample_data.csv", row.names = TRUE)


mifish_2023_glommed_asv_table <- data.frame(otu_table(ps_mifish_2023_glommed))
mifish_2023_glommed_tax_table <- rownames_to_column(data.frame(tax_table(ps_mifish_2023_glommed)), var = "1stRepresentative_ASV")
mifish_2023_glommed_sample_data <- data.frame(sample_data(ps_mifish_2023_glommed))
write.csv(mifish_2023_glommed_asv_table,"filtered_data/mifish_2023_glommed_asv_table.csv", row.names = TRUE)
write.csv(mifish_2023_glommed_tax_table,"filtered_data/mifish_2023_glommed_tax_table.csv", row.names = FALSE)
write.csv(mifish_2023_glommed_sample_data,"filtered_data/mifish_2023_glommed_sample_data.csv", row.names = TRUE)




mifish_2024_glommed_asv_table <- data.frame(otu_table(ps_mifish_2024_glommed))
mifish_2024_glommed_tax_table <- rownames_to_column(data.frame(tax_table(ps_mifish_2024_glommed)), var = "1stRepresentative_ASV")
mifish_2024_glommed_sample_data <- data.frame(sample_data(ps_mifish_2024_glommed))
write.csv(mifish_2024_glommed_asv_table,"filtered_data/mifish_2024_glommed_asv_table.csv", row.names = TRUE)
write.csv(mifish_2024_glommed_tax_table,"filtered_data/mifish_2024_glommed_tax_table.csv", row.names = FALSE)
write.csv(mifish_2024_glommed_sample_data,"filtered_data/mifish_2024_glommed_sample_data.csv", row.names = TRUE)



elas_2021_glommed_asv_table <- data.frame(otu_table(ps_elas_2021_glommed))
elas_2021_glommed_tax_table <- rownames_to_column(data.frame(tax_table(ps_elas_2021_glommed)), var = "1stRepresentative_ASV")
elas_2021_glommed_sample_data <- data.frame(sample_data(ps_elas_2021_glommed))
write.csv(elas_2021_glommed_asv_table,"filtered_data/elas_2021_glommed_asv_table.csv", row.names = TRUE)
write.csv(elas_2021_glommed_tax_table,"filtered_data/elas_2021_glommed_tax_table.csv", row.names = FALSE)
write.csv(elas_2021_glommed_sample_data,"filtered_data/elas_2021_glommed_sample_data.csv", row.names = TRUE)

elas_2023_glommed_asv_table <- data.frame(otu_table(ps_elas_2023_glommed))
elas_2023_glommed_tax_table <- rownames_to_column(data.frame(tax_table(ps_elas_2023_glommed)), var = "1stRepresentative_ASV")
elas_2023_glommed_sample_data <- data.frame(sample_data(ps_elas_2023_glommed))
write.csv(elas_2023_glommed_asv_table,"filtered_data/elas_2023_glommed_asv_table.csv", row.names = TRUE)
write.csv(elas_2023_glommed_tax_table,"filtered_data/elas_2023_glommed_tax_table.csv", row.names = FALSE)
write.csv(elas_2023_glommed_sample_data,"filtered_data/elas_2023_glommed_sample_data.csv", row.names = TRUE)

elas_2024_glommed_asv_table <- data.frame(otu_table(ps_elas_2024_glommed))
elas_2024_glommed_tax_table <- rownames_to_column(data.frame(tax_table(ps_elas_2024_glommed)), var = "1stRepresentative_ASV")
elas_2024_glommed_sample_data <- data.frame(sample_data(ps_elas_2024_glommed))
write.csv(elas_2024_glommed_asv_table,"filtered_data/elas_2024_glommed_asv_table.csv", row.names = TRUE)
write.csv(elas_2024_glommed_tax_table,"filtered_data/elas_2024_glommed_tax_table.csv", row.names = FALSE)
write.csv(elas_2024_glommed_sample_data,"filtered_data/elas_2024_glommed_sample_data.csv", row.names = TRUE)
```

